### 2013-05-16 paper/main.tex
paper started

\documentclass[a4paper,draft]{amsart}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bbm}
\usepackage{a4wide}

\parindent=0pt
\parskip\smallskipamount

\let\set\mathbbm
\def\<#1>{\langle#1\rangle}
\let\ideal\unlhd
\def\i{\mathrm{i}}
\def\e{\mathrm{e}}

\newtheorem{thm}{Theorem}
\newtheorem{prop}{Proposition}
\newtheorem{defn}{Definition}
\newtheorem{conj}{Conjecture}

\newcommand\todo[1][.]{\edef\tmpa{.}\edef\tmpb{#1}%
  \ifx\tmpa\tmpb
    \typeout{To Be on page \thepage}\fbox{\bf To Be}
  \else
    \typeout{To Be on page \thepage: #1}\fbox{{\bf To Be:} #1}
  \fi
}

\begin{document}

 \author[Manuel Kauers, Maximilian Jaroschek, Fredrik Johansson]
   {Manuel Kauers\,$^\ast$, Maximilian Jaroschek\,$^\ast$, Fredrik Johansson\,$^\ast$}
 \address{Manuel Kauers, Research Institute for Symbolic Computation (RISC), J. Kepler University Linz, Austria}
 \email{mkauers@risc.uni-linz.ac.at}
 \address{Maximilian Jaroschek, Research Institute for Symbolic Computation (RISC), J. Kepler University Linz, Austria}
 \email{mjarosch@risc.uni-linz.ac.at}
 \address{Fredrik Johansson, Research Institute for Symbolic Computation (RISC), J. Kepler University Linz, Austria}
 \email{fjohanss@risc.uni-linz.ac.at}
 \thanks{$^\ast$ Supported by the Austrian FWF grant Y464-N18.}

 \title{Ore Polynomials in Sage}

 \begin{abstract}
We present a Sage implementation of Ore algebras. The main features for the most
common instances include basic arithmetic and actions; gcrd and lclm; D-finite
closure properties; natural transformations between related algebras; guessing;
desingularization; solvers for polynomials, rational functions and (generalized)
power series. This paper is a tutorial on how to use the package.
 \end{abstract}

 \maketitle

%%% page limit: 20, goal: 15--20.

\section{Introduction}

motivation

similar packages: gfun, mallinger, q-gfun, future: mgfun, holonomicfunctions

overview over the package

short demo

outlook, pending tasks (multivariate!)

how to get it and how to install it.

\section{Ore Algebras and Ore Polynomials}

\subsection{In Theory}

basic facts and definitions

reference to literature

\subsection{In Sage}

ground rings

general sigma and delta

predefined shortcuts: D, T, S, F, J, Q, C

\section{General Methods}

arithmetic, lclm, gcrd, action, coercion and conversion, pretty printing, coefficient extraction,
closure properties

\section{Special Methods Standard Algebras}

special methods available for univ algebras with univ algebra as ground ring

\subsection{Differential Operators}

D and theta

\subsection{Recurrence Operators}

delta and S

\subsection{The $q$-Case}

J and Q

\section{Guessing}

\section{Further Examples}

Some longer session showing how to solve a meaningful problem by using several features of the code. Ideas?

\subsection{}

\subsection{}

 \bibliographystyle{plain}
 \bibliography{all}

\end{document}

### 2013-05-16 ore_operator_1_1.py
general factorization stub added

@@ -684,6 +684,48 @@ class UnivariateOreOperatorOverUnivariateRing(UnivariateOreOperator):
         from sage.matrix.constructor import matrix      
         return s.factorial(other.leading_coefficient(), n) * matrix(Alg.base_ring().fraction_field(), mat).det()
 
+    def factor(self):
+        """
+        Returns a factorization of this operator into linear factors, if possible.
+
+        More precisely, the output will be a list  `[L1,L2,...]` of operators such that 
+        
+          * `L1*L2*...` is equal to ``self``
+          * `L2,L3,...` are monic first order operators
+          * `L1` has no first order right hand factor
+
+        This method requires the method ``right_factors()`` to be implemented. 
+
+        EXAMPLE::
+
+          sage: R.<n> = ZZ['n']; A.<Sn> = OreAlgebra(R, 'Sn')
+          sage: L = (-2*n^4 - 17*n^3 - 45*n^2 - 33*n + 9)*Sn^3 + (6*n^4 + 57*n^3 + 168*n^2 + 148*n - 15)*Sn^2 + (-4*n^4 - 44*n^3 - 157*n^2 - 195*n - 38)*Sn + 4*n^3 + 34*n^2 + 80*n + 44
+          sage: L.factor()
+          [-2*n^4 - 17*n^3 - 45*n^2 - 33*n + 9,
+          Sn + (-4*n^5 - 44*n^4 - 171*n^3 - 295*n^2 - 230*n - 66)/(4*n^5 + 44*n^4 + 175*n^3 + 291*n^2 + 147*n - 45),
+          Sn + (2*n + 5)/(-2*n^2 - 7*n - 6),
+          Sn + (-2*n - 4)/(n + 1)]
+          sage: reduce(lambda p,q: p*q, _) - L
+          0
+        
+        """
+        from sage.structure.factorization import Factorization
+
+        if self.is_zero():
+            raise ArithmeticError, "Factorization of 0 not defined."
+        elif self.order() == 0:
+            return [self]
+        elif self.order() == 1:
+            lc = self.leading_coefficient()
+            return [self.parent()(lc), (~lc)*self]
+
+        rf = self.right_factors(early_termination=True)
+        if len(rf) == 0:
+            return [self]
+        else:
+            Q = rf[0].monic(); P = self // Q
+            return P.factor() + [Q]
+
 #############################################################################################################
 
 class UnivariateDifferentialOperatorOverUnivariateRing(UnivariateOreOperatorOverUnivariateRing):

### 2013-05-17 ore_operator.py
annihilator_of_polynomial

@@ -1040,7 +1040,7 @@ class UnivariateOreOperator(OreOperator):
         operator `C` such that for all \"functions\" `f` and `g` with `A.f=B.g=0`
         we have `C.(fg)=0`.
 
-        The function requires that a product rule is associated to the ore algebra
+        The method requires that a product rule is associated to the Ore algebra
         where ``self`` and ``other`` live. (See docstring of OreAlgebra for information
         about product rules.)
 
@@ -1247,6 +1247,109 @@ class UnivariateOreOperator(OreOperator):
 
         L = A.parent()(list(sol[0]))
         return L
+
+    def annihilator_of_polynomial(self, poly, solver=None):
+        """
+        Returns an annihilating operator of a polynomial expression evaluated at solutions of ``self``.
+
+        INPUT:
+
+        - ``poly`` -- a multivariate polynomial, say in the variables `x0,x1,x2,...`, with coefficients
+          in the base ring of the parent of ``self``.
+          The number of variables in the parent of ``poly`` must be at least the order of ``self``.
+        
+        OUTPUT:
+
+          An operator `L` with the following property. 
+          Let `A` be the parent of ``self``.
+          For a function `f` write `Df,D^2f,...` for the functions obtained from `f` by letting the generator
+          of `A` act on them. 
+          Then the output `L` is such that for all `f` annihilated by ``self`` we have
+          `L(p(f,Df,D^2f,...))=0`, where `p` is the input polynomial.
+
+        The method requires that a product rule is associated to `A`. 
+        (See docstring of OreAlgebra for information about product rules.)
+
+        NOTE:
+
+          Even for small input, the output operator may be very large, and its computation may need a lot of time.
+
+        EXAMPLES::
+
+          sage: K.<x> = ZZ['x']; K = K.fraction_field(); R.<n> = K['n']
+          sage: A.<Sn> = OreAlgebra(R, 'Sn'); R.<y0,y1,y2> = K['n'][]
+          sage: L = (n+2)*Sn^2 - (2*n+3)*x*Sn + (n+1)
[…]

### 2013-05-27 guessing.py
alternative guessing options

@@ -566,12 +566,16 @@ def _guess_via_gcrd(data, A, **kwargs):
 
     if kwargs.has_key('degree'):
         max_deg = kwargs['degree']; del kwargs['degree']
+    elif kwargs.has_key('max_degree'):
+        max_deg = kwargs['max_degree']; del kwargs['max_degree']
 
     if kwargs.has_key('min_degree'):
         min_deg = kwargs['min_degree']; del kwargs['min_degree']
 
     if kwargs.has_key('order'):
         max_ord = kwargs['order']; del kwargs['order']
+    elif kwargs.has_key('max_order'):
+        max_ord = kwargs['max_order']; del kwargs['max_order']
 
     if kwargs.has_key('min_order'):
         min_ord = kwargs['min_order']; del kwargs['min_order']

### 2013-05-27 ore_operator.py
some changes in improvedPRS

@@ -1335,7 +1335,7 @@ class UnivariateOreOperator(OreOperator):
 
 def __primitivePRS__(r,additional):
     """
-    Computes one division step in the subresultant polynomial remainder sequence.
+    Computes one division step in the primitive polynomial remainder sequence.
     """
 
     orddiff = r[0].order()-r[1].order()
@@ -1380,11 +1380,11 @@ def __improvedPRS__(r,additional):
     if (len(additional)==0):
         essentialPart = gcd(sigma(r[0].leading_coefficient(),-orddiff),r[1].leading_coefficient())
         phi = Rbase.one()
-        beta = (-Rbase.one())**(orddiff+1)*R.sigma().factorial(sigma(phi,1),orddiff)
+        beta = (-Rbase.one())**(orddiff+1)*sigma().factorial(sigma(phi,1),orddiff)
     else:
         (d2,oldalpha,k,essentialPart,phi) = (additional.pop(),additional.pop(),additional.pop(),additional.pop(),additional.pop())
-        phi = oldalpha / R.sigma().factorial(sigma(phi,1),d2-d1-1)
-        beta = oldalpha.parent()(((-Rbase.one())**(orddiff+1)*R.sigma().factorial(sigma(phi,1),orddiff)*k))
+        phi = oldalpha / sigma().factorial(sigma(phi,1),d2-d0-1)
+        beta = oldalpha.parent()(((-Rbase.one())**(orddiff+1)*sigma().factorial(sigma(phi,1),orddiff)*k))
         essentialPart = sigma(essentialPart,-orddiff)
 
     k = r[1].leading_coefficient()//essentialPart
@@ -1414,11 +1414,12 @@ def __subresultantPRS__(r,additional):
 
     if (len(additional)==0):
         phi = -Rbase.one()
-        beta = (-Rbase.one())*R.sigma().factorial(sigma(phi,1),orddiff)
+        beta = (-Rbase.one())*sigma().factorial(sigma(phi,1),orddiff)
     else:
         (d2,phi) = (additional.pop(),additional.pop())
-        phi = R.sigma().factorial(-r[0].leading_coefficient(),d0-d1) / R.sigma().factorial(sigma(phi,1),d0-d1-1)
-        beta = (-Rbase.one())*R.sigma().factorial(sigma(phi,1),orddiff)*r[0].leading_coefficient()
+        orddiff2 = d2-d0
[…]

### 2013-05-27 doc/dfinite_function.py
some cleanup on the docstrings to support automated html doc generation.

@@ -1,5 +1,431 @@
 
-from sage.all import *
-sys.path.append("..")
-load("../dfinite_function.py")
-del allocatemem
+"""
+dfinite_function
+================
+
+"""
+
+from sage.rings.ring import Algebra
+
+class DFiniteFunctionRing(Algebra):
+    """
+    A Ring of Dfinite objects (functions or sequences)
+    """
+    _no_generic_basering_coercion = False
+    
+    def __init__(self, base_ring, name=None, element_class=None, category=None):
+        """
+        """
+        raise NotImplementedError
+        
+    def _element_constructor_(self, x=None, check=True, is_gen = False, construct=False, **kwds):
+        r"""
+        Convert ``x`` into this algebra, possibly non-canonically.
+        """
+        raise NotImplementedError
+
+    def is_integral_domain(self, proof = True):
+        """
+        """
+        return False
+
+    def is_noetherian(self):
+        """
+        """
+        return False
+            
+    def construction(self):
+        """
+        """
+        raise NotImplementedError
+
+    def _coerce_map_from_(self, P):
+        """
+        """
+        raise NotImplementedError
+
+    def _sage_input_(self, sib, coerced):
+        r"""
+        Produce an expression which will reproduce this value when
+        evaluated.
+        """
+        raise NotImplementedError
+
+    def _is_valid_homomorphism_(self, codomain, im_gens):
+        """
+        """
+        raise NotImplementedError
+
+    def __hash__(self):
+        # should be faster than just relying on the string representation
+        try:
+            return self._cached_hash
+        except AttributeError:
+            pass
+        h = self._cached_hash = hash((self.base_ring(),self.variable_name()))
+        return h
+
+    def _repr_(self):
+        try:
+            return self._cached_repr
+        except AttributeError:
+            pass
+        raise NotImplementedError
+
+    def _latex_(self):
+        """
+        """
+        raise NotImplementedError
+
+    def base_extend(self, R):
+        """
+        """
+        raise NotImplementedError
+
+    def ore_algebra(self):
+        """
+        """
+        raise NotImplementedError
+                
+    def characteristic(self):
+        """
+        Return the characteristic of this Ore algebra, which is the
+        same as that of its base ring.
+        """
+        return self.base_ring().characteristic()
+
+    def is_finite(self):
+        """
+        Return False since polynomial rings are not finite (unless the base
+        ring is 0.)
+        """
+        R = self.base_ring()
+        if R.is_finite() and R.order() == 1:
[…]

### 2013-05-27 doc/ore_algebra.py

@@ -1,5 +1,1720 @@
 
-from sage.all import *
-sys.path.append("..")
-load("../ore_algebra.py")
-del allocatemem
+"""
+ore_algebra
+===========
+
+
+
+"""
+
+######### development mode ###########
+
+try:
+    if sys.modules.has_key('nullspace'):
+        del sys.modules['nullspace']
+except:
+    pass
+try:
+    if sys.modules.has_key('ore_operator'):
+        del sys.modules['ore_operator']
+except:
+    pass
+try:
+    if sys.modules.has_key('ore_operator_1_1'):
+        del sys.modules['ore_operator_1_1']
+except:
+    pass
+try:
+    if sys.modules.has_key('ore_operator_mult'):
+        del sys.modules['ore_operator_mult']
+except:
+    pass
+try:
+    if sys.modules.has_key('generalized_series'):
+        del sys.modules['generalized_series']
+except:
+    pass
+
+def taylor_shift_univ_int_poly(p, i):
+    ## assuming that p is an element of ZZ['x']
+    return p(p.parent().gen() + i)
+def taylor_shift_univ_int_ratfun(q, i):
+    ## assuming that q is an element of Fract(ZZ['x'])
+    num = taylor_shift_univ_int_poly(q.numerator(), i)
+    den = taylor_shift_univ_int_poly(q.denominator(), i)
+    return q.parent()(num, den, coerce=False, reduce=False)
+def taylor_shift_univ_modp_poly(p, i):
+    ## assuming that p is an element of GF(m)['x']
+    return p(p.parent().gen() + i)
+def taylor_shift_univ_modp_ratfun(q, i):
+    ## assuming that q is an element of Fract(GF(m)['x'])
+    num = taylor_shift_univ_modp_poly(q.numerator(), i)
+    den = taylor_shift_univ_modp_poly(q.denominator(), i)
+    return q.parent()(num, den, coerce=False, reduce=False)
+# if sage version is 5.10.beta3 or later, use taylor shift of flint2
+import datetime
+try:
+    v = version()
+except:
+    v = "2013-01-01"    
+if datetime.date(int(v[-10:-6]), int(v[-5:-3]), int(v[-2:])) >= datetime.date(2013, 05, 15):
+    try:
+        load("shift.spyx") # cython code defining 'taylor_shift_univ_int_poly' and 'taylor_shift_univ_modp_poly'
+    except:
+        pass
+
+#######################################
+
+from sage.structure.element import RingElement
[…]

### 2013-05-27 doc/ore_operator.py

@@ -1,5 +1,1610 @@
 
-from sage.all import *
-sys.path.append("..")
-load("../ore_operator.py")
-del allocatemem
+"""
+ore_operator
+============
+
+"""
+
+from sage.structure.element import RingElement, canonical_coercion
+from sage.rings.arith import gcd, lcm
+from sage.rings.rational_field import QQ
+from sage.rings.integer_ring import ZZ
+from sage.rings.infinity import infinity
+
+class OreOperator(RingElement):
+    """
+    An Ore operator. This is an abstract class whose instances represent elements of ``OreAlgebra``.
+
+    In addition to usual ``RingElement`` features, Ore operators provide coefficient extraction
+    functionality and the possibility of letting an operator act on another object. The latter
+    is provided through ``__call__``.
+
+    """
+
+    # constructor
+
+    def __init__(self, parent, is_gen = False, construct=False):
+        RingElement.__init__(self, parent)
+        self._is_gen = is_gen
+
+    def __copy__(self):
+        """
+        Return a "copy" of self. This is just self, since in Sage
+        operators are immutable this just returns self again.
+        """
+        return self
+
+    # action
+
+    def __call__(self, f, **kwds):
+        """
+        Lets ``self`` act on ``f`` and returns the result.
+        The meaning of the action corresponding to the generator
+        of the Ore algebra can be specified with a keyword arguments
+        whose left hand sides are the names of the generator and the
+        right hand side some callable object. If no such information
+        is provided for some generator, a default function is used.
+        The choice of the default depends on the subclass. 
+
+        The parent of ``f`` must be a ring supporting conversion
+        from the base ring of ``self``. (There is room for generalization.)
+
+        EXAMPLES::
+
+           # In differential operator algebras, generators acts as derivations
+           sage: R.<x> = QQ['x']
+           sage: A.<Dx> = OreAlgebra(R.fraction_field(), "Dx")
+           sage: (Dx^5)(x^5) # acting on base ring elements
+           120
+           sage: (x*Dx - 1)(x)
+           0
+           sage: RR = PowerSeriesRing(QQ, "x", 5)
+           sage: 1/(1-RR.gen())
+           1 + x + x^2 + x^3 + x^4 + O(x^5)
[…]

### 2013-05-27 doc/ore_operator_1_1.py

@@ -1,5 +1,3519 @@
 
-from sage.all import *
-sys.path.append("..")
-load("../ore_operator_1_1.py")
-del allocatemem
+"""
+ore_operator_1_1
+================
+
+Special classes for operators living in algebras with one generator with base rings that also have
+one generator.
+
+"""
+
+from sage.structure.element import RingElement, canonical_coercion
+from sage.rings.arith import gcd, lcm
+from sage.rings.rational_field import QQ
+from sage.rings.integer_ring import ZZ
+from sage.rings.infinity import infinity
+from sage.rings.qqbar import QQbar
+from sage.rings.polynomial.polynomial_ring_constructor import PolynomialRing
+
+from ore_operator import *
+from generalized_series import *
+from generalized_series import _generalized_series_shift_quotient, _binomial ## why not implied by the previous line?
+
+class UnivariateOreOperatorOverUnivariateRing(UnivariateOreOperator):
+    """
+    Element of an Ore algebra with a single generator and a commutative rational function field as base ring.     
+    """
+    # Overview of dependencies between degree and denominator bounding functions:
+    #
+    #   * _degree_bound()          ==> requires indicial_polynomial(1/x)
+    #
+    #   * _denominator_bound()     ==> requires dispersion() and implements Abramov's algorithm
+    #                                  Not universally correct. May have to be adapted by subclasses.
+    #
+    #   * indicial_polynomial(p)   ==> requires _coeff_list_for_indicial_polynomial when p is x or 1/x
+    #                                  ABSTRACT for other arguments p.
+    #
+    #   * dispersion(p)            ==> requires spread(p)
+    #
+    #   * _coeff_list_for_ind...   ==> ABSTRACT
+    #
+    #   * spread(p)                ==> ABSTRACT
+    #
+
+    def __init__(self, parent, *data, **kwargs):
+        super(UnivariateOreOperator, self).__init__(parent, *data, **kwargs)
+
+    def _normalize_base_ring(self):
+        """
+        Rewrites ``self`` into an operator from an algebra whose base ring is a univariate
+        polynomial ring over a field.
+
+        Returns a tuple ``(A, R, K, L)`` where
+
+         * ``L`` is the new operator
+ 
+         * ``A`` is the parent of ``L``
+
[…]

### 2013-05-27 generalized_series.py

@@ -47,7 +47,7 @@ def GeneralizedSeriesMonoid(base, x, type="continuous"):
 
     If the type is ``"discrete"``, the domain contains series objects of the form
 
-    `(x/e)^(x*u/v)\rho^x\exp\bigl(c_1 x^{1/(m*v)} +...+ c_{v-1} x^{1-1/(m*v)}\bigr)x^\alpha p(x^{-1/(m*v)},\log(x))`
+    `(x/e)^{x u/v}\rho^x\exp\bigl(c_1 x^{1/(m*v)} +...+ c_{v-1} x^{1-1/(m*v)}\bigr)x^\alpha p(x^{-1/(m*v)},\log(x))`
 
     where
 

### 2013-05-27 guessing.py

@@ -109,8 +109,8 @@ def guess(data, algebra, **kwargs):
 
     .. NOTE::
 
-    This method is designed to find equations for D-finite objects. It may exhibit strange
-    behaviour for objects which are holonomic but not D-finite. 
+      This method is designed to find equations for D-finite objects. It may exhibit strange
+      behaviour for objects which are holonomic but not D-finite. 
 
     EXAMPLES::
 
@@ -246,8 +246,8 @@ def guess_raw(data, A, order=-1, degree=-1, lift=None, solver=None, cut=25, ensu
 
     .. NOTE::
 
-    This is a low-level method. Don't call it directly unless you know what you
-    are doing. In usual applications, the right method to call is ``guess``.
+      This is a low-level method. Don't call it directly unless you know what you
+      are doing. In usual applications, the right method to call is ``guess``.
 
     EXAMPLES::
 

### 2013-05-27 ore_algebra.py

@@ -53,7 +53,10 @@ def taylor_shift_univ_modp_ratfun(q, i):
     return q.parent()(num, den, coerce=False, reduce=False)
 # if sage version is 5.10.beta3 or later, use taylor shift of flint2
 import datetime
-v = version()
+try:
+    v = version()
+except:
+    v = "2013-01-01"    
 if datetime.date(int(v[-10:-6]), int(v[-5:-3]), int(v[-2:])) >= datetime.date(2013, 05, 15):
     try:
         load("shift.spyx") # cython code defining 'taylor_shift_univ_int_poly' and 'taylor_shift_univ_modp_poly'

### 2013-05-27 ore_operator_1_1.py

@@ -106,10 +106,10 @@ class UnivariateOreOperatorOverUnivariateRing(UnivariateOreOperator):
 
         .. NOTE::
 
-        - Even if no ``rhs`` is given, the output will be a list of tuples ``[(p1,), (p2,),...]``
-          and not just a list of plain polynomials.
-        - If no ``degree`` is given, a basis of all the polynomial solutions is returned.
-          This feature may not be implemented for all algebras. 
+          - Even if no ``rhs`` is given, the output will be a list of tuples ``[(p1,), (p2,),...]``
+            and not just a list of plain polynomials.
+          - If no ``degree`` is given, a basis of all the polynomial solutions is returned.
+            This feature may not be implemented for all algebras. 
 
         EXAMPLES::
 
@@ -191,12 +191,12 @@ class UnivariateOreOperatorOverUnivariateRing(UnivariateOreOperator):
 
         .. NOTE::
 
-        - Even if no ``rhs`` is given, the output will be a list of tuples ``[(p1,), (p2,),...]``
-          and not just a list of plain polynomials.
-        - If no ``denominator`` is given, a basis of all the rational solutions is returned.
-          This feature may not be implemented for all algebras. 
-        - If no ``degree`` is given, a basis of all the polynomial solutions is returned.
-          This feature may not be implemented for all algebras. 
+          - Even if no ``rhs`` is given, the output will be a list of tuples ``[(p1,), (p2,),...]``
+            and not just a list of plain polynomials.
+          - If no ``denominator`` is given, a basis of all the rational solutions is returned.
+            This feature may not be implemented for all algebras. 
+          - If no ``degree`` is given, a basis of all the polynomial solutions is returned.
+            This feature may not be implemented for all algebras. 
 
         EXAMPLES::
 
@@ -2083,7 +2083,7 @@ class UnivariateRecurrenceOperatorOverUnivariateRing(UnivariateOreOperatorOverUn
 
         These are solutions of the form
 
-          `(x/e)^(x*u/v)\rho^x\exp\bigl(c_1 x^{1/m} +...+ c_{v-1} x^{1-1/m}\bigr)x^\alpha p(x^{-1/m},\log(x))`
+          `(x/e)^{x u/v}\rho^x\exp\bigl(c_1 x^{1/m} +...+ c_{v-1} x^{1-1/m}\bigr)x^\alpha p(x^{-1/m},\log(x))`
 
         where
 

### 2013-05-27 doc/guessing.py
bugs in nullspace.py fixed.
now "sage -t nullspace.py" runs without complaints.

two missing files added to doc directory

@@ -1,0 +1,892 @@
+
+"""
+guessing
+========
+
+"""
+
+######### development mode ###########
+
+try:
+    if sys.modules.has_key('ore_algebra'):
+        del sys.modules['ore_algebra']
+except:
+    pass
+
+#######################################
+
+from sage.rings.integer_ring import ZZ
+from sage.rings.rational_field import QQ
+from sage.rings.finite_rings.all import GF, is_FiniteField
+from sage.matrix.constructor import Matrix, matrix
+from sage.rings.arith import xgcd
+from sage.parallel.decorate import parallel
+from sage.rings.polynomial.polynomial_ring import is_PolynomialRing
+
+from ore_algebra import *
+
+from datetime import datetime
+import nullspace
+import math
+
+def guess_rec(data, n, S, **kwargs):
+    """
+    Shortcut for ``guess`` applied with an Ore algebra of shift operators in `S` over `K[n]`
+    where `K` is the parent of ``data[0]``.
+
+    See the docstring of ``guess`` for further information.    
+    """
+    R = data[0].parent()[n]; x = R.gen()
+    return guess(data, OreAlgebra(R, (S, {n:n+R.one()}, {})), **kwargs)
+
+def guess_deq(data, x, D, **kwargs):
+    """
+    Shortcut for ``guess`` applied with an Ore algebra of differential operators in `D` over `K[x]`
+    where `K` is the parent of ``data[0]``.
+
+    See the docstring of ``guess`` for further information.    
+    """
+    R = data[0].parent()[x]; x = R.gen()
+    return guess(data, OreAlgebra(R, (D, {}, {x:R.one()})), **kwargs)
+
+def guess_qrec(data, qn, Q, q, **kwargs):
+    """
+    Shortcut for ``guess`` applied with an Ore algebra of `q`-recurrence operators in `Q` over `K[qn]`
+    where `K` is the parent of `q`.
+
+    See the docstring of ``guess`` for further information.    
+    """
+    R = q.parent()[qn]; x = R.gen()
+    return guess(data, OreAlgebra(R, (Q, {qn:q*qn}, {qn:R.one()})), **kwargs)
+
+def guess(data, algebra, **kwargs):
+    """
+    Searches for an element of the algebra which annihilates the given data.
+
+    INPUT:
+
+    - ``data`` -- a list of elements of the algebra's base ring's base ring `K` (or at least
[…]

### 2013-05-27 generalized_series.py
bugfixes in generalized_series.py
now "sage -t generalized_series.py" runs without complaints.

@@ -85,28 +85,30 @@ def GeneralizedSeriesMonoid(base, x, type="continuous"):
 
       sage: G = GeneralizedSeriesMonoid(QQ, 'x')
       sage: G
-      Differential monoid of generalized series in x over Rational Field
+      Monoid of continuous generalized series in x over Rational Field
+      sage: x = QQ['x'].gen()
       sage: G(x+2*x^3 + 4*x^4 + O(x^5))
       x*(1 + 2*x^2 + 4*x^3 + O(x^4))
       sage: G(x+2*x^3 + 4*x^4 + O(x^5), ramification=2)
       x^(1/2)*(1 + 2*x^(2/2) + 4*x^(3/2) + O(x^(4/2)))
       sage: G(x+2*x^3 + 4*x^4 + O(x^5), ramification=3)
       x^(1/3)*(1 + 2*x^(2/3) + 4*x^(3/3) + O(x^(4/3)))
-      sage: _.derivative()
+      sage: f = _
+      sage: f.derivative()
       x^(-2/3)*(1/3 + 2*x^(2/3) + 16/3*x^(3/3) + O(x^(4/3)))
-      sage: _*__
+      sage: _*f
       x^(-1/3)*(1/3 + 8/3*x^(2/3) + 20/3*x^(3/3) + O(x^(4/3)))
       sage: (G(1+x, ramification=2)*G(1+x, ramification=3)).ramification()
       6
       sage: K = QQ.extension(x^2-2, 'a'); a = K.gen()
       sage: a*G(x)
-      a*x
+      x*a
       sage: _.parent()
-      Differential monoid of generalized series in x over Number Field in a with defining polynomial x^2 - 2
+      Monoid of continuous generalized series in x over Number Field in a with defining polynomial x^2 - 2
       sage: G(x).base_extend(x^3+5, 'b')
       x
       sage: _.parent()
-      Differential monoid of generalized series in x over Number Field in b with defining polynomial x^3 - 5
+      Monoid of continuous generalized series in x over Number Field in b with defining polynomial x^3 + 5
 
     """
     M = GeneralizedSeriesMonoid_class(base, x, type)
@@ -303,14 +305,14 @@ class GeneralizedSeriesMonoid_class(Parent):
 
            sage: G = GeneralizedSeriesMonoid(QQ, 'x', 'continuous')
            sage: G
-           Differential monoid of generalized series in x over Rational Field
+           Monoid of continuous generalized series in x over Rational Field
[…]

### 2013-05-28 guessing.py
bugs in guessing.py fixed.
now "sage -t guessing.py" runs without complaints.

@@ -120,8 +120,7 @@ def guess(data, algebra, **kwargs):
       sage: R.<t> = QQ['t']
       sage: rec = guess([1/(i+t) + t^i for i in xrange(100)], OreAlgebra(R['n'], 'Sn'))
       sage: rec
-      ((-51891840*t + 51891840)*n^2 + (-103783680*t^2 - 51891840*t + 103783680)*n - 51891840*t^3 - 103783680*t^2)*Sn^2 + ((51891840*t^2 - 51891840)*n^2 + (103783680*t^3 + 155675520*t^2 - 103783680*t - 51891840)*n + 51891840*t^4 + 155675520*t^3 + 51891840*t^2 - 51891840*t)*Sn + (-51891840*t^2 + 51891840*t)*n^2 + (-103783680*t^3 + 51891840*t)*n - 51891840*t^4 - 51891840*t^3 + 51891840*t^2
-      sage: 
+      ((t - 1)*n^2 + (2*t^2 + t - 2)*n + t^3 + 2*t^2)*Sn^2 + ((-t^2 + 1)*n^2 + (-2*t^3 - 3*t^2 + 2*t + 1)*n - t^4 - 3*t^3 - t^2 + t)*Sn + (t^2 - t)*n^2 + (2*t^3 - t)*n + t^4 + t^3 - t^2
     
     """
 
@@ -884,9 +883,13 @@ def _rat_recon(a, m, u=None):
         if s < score:
             out = (p, q); score = s
             if score < early_termination_bound:
-                return out
+                break
 
     if score < bound:
-        return out
+        if K is ZZ:
+            return out
+        else:
+            lc = out[1].leading_coefficient()
+            return (out[0]/lc, out[1]/lc)
     else:
         raise ArithmeticError

### 2013-05-28 ore_operator.py

@@ -515,7 +515,13 @@ class OreOperator(RingElement):
         num = self.numerator().primitive_part(proof=proof)
         c = num.leading_coefficient()
         while not c.is_unit() and c.parent() is not c.parent().base_ring():
-            c = c.leading_coefficient()
+            try:
+                c = c.leading_coefficient()
+            except:
+                try:
+                    c = c.lc()
+                except:
+                    break
         if c.is_unit():
             return self.parent()((~c)*num)
         else:

### 2013-05-28 paper/main.tex
some cleanup on the docstrings to support automated html doc generation.
paper: introduction

@@
-\documentclass[a4paper,draft]{amsart}
+\documentclass{amsart}
 \usepackage{amsmath}
@@
 \usepackage{a4wide}
+\usepackage{sagetex}
 
@@
 
-motivation
+In computer algebra, objects are often described implicitly through equations
+they satisfy. For example, the exponential function $\exp(x)$ is uniquely
+specified by the linear differential equation $f'(x)-f(x)=0$ and the initial
+value $f(0)=1$.  Likewise, the sequence $F_n$ of Fibonacci numbers is uniquely
+determined by the linear recurrence $F_{n+2}-F_{n+1}-F_n=0$ and the two initial
+values $F_0=0$,~$F_1=1$.  Especially for representing functions or sequences
+that cannot be expressed in ``closed form'', the differential or difference
+equations they may satisfy provide an attractive way to store them on the
+computer. The question is then how to calculate with objects which are given in
+this form.
 
-similar packages: gfun, mallinger, q-gfun, future: mgfun, holonomicfunctions
+Algorithms for Ore algebras provide a systematic answer to this
+question~\cite{..,..}.  Invented in the early 20th century~\cite{..} with the
+objective of providing a unified theory for various kinds of linear operators,
+they have been used for many years in computer algebra systems, for example in
+the Maple packages gfun~\cite{..} and mgfun~\cite{..}, in the Mathematica
+packages by Mallinger~\cite{..} and Koutschan~\cite{..,..}, and
+elsewhere~\cite{..}.
 
-overview over the package
+The purpose of this paper is to introduce an implementation of a collection of
+algorithms related to Ore algebras for the computer algebra system
+Sage~\cite{..}. It is addressed to first-time users who are already familiar
+with Sage, and with the theory of Ore algebras and their use for doing symbolic
+computation for special functions. Readers unfamiliar with Sage are referred
+to~\cite{..}, and readers unfamiliar with Ore algebras may wish to consult the
+recent tutorial~\cite{..} and the references given there for an introduction to
+the subject.
 
-short demo
+At the time of writing, the package we describe here is still under construction
+and has not yet been incorporated into the official Sage distribution. Readers
+who want to try it out are invited to download the current development version
+from [URL] and are encouraged to send us bug reports and other comments. We hope
+that the community will find the code useful.
 
-outlook, pending tasks (multivariate!)
+SHORT DEMO HOW TO INSTALL AND LAUNCH IT (TO BE DETERMINED)
 
-how to get it and how to install it.
+DISCLAIMER: SO FAR WE MAINLY SUPPORT UNIVARIATE OPERATORS.
+
+HOWEVER: FOR THESE WE HAVE MORE OPERATIONS THAT GFUN OR MALLINGER.
 
 \section{Ore Algebras and Ore Polynomials}
+
+\begin{sageexample}
+  sage: x = 5
+  sage: x + 1
+\end{sageexample}
 
@@
 arithmetic, lclm, gcrd, action, coercion and conversion, pretty printing, coefficient extraction,
-closure properties
+closure properties, associate solutions.
 

### 2013-05-29 ore_algebra.py
bugs in ore_algebra.py fixed.
now "sage -t ore_algebra.py" runs with only a single complaint,
which however seems to be a factorization bug in Sage 5.8.

@@ -47,16 +47,18 @@ def taylor_shift_univ_modp_poly(p, i):
     ## assuming that p is an element of GF(m)['x']
     return p(p.parent().gen() + i)
 def taylor_shift_univ_modp_ratfun(q, i):
-    ## assuming that q is an element of Fract(GF(m)['x'])
+    ## assuming that q is an element of Frac(GF(m)['x'])
     num = taylor_shift_univ_modp_poly(q.numerator(), i)
     den = taylor_shift_univ_modp_poly(q.denominator(), i)
     return q.parent()(num, den, coerce=False, reduce=False)
+
 # if sage version is 5.10.beta3 or later, use taylor shift of flint2
 import datetime
 try:
     v = version()
 except:
     v = "2013-01-01"    
+
 if datetime.date(int(v[-10:-6]), int(v[-5:-3]), int(v[-2:])) >= datetime.date(2013, 05, 15):
     try:
         load("shift.spyx") # cython code defining 'taylor_shift_univ_int_poly' and 'taylor_shift_univ_modp_poly'
@@ -89,7 +91,7 @@ def is_OreAlgebra(A):
     """
     return isinstance(A, OreAlgebra_generic)
 
-def _is_suitable_base_ring(R):
+def is_suitable_base_ring(R):
     """
     Checks whether `R` is suitable as base ring of an Ore algebra.
     This is the case if and only if: (a) `R` is one of `ZZ`, `QQ`,
@@ -102,10 +104,10 @@ def _is_suitable_base_ring(R):
 
     EXAMPLES::
 
-       sage: R = GF(1091)['x,y'].fraction_field()['z']
-       sage: _is_suitable_base_ring(R)
+       sage: R = GF(1091)['x, y'].fraction_field()['z']
+       sage: is_suitable_base_ring(R)
        True
-       sage: _is_suitable_base_ring(GF(9, 'a'))
+       sage: is_suitable_base_ring(GF(9, 'a'))
        False
     
     """
@@ -113,13 +115,13 @@ def _is_suitable_base_ring(R):
     if (p == 0 and (R is ZZ or R is QQ or is_NumberField(R))) or (p > 0 and R is GF(p)):
         return True
     elif is_FractionField(R):
-        return _is_suitable_base_ring(R.ring())
+        return is_suitable_base_ring(R.ring())
     elif is_PolynomialRing(R) or is_MPolynomialRing(R):
[…]

### 2013-05-29 ore_algebra.py
implicit generator declaration in OreAlgebra:
now you can say

  A.<Dx> = OreAlgebra(ZZ['x'])

without having to specify 'Dx' as second argument on the right,
similar as

  R.<x> = PolynomialRing(QQ)

@@ -770,6 +770,11 @@ def OreAlgebra(base_ring, *generators, **kwargs):
     if not is_suitable_base_ring(R):
         raise TypeError, "The base ring is not of the required form."
     if len(gens) == 0:
+        try:
+            gens = list(kwargs['names'])
+        except AttributeError:
+            raise TypeError, "There must be at least one generator"
+    if len(gens) == 0:
         raise TypeError, "There must be at least one generator"
 
     product_rules = []

### 2013-06-03 ore_operator.py
bugfix in a docstring

@@ -73,7 +73,7 @@ class OreOperator(RingElement):
            x^3 - 3*x^2 + 2*x
            sage: factor(_)
            (x - 2) * (x - 1) * x
-           sage: (Sx - 1)(1/4*x*(x-1)*(x-2)*(x-3), Sx=lambda p:p(2*x)) # let Sx act as q-shift
+           sage: (Sx - 1)(1/4*x*(x-1)*(x-2)*(x-3), action=lambda p:p(2*x)) # let Sx act as q-shift
            15/4*x^4 - 21/2*x^3 + 33/4*x^2 - 3/2*x
 
         """

### 2013-06-03 paper/main.tex
paper: section 2

@@
 \def\e{\mathrm{e}}
+
+\def\Bold#1{\mathbf{#1}}
 
@@
 Sage~\cite{..}. It is addressed to first-time users who are already familiar
-with Sage, and with the theory of Ore algebras and their use for doing symbolic
-computation for special functions. Readers unfamiliar with Sage are referred
+with Sage, and with the theory of Ore algebras and its use for doing symbolic
+computation related to special functions. Readers unfamiliar with Sage are referred
 to~\cite{..}, and readers unfamiliar with Ore algebras may wish to consult the
@@
 and has not yet been incorporated into the official Sage distribution. Readers
-who want to try it out are invited to download the current development version
-from [URL] and are encouraged to send us bug reports and other comments. We hope
+who want to try it out are invited to download the current latest version from 
+[URL] and are encouraged to send us bug reports and other comments. We hope 
 that the community will find the code useful.
 
-SHORT DEMO HOW TO INSTALL AND LAUNCH IT (TO BE DETERMINED)
-
-DISCLAIMER: SO FAR WE MAINLY SUPPORT UNIVARIATE OPERATORS.
-
-HOWEVER: FOR THESE WE HAVE MORE OPERATIONS THAT GFUN OR MALLINGER.
-
-\section{Ore Algebras and Ore Polynomials}
+Creating an Ore algebra object is easy. The following instructions show how to 
+load the code and then create an Ore algebra $A$ of linear differential operators
+and an Ore algebra $B$ of recurrence operators. Observe the correct application
+of the respective commutation rules in both cases.
 
 \begin{sageexample}
-  sage: x = 5
-  sage: x + 1
+  sage: load("ore_algebra.py");  #### TO BE FIXED ####
+  sage: R.<x> = PolynomialRing(ZZ); A.<Dx> = OreAlgebra(R)
+  sage: A
+  sage: A.random_element()
+  sage: Dx*x
+  sage: B.<Sx> = OreAlgebra(R)
+  sage: B
+  sage: Sx*x
 \end{sageexample}
 
-\subsection{In Theory}
+More details on the construction of Ore algebras are given in the following section.
+Also Ore algebras with several generators are already supported. However, so far we offer
+no functionality beyond addition and multiplication for these. More functionality 
+is available for algebras with one generator. We describe some of it in 
+Section~\ref{sec:3}. Even more functionality has been implemented for special 
+algebras like the ones created above, which are most frequently used in practice. 
+See Section~\ref{sec:4} for an overview. Finally, the package also comes with a 
+fine tuned guessing engine, whose main features we describe in Section~\ref{sec:5}.
 
-basic facts and definitions
+\section{Ore Algebras}
 
-reference to literature
+An Ore algebra is determined by a base ring and a finite number of generators.
+In the examples above, the base ring was $\Bold{Z}[x]$, and the generators were
+$\mathrm{Dx}$ and $\mathrm{Sx}$, respectively. If no other information is provided
+in the arguments, the \verb|OreAlgebra| constructor chooses the nature of the 
+generators according to their name: a generator called $\mathrm{Dt}$ represents
+the standard derivation $d/dt$ acting on the generator $t$ of the base ring, 
+a generator called $Sn$ represents the standard shift operator sending the
+generator $n$ of the base ring to $n+1$.
 
-\subsection{In Sage}
+For this way of generating algebras, generator names must be composed of
+one of the following single-letter prefixes followed by the name of a generator
+of the base ring.
 
-ground rings
+\begin{center}
+  \begin{tabular}{|c|c|c|}\hline
+    Prefix & Name & Commutation rule \\\hline
+     D & Standard derivation $d/dx$ & $\mathrm{Dx}\,x=x\,\mathrm{Dx}+1$ \\
+     S & Standard shift $x\leadsto x+1$ & $\mathrm{Sx}\,x=(x+1)\,\mathrm{Sx}$ \\
+     T or $\Theta$ & Eulerian derivation $x\,d/dx$ & $\mathrm{Tx}\,x=x\,\mathrm{Tx}+x$ \\
+     F or $\Delta$ & Forward difference $\Delta_x$ & $\mathrm{Fx}\,x=(x+1)\mathrm{Fx}+1$ \\
+     Q & $q$-shift $x\leadsto q\,x$ & $\mathrm{Qx}\,x=q\,x\,\mathrm{Qx}$ \\
+     J & $q$-derivation (``Jackson derivation'') & $\mathrm{Jx}\,x=q\,x\,\mathrm{Jx}+1$ \\ 
+     C & commutative generator & $\mathrm{Cx}\,x = x\,\mathrm{Cx}$ \\\hline
+  \end{tabular}
+\end{center}
 
-general sigma and delta
+For the $q$-shift and the $q$-derivation, the base ring must contain an element~$q$.
+The element playing the role of $q$ can be specified as an optional argument.
 
-predefined shortcuts: D, T, S, F, J, Q, C
+\begin{sageexample}
+  sage: R.<x> = PolynomialRing(ZZ['q'])
+  sage: A.<Qx> = OreAlgebra(R)
+  sage: Qx*x
+  sage: A.<Qx> = OreAlgebra(R, q=2)
+  sage: Qx*x
+\end{sageexample}
 
-\section{General Methods}
+In general, the commutation rules of a generator~$X$ of an Ore algebra~$A$ with
+base ring~$R$ are governed by two maps, $\sigma\colon R\to R$ and $\delta\colon R\to R$,
+where $\sigma$ is a ring endomorphism (i.e., $\sigma(a+b)=\sigma(a)+\sigma(b)$ and
+$\sigma(ab)=\sigma(a)\sigma(b)$ for all $a,b\in R$) and $\delta$ is a skew-derivation
+for $\sigma$ (i.e., $\delta(a+b)=\delta(a)+\delta(b)$ and $\delta(ab)=\delta(a)b
++\sigma(a)\delta(b)$ for all $a,b\in R$). With two such maps being given, the
+generator $X$ satisfies the commutation rule $Xa=\sigma(a)X+\delta(a)$ for every
+$a\in R$. If there is more than one generator, then each of them has its own pair
+of maps $\sigma,\delta$. Different generators commute with each other; 
+noncommutativity only takes place ``between'' generators and base ring elements.
 
-arithmetic, lclm, gcrd, action, coercion and conversion, pretty printing, coefficient extraction,
-closure properties, associate solutions.
+It is possible to create an Ore algebra with user specified commutation rules.
+In this form, each generator must be declared by a tuple $(X,\sigma,\delta)$, 
+where $X$ is the name of the generator (a string), and $\sigma$ and $\delta$
+are dictionaries which contain the images of the generators of the base
+ring under the respective map. Here is how to specify an algebra of difference
+operators in this way:
 
-\section{Special Methods Standard Algebras}
+\begin{sageexample}
+  sage: R.<x> = ZZ['x']
+  sage: A = OreAlgebra(R, ('X', {x:x+1}, {x:1}))
+  sage: X = A.gen()
+  sage: X*x
+\end{sageexample}
+
+As another example, here is how to define an algebra of differential operators 
+whose base ring is a differential field $K=\set Q(x,y,z)$ where $y$ represents
+$\exp(x)$ and $z$ represents $\log(x)$:
+
+\begin{sageexample}
+  sage: K = ZZ['x','y','z'].fraction_field()
+  sage: x,y,z = K.gens()
+  sage: A = OreAlgebra(K, ('D', {}, {x:1, y:y, z:1/x}))
+  sage: D = A.gen()
+  sage: D*x, D*y, D*z
+\end{sageexample}
+
+In the dictionary specifying~$\sigma$, omitted generators are understood to be
+mapped to themselves, so that \verb|{}| in the definition of $A$ in the example
+above is equivalent to \verb|{x:x,y:y,z:z}|.  Likewise, in the dictionaries
+specifying~$\delta$, omitted generators are understood to be mapped to zero.
+
+For Ore algebras with several generators, it is possible to mix specifications
+of generators via triples $(X,\sigma,\delta)$ with generators using the naming
+convention shortcuts as explained before. Continuing the previous example,
+here is a way to define an algebra $A$ over $K$ with two generators, a $D$ 
+that behaves like before, and in addition an $Sx$ which acts like the standard
+shift on~$x$.
+
+\begin{sageexample}
+  sage: A = OreAlgebra(K, ('D', {}, {x:1, y:y, z:1/x}), 'Sx')
+  sage: D, Sx = A.gens()
+  sage: D*x, Sx*x
+\end{sageexample}
+
+Not every ring is suitable as base ring of an Ore algebra. Base rings must
+themselves be polynomial rings (univariate or multivariate), or fraction 
+fields of polynomial rings. Their base rings in turn may be either $\set Z$,
+$\set Q$, a prime field $GF(p)$, or a number field $\set Q(\alpha)$, or
+--- recursively --- some ring which itself would be suitable as base ring
+of an Ore algebra.
+
+\begin{sageexample}
+  sage: ZZ['x'].fraction_field()['y','z'] ### OK
+  sage: GF(1091)['x','y','z']['u'] ### OK
+  sage: ( ZZ['x','y','z'].quotient_ring(x^2+y^2+z^2-1) )['u'] ### not OK
+  sage: GF(9, 'a')['x'] ### not OK
+\end{sageexample}
+
+Note that the maps $\sigma$ and $\delta$ must leave all the elements of
+the base ring's base ring fixed. They may only have nontrivial images for
+the top level generators.
+
+The constituents of an Ore algebra $A$ can be accessed through the methods
+summarized in the following table. Further methods can be found in the 
+documentation.
+
+\begin{center}
+  \begin{tabular}{|l|p{.55\hsize}|}
+    \hline
+    Method name & short description \\\hline
+    \verb|associated_commutative_algebra()| & returns a polynomial ring with the
+       same base ring as $A$ and whose generators are named like the generators
+       of~$A$\\
+    \verb|base_ring()| & returns the base ring of $A$\\
+    \verb|delta(i)| & returns a callable object representing the delta map
+       associated to the $i$th generator (default: $i=0$) \\
+    \verb|gen(i)| & returns the $i$th generator (default: $i=0$)\\
+    \verb|sigma(i)| & returns a callable object representing the sigma map
+       associated to the $i$th generator (default: $i=0$) \\
+    \verb|var(i)| & returns the name of the $i$th generator (default: $i=0$)\\\hline
+  \end{tabular}
+\end{center}
+
+\smallskip
+
+Examples:
+
+\begin{sageexample}
+  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
+  sage: A
+  sage: A.associated_commutative_algebra()
+  sage: A.base_ring()
+  sage: A.gen()
+  sage: s = A.sigma(); d = A.delta(); 
+  sage: s(x^5), d(x^5)
+\end{sageexample}
+
+\section{Ore Polynomials}\label{sec:3}
+
+arithmetic, lclm, gcrd, action, coercion and conversion, pretty printing, 
+coefficient extraction, closure properties, associate solutions.
+
+\section{Special Methods for Standard Algebras}\label{sec:4}
 
@@
 
-\section{Guessing}
+\section{Guessing}\label{sec:5}
 
@@
  \bibliographystyle{plain}
- \bibliography{all}
+ \bibliography{main}
 

### 2013-06-03 ore_algebra.py

@@ -1075,8 +1075,8 @@ class OreAlgebra_generic(Algebra):
         except AttributeError:
             pass
         r = self.base_ring()._latex_() + "\\langle "
-        for x in self._gens:
-            r = r + x[0] + ", "
+        for x in self.gens():
+            r = r + x._latex_() + ", "
         r = r[:-2] + "\\rangle "
         self._cached_latex = r
         return r

### 2013-06-07 paper/main.tex
paper, section 5: guessing

@@
 \usepackage{sagetex}
+\usepackage{tikz}
 
@@
 \begin{sageexample}
-  sage: load("ore_algebra.py");  #### TO BE FIXED ####
+  sage: load("ore_algebra.py", "guessing.py");  #### TO BE FIXED ####
   sage: R.<x> = PolynomialRing(ZZ); A.<Dx> = OreAlgebra(R)
@@
 
-\section{Further Examples}
+[[CITE RUBEY.]]
 
-Some longer session showing how to solve a meaningful problem by using several features of the code. Ideas?
+Guessing is one of the most popular features of software packages like gfun.  It
+is a technique for detecting patterns in sequences of numbers, which is often
+used in combinatorics. The basic idea is simple. If we are given a finite array
+of numbers, though of as the first terms of an infinite sequence, we may want to
+know whether this sequence satisfies a recurrence. The algorithm behind a
+guessing engine searches for small equations matching the given data.
+Generically, no such equations exist, so if some are found, it is fair to
+``guess'' that they are in fact valid equations for the whole infinite sequence.
 
-\subsection{}
+We provide a guessing function which takes as input a list of terms and an Ore
+algebra, and returns as output an operator which matches the given data and
+which, in some measure, would be unlikely to exist for random data.
 
-\subsection{}
+\begin{sageexample}
+  sage: data = [ 0, 1, 1, 2, 3, 5, 8, 13, 21, 34, 55 ]
+  sage: L = guess(data, OreAlgebra(ZZ['n'], 'Sn'))
+  sage: L
+  sage: L(data)
+  sage: M = guess(data, OreAlgebra(ZZ['x'], 'Dx'))
+  sage: M
+  sage: M(x/(1-x-x^2))
+\end{sageexample}
+
+If an algebra of differential operators is supplied as second argument, the data
+is understood as the first few coefficients of a power series. The output
+operator is expected to have this power series as solution.
+
+It can happen that the procedure is unable to find an operator matching the 
+given data. In this case, an exception is raised. There are two possible 
+explanations for such an event. Either the sequence in question does not 
+satisfy any equations, or it does but the equations are so big that more
+data is needed to detect them.
+
+A number of options are available for governing the search for relations.  In
+order to explain them, we first need to give some details on the underlying
+algorithms. For simplicity of language, we restrict here to the case of
+recurrence operators. The situation for differential operators is very similar.
+
+For the most typical situations, there are two important hyperbolas. One
+describes the region in the $(r,d)$-plane consisting of all points for which
+there exists an operator of order~$r$ and degree~$d$ truly satisfied by the
+sequence in question. (See~\cite{..} for an explanation why the boundary of this
+region is usually a hyperbola.) The second describes the region of all points
+$(r,d)$ for which an operator of order~$r$ and degree~$d$ can be detected when
+$N$ terms are provided as input. This region is determined by the requirement
+$(r+1)(d+2)<N$.
+
+The method tests a sequence of points $(r_1,d_1)$, $(r_2,d_2)$, \dots\ right
+below this second parabola. Success at a point $(r_i,d_i)$ means that some
+evidence for an operator of order $\leq r_i$ and degree $\leq d_i$ has been found. 
+This operator however is not explicitly computed. Instead, the method uses the
+partial information found about this operator to calculate an operator which with 
+high probability is the minimal order operator satisfied by the sequence in question. 
+This operator is usually more interesting than the one at $(r_i,d_i)$, and its 
+computation is usually more efficient.
+
+Using the option \verb|path|, the user can specify a list of points $(r_i,d_i)$
+which should be used instead of the standard path. By setting the options \verb|min_degree|,
+\verb|max_degree|, \verb|min_order|, \verb|max_order|, all points $(r,d)$ of the path
+are discarded for which $r$ or $d$ is not within the specified bounds. These options
+can be used to accelerate the search in situations where the user has some knowledge
+(or intuition) about the size of the expected equations.
+
+\hangindent=-6.3cm\hangafter=-14\leavevmode
+\smash{\raise-5.3cm\rlap{\kern\hsize\kern-5.1cm\begin{tikzpicture}[scale=.45]
+  \draw[->] (10.5,0) node {$r$} (-.2,0) -- (10, 0) ;
+  \draw[->] (0,10.5) node {$d$} (0,-.2) -- (0, 10) ;
+  \draw (1.2, 3.5) -- (1.4, 4) (1.2, 3) -- (1.6, 4) (1.2, 2.5) -- (1.8, 4) (1.2, 2) -- (2, 4)
+        (1.4, 2) -- (2.2, 4) (1.6, 2) -- (2.4, 4) (1.8, 2) -- (2.6, 4) (2, 2) -- (2.8, 4) 
+        (2.2, 2) -- (3, 4) (2.4, 2) -- (3.2, 4) (2.6, 2) -- (3.4, 4) (2.8,2)--(3.6,4) (3,2)--(3.8,4)
+        (3.2, 2) -- (4, 4) (3.4, 2) -- (4.2, 4) (3.6, 2) -- (4.4, 4) (3.8,2)--(4.6,4) (4,2)--(4.8,4)
+        (4.2, 2) -- (5, 4) (4.4, 2) -- (5.2, 4) (4.6, 2) -- (5.4, 4) (4.8,2)--(5.6,4) (5,2)--(5.8,4); 
+  \fill[lightgray] (.75, 10) .. controls (1.5,2.5) and (2.5,1.5) .. (10, .75) -- (10, 10) -- (.75, 10);
+  \draw[thick] (2, 10) .. controls (2, 1.2) and (2, 1.2) .. (10, 1.2); 
+  \draw (-.2, 2) node {\vbox{\llap{\footnotesize min }\kern-3pt\llap{\footnotesize degree }}} -- (10, 2) 
+        (-.2, 4) node {\vbox{\llap{\footnotesize max }\kern-3pt\llap{\footnotesize degree }}} -- (10, 4);
+  \draw (1.2, -.2) node {\hbox to0pt{\rule{0pt}{2em}\hss\footnotesize min order\hss}} -- (1.2, 10)
+        (5, -.2) node {\hbox to0pt{\rule{0pt}{2em}\hss\footnotesize max order\hss}} -- (5, 10);
+  \fill (2.5, 3) circle(4pt) node[right] {\rule[-1em]{0pt}{0pt}$A$} (2, 9) circle(4pt) node[right] {$B$} ;
+\end{tikzpicture}}}%
+The figure on the right illustrates the typical situation for guessing problems
+that are not too small and not too artificial. The light gray region indicates
+the area which is not accessible with the given amount of data. Only the points
+$(r,d)$ below it can be tested for an operator of order~$r$ and degree~$d$ that
+fits to the given data. Let's assume that operators exist on and above the solid
+black hyperbola. The user will usually not know this curve in advance but may
+have some expectations about it and can restrict the search accordingly, for
+example to the dashed area shown in the figure. The method will detect the
+existence of an operator, say at point~$A$, and construct from the information
+gained at this point an operator of minimal possible order, which may correspond
+to point~$B$.  This operator is returned as output. Note that the degree of the
+output may exceed the value of \verb|max_degree|, and its order may be smaller
+than $\verb|min_order|$:
+
+\begin{sageexample}
+sage: data = [(n+1)^10*2^n + 3^n for n in xrange(200)]
+sage: L = guess(data, OreAlgebra(ZZ['n'],'Sn'), min_order=3, max_degree=5)
+sage: L.order(), L.degree()
+\end{sageexample}
+
+In order to test a specific point $(r,d)$, the data array must contain at least $(r+1)(d+2)$ terms. 
+If it has more terms, the guess becomes more trustworthy, but also the computation time increases.
+By setting the option \verb|ensure| to a positive integer~$e$, the user can request that only 
+such points $(r,d)$ should be tested for which the data array contains at least $e$ more terms than
+needed. This increases the reliability. By setting the option \verb|cut| to a positive integer~$c$,
+the user requests that for testing a point $(r,d)$, the method should take into account at most $c$
+more terms than needed. If the data array contains more terms, superfluous ones are ignored in the
+interested of better performance. We must always have $0\leq e\leq c\leq \infty$. The default setting
+is $e=0$, $c=\infty$.
+
+Finally, there is an option \verb|infolevel| which, when set to a positive integer, causes the method
+to print some output about the progress of the computation. Higher values lead to more detailed output.
+
+\section{Concluding Remarks}
+
+...
 

### 2013-06-07 paper/guessing.py
guessing improvement: avoid some unnecessary shifts during the reconstruction

@@ -1,0 +1,895 @@
+
+"""
+guessing
+========
+
+"""
+
+######### development mode ###########
+
+try:
+    if sys.modules.has_key('ore_algebra'):
+        del sys.modules['ore_algebra']
+except:
+    pass
+
+#######################################
+
+from sage.rings.integer_ring import ZZ
+from sage.rings.rational_field import QQ
+from sage.rings.finite_rings.all import GF, is_FiniteField
+from sage.matrix.constructor import Matrix, matrix
+from sage.rings.arith import xgcd
+from sage.parallel.decorate import parallel
+from sage.rings.polynomial.polynomial_ring import is_PolynomialRing
+
+from ore_algebra import *
+
+from datetime import datetime
+import nullspace
+import math
+
+def guess_rec(data, n, S, **kwargs):
+    """
+    Shortcut for ``guess`` applied with an Ore algebra of shift operators in `S` over `K[n]`
+    where `K` is the parent of ``data[0]``.
+
+    See the docstring of ``guess`` for further information.    
+    """
+    R = data[0].parent()[n]; x = R.gen()
+    return guess(data, OreAlgebra(R, (S, {n:n+R.one()}, {})), **kwargs)
+
+def guess_deq(data, x, D, **kwargs):
+    """
+    Shortcut for ``guess`` applied with an Ore algebra of differential operators in `D` over `K[x]`
+    where `K` is the parent of ``data[0]``.
+
+    See the docstring of ``guess`` for further information.    
+    """
+    R = data[0].parent()[x]; x = R.gen()
+    return guess(data, OreAlgebra(R, (D, {}, {x:R.one()})), **kwargs)
+
+def guess_qrec(data, qn, Q, q, **kwargs):
+    """
+    Shortcut for ``guess`` applied with an Ore algebra of `q`-recurrence operators in `Q` over `K[qn]`
+    where `K` is the parent of `q`.
+
+    See the docstring of ``guess`` for further information.    
+    """
+    R = q.parent()[qn]; x = R.gen()
+    return guess(data, OreAlgebra(R, (Q, {qn:q*qn}, {qn:R.one()})), **kwargs)
+
+def guess(data, algebra, **kwargs):
+    """
+    Searches for an element of the algebra which annihilates the given data.
+
+    INPUT:
+
+    - ``data`` -- a list of elements of the algebra's base ring's base ring `K` (or at least
[…]

### 2013-06-10 paper/main.tex
add a sequence example

@@
 
-delta and S
+We compute an annihilator for the sequence
+$c(n) = \sum_{k=0}^n 1 / k!$:
+
+\begin{sageexample}
+  sage: R.<n> = ZZ[]; RS.<Sn> = OreAlgebra(R, 'Sn')
+  sage: inverse_factorials = (n + 1) * Sn - 1
+  sage: partial_sums = inverse_factorials.annihilator_of_sum()
+  sage: partial_sums
+\end{sageexample}
+
+The \verb|to_list| method returns the first few values of a sequence,
+given the initial values:
+
+\begin{sageexample}
+  sage: L = partial_sums.to_list([1, 2], 8)
+  sage: L
+  sage: N(L[7])
+\end{sageexample}
+
+In some cases (for example when the base ring is $\mathbb{Z}$ or
+$\mathbb{Z}[x]$), isolated values can be
+computed asymptotically faster for large $n$ using the binary splitting
+technique.
+The \verb|forward_matrix_bsplit| method, called with argument $n$,
+returns a matrix $P$ and a denominator $Q$ such that $P / Q$ multiplied a
+column vector of initial values $c_0, c_1, \ldots$
+yields $c_n, c_{n+1}, \ldots$.
+This way, computing $10^5$ digits of $e$ takes a fraction of a second:
+
+\begin{sageexample}
+  sage: e_approx = N(e, 400000)
+  sage: P, Q = partial_sums.forward_matrix_bsplit(25600)
+  sage: u = Matrix([[e_approx], [e_approx]]) - P * Matrix([[1], [2]]) / Q
+  sage: u.change_ring(RealField(20))
+\end{sageexample}
 

### 2013-06-10 paper/main.tex
paper: references added, section on ore polynomials

--- 2013-06-07 paper/main.tex
@@
 Algorithms for Ore algebras provide a systematic answer to this
-question~\cite{..,..}.  Invented in the early 20th century~\cite{..} with the
-objective of providing a unified theory for various kinds of linear operators,
-they have been used for many years in computer algebra systems, for example in
-the Maple packages gfun~\cite{..} and mgfun~\cite{..}, in the Mathematica
-packages by Mallinger~\cite{..} and Koutschan~\cite{..,..}, and
-elsewhere~\cite{..}.
+question~\cite{bronstein96,chyzak98}.  Invented in the first half of the 20th
+century~\cite{ore33} with the objective of providing a unified theory for
+various kinds of linear operators, they have been used for many years in
+computer algebra systems, for example in the Maple packages
+OreTools~\cite{abramov03}, gfun~\cite{salvy94} or mgfun~\cite{chyzak98a}, in the
+Mathematica packages by Mallinger~\cite{mallinger96} and
+Koutschan~\cite{koutschan10c,koutschan09}.
 
@@
 algorithms related to Ore algebras for the computer algebra system
-Sage~\cite{..}. It is addressed to first-time users who are already familiar
+Sage~\cite{sage}. It is addressed to first-time users who are already familiar
 with Sage, and with the theory of Ore algebras and its use for doing symbolic
 computation related to special functions. Readers unfamiliar with Sage are referred
-to~\cite{..}, and readers unfamiliar with Ore algebras may wish to consult the
-recent tutorial~\cite{..} and the references given there for an introduction to
+to~\cite{sage}, and readers unfamiliar with Ore algebras may wish to consult the
+recent tutorial~\cite{kauers13} and the references given there for an introduction to
 the subject.
@@
 algebras like the ones created above, which are most frequently used in practice. 
-See Section~\ref{sec:4} for an overview. Finally, the package also comes with a 
-fine tuned guessing engine, whose main features we describe in Section~\ref{sec:5}.
+Section~\ref{sec:4} describes some of these features.
 
@@
 whose base ring is a differential field $K=\set Q(x,y,z)$ where $y$ represents
-$\exp(x)$ and $z$ represents $\log(x)$:
+$\exp(x)$ and $z$ represents~$\log(x)$:
 
@@
 
-arithmetic, lclm, gcrd, action, coercion and conversion, pretty printing, 
-coefficient extraction, closure properties, associate solutions.
+Ore polynomials are elements of Ore algebras, i.e., Sage objects whose parent is an Ore algebra
+object as described in the previous section. They can be constructed by addition and multiplication
+from generators and elements of the base ring.
 
-\section{Special Methods for Standard Algebras}\label{sec:4}
+\begin{sageexample}
+  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
+  sage: (5*x^2+3*x-7)*Dx^2 + (3*x^2+8*x-1)*Dx + (9*x^2-3*x+8)
+\end{sageexample}
 
-special methods available for univ algebras with univ algebra as ground ring
+Alternatively, an Ore polynomial can be constructed from any piece of data that is also accepted
+by the constructor of the associated commutative algebra. The associated commutative algebra of
+an Ore algebra is the commutative polynomial ring with the same base ring as the Ore algebra 
+and with generators that are named like the generators of the Ore algebra. In particular, it is
+possible to create an Ore polynomial from the corresponding commutative polynomial, from a 
+coefficient list, or even from a string representation.
 
-\subsection{Differential Operators}
+\begin{sageexample}
+  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
+  sage: Ac = A.associated_commutative_algebra(); Ac
+  sage: A(Ac.random_element())
+  sage: A([5*x,7*x-3,3*x+1])
+  sage: A("(5*x^2+3*x-7)*Dx^2 + (3*x^2+8*x-1)*Dx + (9*x^2-3*x+8)")
+\end{sageexample}
 
-D and theta
+Ore polynomials can also be created from Ore polynomials that belong to other algebras, provided
+that such a conversion is meaningful.
 
-\subsection{Recurrence Operators}
+\begin{sageexample}
+  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
+  sage: L = (5*x^2+3*x-7)*Dx^2 + (3*x^2+8*x-1)*Dx + (9*x^2-3*x+8)
+  sage: L.parent()
+  sage: B = OreAlgebra(QQ['x'], 'Dx')
+  sage: L = B(L)
+  sage: L.parent()
+\end{sageexample}
 
-delta and S
+In accordance with the default Sage coercion model, such conversions take place automatically (if possible) 
+when operators from different algebras are added or multiplied. Note that the result of such an operation
+need not belong to either of the parents of the operands but may instead have a suitable ``common extension'' 
+as parent.
 
-\subsection{The $q$-Case}
+\begin{sageexample}
+  sage: A = OreAlgebra(ZZ['t']['x'], 'Dx'); 
+  sage: B = OreAlgebra(QQ['x'].fraction_field(), 'Dx')
+  sage: L = A.random_element() + B.random_element()
+  sage: L.parent()
+\end{sageexample}
 
-J and Q
+\section{Selected Methods}\label{sec:4}
 
-\section{Guessing}\label{sec:5}
+\subsection{gcrd and lclm?}
 
-[[CITE RUBEY.]]
+-- options for selecting algorithms / remainder sequence
 
-Guessing is one of the most popular features of software packages like gfun.  It
-is a technique for detecting patterns in sequences of numbers, which is often
-used in combinatorics. The basic idea is simple. If we are given a finite array
-of numbers, though of as the first terms of an infinite sequence, we may want to
-know whether this sequence satisfies a recurrence. The algorithm behind a
-guessing engine searches for small equations matching the given data.
+-- cite max's paper
+
+-- available for all (univariate) algebras
+
+\subsection{closure properties?}
+
+-- table mapping function names to arithmetic
+
+-- example for how to prove some identity with the methods provided.
+
+\subsection{solutions?}
+
+-- polynomial, rational, generalized series solutions
+
+-- available for differential and shift algebras
+
+\subsection{action?}
+
+-- should be mentioned somewhere, although there is not much to say about it.
+
+\subsection{to list?}
+
+-- options for handling singularities
+
+-- rectangular splitting
+
+-- available for shift algebras.
+
+\subsection{Guessing}
+
+Guessing is a technique for detecting patterns in sequences of numbers, which is
+often used in combinatorics. It is one of the most popular features of packages
+like gfun, and there are even some special purpose packages dedicated to this
+technique~\cite{kauers09a,hebisch11}.  The basic idea is simple. If we are given
+a finite array of numbers, though of as the first terms of an infinite sequence,
+we may want to know whether this sequence satisfies a recurrence. The algorithm
+behind a guessing engine searches for small equations matching the given data.
 Generically, no such equations exist, so if some are found, it is fair to
@@
 there exists an operator of order~$r$ and degree~$d$ truly satisfied by the
-sequence in question. (See~\cite{..} for an explanation why the boundary of this
-region is usually a hyperbola.) The second describes the region of all points
-$(r,d)$ for which an operator of order~$r$ and degree~$d$ can be detected when
-$N$ terms are provided as input. This region is determined by the requirement
-$(r+1)(d+2)<N$.
+sequence in question. (See~\cite{jaroschek13a} for an explanation why the
+boundary of this region is usually a hyperbola.) The second describes the region
+of all points $(r,d)$ for which an operator of order~$r$ and degree~$d$ can be
+detected when $N$ terms are provided as input. This region is determined by the
+requirement $(r+1)(d+2)<N$.
 

### 2013-06-10 paper/main.tex

@@
 
--- options for handling singularities
+-- available for shift and q-shift algebras.
 
--- rectangular splitting
+We compute an annihilator for the sequence
+$c(n) = \sum_{k=0}^n 1 / k!$:
 
--- available for shift algebras.
+\begin{sageexample}
+  sage: R.<n> = ZZ[]; RS.<Sn> = OreAlgebra(R, 'Sn')
+  sage: inverse_factorials = (n + 1) * Sn - 1
+  sage: partial_sums = inverse_factorials.annihilator_of_sum()
+  sage: partial_sums
+\end{sageexample}
+
+The \verb|to_list| method returns the first few values of a sequence,
+given the initial values:
+
+\begin{sageexample}
+  sage: L = partial_sums.to_list([1, 2], 8)
+  sage: L
+  sage: N(L[7])
+\end{sageexample}
+
+In some cases (for example when the base ring is $\mathbb{Z}$ or
+$\mathbb{Z}[x]$), isolated values can be
+computed asymptotically faster for large $n$ using the binary splitting
+technique.
+The \verb|forward_matrix_bsplit| method, called with argument $n$,
+returns a matrix $P$ and a denominator $Q$ such that $P / Q$ multiplied a
+column vector of initial values $c_0, c_1, \ldots$
+yields $c_n, c_{n+1}, \ldots$.
+This way, computing $10^5$ digits of $e$ takes a fraction of a second:
+
+\begin{sageexample}
+  sage: e_approx = N(e, 400000)
+  sage: P, Q = partial_sums.forward_matrix_bsplit(25600)
+  sage: u = Matrix([[e_approx], [e_approx]]) - P * Matrix([[1], [2]]) / Q
+  sage: u.change_ring(RealField(20))
+\end{sageexample}
 

### 2013-06-10 paper/guessing.py
corrected recent commit about guessing. (mistakenly went into the wrong directory.)

@@ -1,895 +1,0 @@
-
-"""
-guessing
-========
-
-"""
-
-######### development mode ###########
-
-try:
-    if sys.modules.has_key('ore_algebra'):
-        del sys.modules['ore_algebra']
-except:
-    pass
-
-#######################################
-
-from sage.rings.integer_ring import ZZ
-from sage.rings.rational_field import QQ
-from sage.rings.finite_rings.all import GF, is_FiniteField
-from sage.matrix.constructor import Matrix, matrix
-from sage.rings.arith import xgcd
-from sage.parallel.decorate import parallel
-from sage.rings.polynomial.polynomial_ring import is_PolynomialRing
-
-from ore_algebra import *
-
-from datetime import datetime
-import nullspace
-import math
-
-def guess_rec(data, n, S, **kwargs):
-    """
-    Shortcut for ``guess`` applied with an Ore algebra of shift operators in `S` over `K[n]`
-    where `K` is the parent of ``data[0]``.
-
-    See the docstring of ``guess`` for further information.    
-    """
-    R = data[0].parent()[n]; x = R.gen()
-    return guess(data, OreAlgebra(R, (S, {n:n+R.one()}, {})), **kwargs)
-
-def guess_deq(data, x, D, **kwargs):
-    """
-    Shortcut for ``guess`` applied with an Ore algebra of differential operators in `D` over `K[x]`
-    where `K` is the parent of ``data[0]``.
-
-    See the docstring of ``guess`` for further information.    
-    """
-    R = data[0].parent()[x]; x = R.gen()
-    return guess(data, OreAlgebra(R, (D, {}, {x:R.one()})), **kwargs)
-
-def guess_qrec(data, qn, Q, q, **kwargs):
-    """
-    Shortcut for ``guess`` applied with an Ore algebra of `q`-recurrence operators in `Q` over `K[qn]`
-    where `K` is the parent of `q`.
-
-    See the docstring of ``guess`` for further information.    
-    """
-    R = q.parent()[qn]; x = R.gen()
-    return guess(data, OreAlgebra(R, (Q, {qn:q*qn}, {qn:R.one()})), **kwargs)
-
-def guess(data, algebra, **kwargs):
-    """
-    Searches for an element of the algebra which annihilates the given data.
-
-    INPUT:
-
-    - ``data`` -- a list of elements of the algebra's base ring's base ring `K` (or at least
[…]

### 2013-06-10 paper/main.tex
paper: minor modifications.

@@
 Section~\ref{sec:4} describes some of these features.
+
+/// disclaimer: we don't explain algorithms here, but only the software ///
 
@@
 
+-- conceptually the base ring is always a field, even when it's not.
+
 \subsection{closure properties?}
@@
 -- example for how to prove some identity with the methods provided.
+
+-- merge with previous section?
+
+\subsection{conversions?}
+
+-- re2de and all its relatives
 
@@
 
-\subsection{to list?}
+\subsection{Computation of Sequence Terms}
 
--- available for shift and q-shift algebras.
+(... a few introductory words ... available for shift and q-shift algebras ...)
 
@@
 \begin{sageexample}
-  sage: R.<n> = ZZ[]; RS.<Sn> = OreAlgebra(R, 'Sn')
+  sage: R.<n> = ZZ[]; RS.<Sn> = OreAlgebra(R)
   sage: inverse_factorials = (n + 1) * Sn - 1
@@
 technique.
-The \verb|forward_matrix_bsplit| method, called with argument $n$,
-returns a matrix $P$ and a denominator $Q$ such that $P / Q$ multiplied a
+The \verb|forward_matrix_bsplit| method, called with argument~$n$,
+returns a matrix~$P$ and a polynomial~$Q$ such that $P / Q$ multiplied by a
 column vector of initial values $c_0, c_1, \ldots$
 yields $c_n, c_{n+1}, \ldots$.
-This way, computing $10^5$ digits of $e$ takes a fraction of a second:
+This way, computing $10^5$ digits of $\mathrm{e}$ takes a fraction of a second:
 
@@
 
+... add a figure with timings? ...
+
 \subsection{Guessing}
 
-Guessing is a technique for detecting patterns in sequences of numbers, which is
-often used in combinatorics. It is one of the most popular features of packages
-like gfun, and there are even some special purpose packages dedicated to this
-technique~\cite{kauers09a,hebisch11}.  The basic idea is simple. If we are given
-a finite array of numbers, though of as the first terms of an infinite sequence,
-we may want to know whether this sequence satisfies a recurrence. The algorithm
-behind a guessing engine searches for small equations matching the given data.
-Generically, no such equations exist, so if some are found, it is fair to
-``guess'' that they are in fact valid equations for the whole infinite sequence.
+Guessing is, in some sense, the reverse operation of \verb|to_list|. It is one
+of the most popular features of packages like gfun, and there are even some
+special purpose packages dedicated to this technique~\cite{kauers09a,hebisch11}.
+The basic idea is simple. If we are given a finite array of numbers, thought of
+as the first terms of an infinite sequence, we may want to know whether this
+sequence satisfies a recurrence. The algorithm behind a guessing engine searches
+for small equations matching the given data.  Generically, no such equations
+exist, so if some are found, it is fair to ``guess'' that they are in fact valid
+equations for the whole infinite sequence.
 

### 2013-06-10 paper/main.tex
expand a bit (note: printing the asymptotic expansion causes latex errors)

--- 2013-06-10 paper/main.tex
@@
 Algorithms for Ore algebras provide a systematic answer to this
-question~\cite{bronstein96,chyzak98}.  Invented in the first half of the 20th
-century~\cite{ore33} with the objective of providing a unified theory for
-various kinds of linear operators, they have been used for many years in
-computer algebra systems, for example in the Maple packages
-OreTools~\cite{abramov03}, gfun~\cite{salvy94} or mgfun~\cite{chyzak98a}, in the
-Mathematica packages by Mallinger~\cite{mallinger96} and
-Koutschan~\cite{koutschan10c,koutschan09}.
+question~\cite{..,..}.  Invented in the early 20th century~\cite{..} with the
+objective of providing a unified theory for various kinds of linear operators,
+they have been used for many years in computer algebra systems, for example in
+the Maple packages gfun~\cite{..} and mgfun~\cite{..}, in the Mathematica
+packages by Mallinger~\cite{..} and Koutschan~\cite{..,..}, and
+elsewhere~\cite{..}.
 
@@
 algorithms related to Ore algebras for the computer algebra system
-Sage~\cite{sage}. It is addressed to first-time users who are already familiar
+Sage~\cite{..}. It is addressed to first-time users who are already familiar
 with Sage, and with the theory of Ore algebras and its use for doing symbolic
 computation related to special functions. Readers unfamiliar with Sage are referred
-to~\cite{sage}, and readers unfamiliar with Ore algebras may wish to consult the
-recent tutorial~\cite{kauers13} and the references given there for an introduction to
+to~\cite{..}, and readers unfamiliar with Ore algebras may wish to consult the
+recent tutorial~\cite{..} and the references given there for an introduction to
 the subject.
@@
 algebras like the ones created above, which are most frequently used in practice. 
-Section~\ref{sec:4} describes some of these features.
+See Section~\ref{sec:4} for an overview. Finally, the package also comes with a 
+fine tuned guessing engine, whose main features we describe in Section~\ref{sec:5}.
 
@@
 whose base ring is a differential field $K=\set Q(x,y,z)$ where $y$ represents
-$\exp(x)$ and $z$ represents~$\log(x)$:
+$\exp(x)$ and $z$ represents $\log(x)$:
 
@@
 
-Ore polynomials are elements of Ore algebras, i.e., Sage objects whose parent is an Ore algebra
-object as described in the previous section. They can be constructed by addition and multiplication
-from generators and elements of the base ring.
+arithmetic, lclm, gcrd, action, coercion and conversion, pretty printing, 
+coefficient extraction, closure properties, associate solutions.
 
-\begin{sageexample}
-  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
-  sage: (5*x^2+3*x-7)*Dx^2 + (3*x^2+8*x-1)*Dx + (9*x^2-3*x+8)
-\end{sageexample}
+\section{Special Methods for Standard Algebras}\label{sec:4}
 
-Alternatively, an Ore polynomial can be constructed from any piece of data that is also accepted
-by the constructor of the associated commutative algebra. The associated commutative algebra of
-an Ore algebra is the commutative polynomial ring with the same base ring as the Ore algebra 
-and with generators that are named like the generators of the Ore algebra. In particular, it is
-possible to create an Ore polynomial from the corresponding commutative polynomial, from a 
-coefficient list, or even from a string representation.
+special methods available for univ algebras with univ algebra as ground ring
 
-\begin{sageexample}
-  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
-  sage: Ac = A.associated_commutative_algebra(); Ac
-  sage: A(Ac.random_element())
-  sage: A([5*x,7*x-3,3*x+1])
-  sage: A("(5*x^2+3*x-7)*Dx^2 + (3*x^2+8*x-1)*Dx + (9*x^2-3*x+8)")
-\end{sageexample}
+\subsection{Differential Operators}
 
-Ore polynomials can also be created from Ore polynomials that belong to other algebras, provided
-that such a conversion is meaningful.
+D and theta
 
-\begin{sageexample}
-  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
-  sage: L = (5*x^2+3*x-7)*Dx^2 + (3*x^2+8*x-1)*Dx + (9*x^2-3*x+8)
-  sage: L.parent()
-  sage: B = OreAlgebra(QQ['x'], 'Dx')
-  sage: L = B(L)
-  sage: L.parent()
-\end{sageexample}
+\subsection{Recurrence Operators}
 
-In accordance with the default Sage coercion model, such conversions take place automatically (if possible) 
-when operators from different algebras are added or multiplied. Note that the result of such an operation
-need not belong to either of the parents of the operands but may instead have a suitable ``common extension'' 
-as parent.
+\begin{center}
+  \begin{tabular}{|l|p{.55\hsize}|}
+    \hline
+    Method name & short description \\\hline
+    \verb|to_D()| & converts annihilator for the coefficients in a power series to a
+        differential operator for the sum \\
+    \verb|to_F()| & converts shift operator to a difference operator\\
+    \verb|to_T()| & converts to a differential operator in terms of the Euler derivative \\
+    \verb|annihilator_of_sum()| & converts an annihilator for $f(n)$ to one for $\sum_{k=0}^n f(k)$\\
+    \verb|annihilator_of_composition()| & converts an annihilator for $f(n)$ to one for $f(\lfloor un+v \rfloor)$ where $u, v \in \mathbb{Q}$ \\
+    \verb|annihilator_of_interlacing()| & interlaces two or more sequences\\
+    \verb|generalized_series_solutions()| & computes asymptotic expansions of sequences annihilated by the operator\\ \hline
+  \end{tabular}
+\end{center}
 
-\begin{sageexample}
-  sage: A = OreAlgebra(ZZ['t']['x'], 'Dx'); 
-  sage: B = OreAlgebra(QQ['x'].fraction_field(), 'Dx')
-  sage: L = A.random_element() + B.random_element()
-  sage: L.parent()
-\end{sageexample}
-
-\section{Selected Methods}\label{sec:4}
-
-\subsection{gcrd and lclm?}
-
--- options for selecting algorithms / remainder sequence
-
--- cite max's paper
-
--- available for all (univariate) algebras
-
-\subsection{closure properties?}
-
--- table mapping function names to arithmetic
-
--- example for how to prove some identity with the methods provided.
-
-\subsection{solutions?}
-
--- polynomial, rational, generalized series solutions
-
--- available for differential and shift algebras
-
-\subsection{action?}
-
--- should be mentioned somewhere, although there is not much to say about it.
-
-\subsection{to list?}
-
--- available for shift and q-shift algebras.
+\smallskip
 
@@
 
+We compute the asymptotic expansion of the sequence of terms
+to estimate how many terms we need to approximate $e$ to a given
+number of digits:
+
+\begin{sageexample}
+  sage: digits = 10^5
+  sage: asymp = inverse_factorials.generalized_series_solutions(3)
+  sage: target = lambda x: log(abs(asymp[0](RR(x))), 10) + digits
+  sage: num_terms = ceil(find_root(target, 1, 10^6))
+  sage: num_terms
+\end{sageexample}
+
 In some cases (for example when the base ring is $\mathbb{Z}$ or
-$\mathbb{Z}[x]$), isolated values can be
-computed asymptotically faster for large $n$ using the binary splitting
-technique.
+$\mathbb{Z}[x]$), isolated values of a sequence can be
+computed asymptotically faster for large $n$ than by listing all
+values, using the binary splitting technique.
 The \verb|forward_matrix_bsplit| method, called with argument $n$,
@@
   sage: e_approx = N(e, 400000)
-  sage: P, Q = partial_sums.forward_matrix_bsplit(25600)
+  sage: P, Q = partial_sums.forward_matrix_bsplit(num_terms)
   sage: u = Matrix([[e_approx], [e_approx]]) - P * Matrix([[1], [2]]) / Q
@@
 
-\subsection{Guessing}
+\subsection{The $q$-Case}
 
-Guessing is a technique for detecting patterns in sequences of numbers, which is
-often used in combinatorics. It is one of the most popular features of packages
-like gfun, and there are even some special purpose packages dedicated to this
-technique~\cite{kauers09a,hebisch11}.  The basic idea is simple. If we are given
-a finite array of numbers, though of as the first terms of an infinite sequence,
-we may want to know whether this sequence satisfies a recurrence. The algorithm
-behind a guessing engine searches for small equations matching the given data.
+J and Q
+
+\section{Guessing}\label{sec:5}
+
+[[CITE RUBEY.]]
+
+Guessing is one of the most popular features of software packages like gfun.  It
+is a technique for detecting patterns in sequences of numbers, which is often
+used in combinatorics. The basic idea is simple. If we are given a finite array
+of numbers, though of as the first terms of an infinite sequence, we may want to
+know whether this sequence satisfies a recurrence. The algorithm behind a
+guessing engine searches for small equations matching the given data.
 Generically, no such equations exist, so if some are found, it is fair to
@@
 there exists an operator of order~$r$ and degree~$d$ truly satisfied by the
-sequence in question. (See~\cite{jaroschek13a} for an explanation why the
-boundary of this region is usually a hyperbola.) The second describes the region
-of all points $(r,d)$ for which an operator of order~$r$ and degree~$d$ can be
-detected when $N$ terms are provided as input. This region is determined by the
-requirement $(r+1)(d+2)<N$.
+sequence in question. (See~\cite{..} for an explanation why the boundary of this
+region is usually a hyperbola.) The second describes the region of all points
+$(r,d)$ for which an operator of order~$r$ and degree~$d$ can be detected when
+$N$ terms are provided as input. This region is determined by the requirement
+$(r+1)(d+2)<N$.
 

### 2013-06-10 paper/main.tex

--- 2013-06-10 paper/main.tex
@@
 
+\begin{center}
+  \begin{tabular}{|l|p{.55\hsize}|}
+    \hline
+    Method name & short description \\\hline
+    \verb|to_D()| & converts annihilator for the coefficients in a power series to a
+        differential operator for the sum \\
+    \verb|to_F()| & converts shift operator to a difference operator\\
+    \verb|to_T()| & converts to a differential operator in terms of the Euler derivative \\
+    \verb|annihilator_of_sum()| & converts an annihilator for $f(n)$ to one for $\sum_{k=0}^n f(k)$\\
+    \verb|annihilator_of_composition()| & converts an annihilator for $f(n)$ to one for $f(\lfloor un+v \rfloor)$ where $u, v \in \mathbb{Q}$ \\
+    \verb|annihilator_of_interlacing()| & interlaces two or more sequences\\
+    \verb|generalized_series_solutions()| & computes asymptotic expansions of sequences annihilated by the operator\\ \hline
+  \end{tabular}
+\end{center}
+
+\smallskip
+
 We compute an annihilator for the sequence
@@
 
+We compute the asymptotic expansion of the sequence of terms
+to estimate how many terms we need to approximate $e$ to a given
+number of digits:
+
+\begin{sageexample}
+  sage: digits = 10^5
+  sage: asymp = inverse_factorials.generalized_series_solutions(3)
+  sage: target = lambda x: log(abs(asymp[0](RR(x))), 10) + digits
+  sage: num_terms = ceil(find_root(target, 1, 10^6))
+  sage: num_terms
+\end{sageexample}
+
 In some cases (for example when the base ring is $\mathbb{Z}$ or
-$\mathbb{Z}[x]$), isolated values can be
-computed asymptotically faster for large $n$ using the binary splitting
-technique.
+$\mathbb{Z}[x]$), isolated values of a sequence can be
+computed asymptotically faster for large $n$ than by listing all
+values, using the binary splitting technique.
 The \verb|forward_matrix_bsplit| method, called with argument~$n$,
-returns a matrix~$P$ and a polynomial~$Q$ such that $P / Q$ multiplied by a
+returns a matrix $P$ and a denominator $Q$ such that $P / Q$ multiplied a
 column vector of initial values $c_0, c_1, \ldots$
@@
 
+\subsection{The $q$-Case}
+
+J and Q
+
+\section{Guessing}\label{sec:5}
+
+[[CITE RUBEY.]]
+
+Guessing is one of the most popular features of software packages like gfun.  It
+is a technique for detecting patterns in sequences of numbers, which is often
+used in combinatorics. The basic idea is simple. If we are given a finite array
+of numbers, though of as the first terms of an infinite sequence, we may want to
+know whether this sequence satisfies a recurrence. The algorithm behind a
+guessing engine searches for small equations matching the given data.
+Generically, no such equations exist, so if some are found, it is fair to
+``guess'' that they are in fact valid equations for the whole infinite sequence.
+=======
 \subsection{Guessing}
@@
 equations for the whole infinite sequence.
+>>>>>>> manuel/master
 

### 2013-06-13 ore_operator.py
minor bugfix in improvedPRS

@@ -892,7 +892,7 @@ class UnivariateOreOperator(OreOperator):
         a11, a12, a21, a22 = RF.one(), RF.zero(), RF.zero(), RF.one()
 
         if prs is None:
-            prs = __classicPRS__ if R.base_ring().is_field() else __improvedPRS__
+            prs = __classicPRS__ if R.base_ring().is_field() else (__improvedPRS__ if retval=="bezout" else __primitivePRS__)
 
         additional = []
 
@@ -903,7 +903,6 @@ class UnivariateOreOperator(OreOperator):
             else:
                 r = r2; bInv = ~beta
                 a11, a12, a21, a22 = a21, a22, bInv*(alpha*a11 - q*a21), bInv*(alpha*a12 - q*a22)
-
         if retval == "syzygy":
             c = a21.denominator().lcm(a22.denominator())
             return (c*a21, c*a22) ### ???correct???
@@ -1544,7 +1543,8 @@ def __monicPRS__(r,additional):
     """
 
     newRem = r[0].quo_rem(r[1])
-    return ((r[1],newRem[1].primitive_part()),newRem[0],r[0].parent().base_ring().one(),r[0].parent().base_ring().one(),True)
+    beta = newRem[1].leading_coefficient() if not newRem[1].is_zero() else r[0].parent().base_ring().one()
+    return ((r[1],newRem[1].primitive_part()),newRem[0],r[0].parent().base_ring().one(),beta,True)
 
 def __improvedPRS__(r,additional):
     """
@@ -1562,18 +1562,18 @@ def __improvedPRS__(r,additional):
     if (len(additional)==0):
         essentialPart = gcd(sigma(r[0].leading_coefficient(),-orddiff),r[1].leading_coefficient())
         phi = Rbase.one()
-        beta = (-Rbase.one())**(orddiff+1)*sigma().factorial(sigma(phi,1),orddiff)
+        beta = (-Rbase.one())**(orddiff+1)*sigma.factorial(sigma(phi,1),orddiff)
     else:
         (d2,oldalpha,k,essentialPart,phi) = (additional.pop(),additional.pop(),additional.pop(),additional.pop(),additional.pop())
-        phi = oldalpha / sigma().factorial(sigma(phi,1),d2-d0-1)
[…]

### 2013-06-13 paper/main.tex
add a table of methods and an example for differential operators

@@
 
+\subsection{Differential Operators}
+
+Useful methods available for differential operators:
+
+\begin{center}
+  \begin{tabular}{|l|p{.55\hsize}|}
+    \hline
+    Method name & short description \\\hline
+    \verb|to_S()| & converts to a recurrence operator for the Taylor series solutions at the origin \\
+    \verb|to_F()| & converts to a difference operator for the Taylor series solutions at the origin \\
+    \verb|to_T()| & rewrites in terms of the Euler derivative \\
+    \verb|annihilator_of_integral()| & converts an annihilator for $f(x)$ to one for $\int f(x) dx$\\
+    \verb|annihilator_of_composition()| & converts an annihilator for $f(x)$ to one for $f(a(x))$ where $a(x)$ is algebraic over the base ring\\
+    \verb|power_series_solutions()| & computes power series solutions of this operator\\
+    \verb|generalized_series_solutions()| & computes generalized series solutions of this operator \\ \hline
+  \end{tabular}
+\end{center}
+
+\smallskip
+
+We compute an annihilator for the error function $\frac{2}{\sqrt{\pi}} \int_0^x e^{-t^2} dt$,
+starting from the differential equation for $e^x$, and produce the
+recurrence for the Taylor series coefficients at the origin:
+
+\begin{sageexample}
+  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R, 'Dx')                            
+  sage: (Dx - 1).annihilator_of_composition(-x**2)
+  Dx + 2*x
+  sage: (Dx + 2*x).annihilator_of_integral()                                     
+  Dx^2 + 2*x*Dx
+  sage: (Dx^2 + 2*x*Dx).to_S(OreAlgebra(ZZ['n'], 'Sn'))                          
+  (n^2 + 3*n + 2)*Sn^2 + 2*n
+  sage: (Dx^2 + 2*x*Dx).power_series_solutions(10)                               
+  [x - 1/3*x^3 + 1/10*x^5 - 1/42*x^7 + 1/216*x^9 + O(x^10), 1 + O(x^10)]
+\end{sageexample}
+
 \subsection{Computation of Sequence Terms}
@@
 (... a few introductory words ... available for shift and q-shift algebras ...)
+
+Useful methods available for recurrence operators:
 
@@
 yields $c_n, c_{n+1}, \ldots$.
-This way, computing $10^5$ digits of $\mathrm{e}$ takes a fraction of a second:
+This way, computing $10^5$ digits of $e$ takes a fraction of a second:
 

### 2013-06-13 paper/main.tex
text for gcrd and lclm. some changes to normalize() so that leading sign is + added message when improvedPRS fails

--- 2013-06-10 paper/main.tex
@@
 \begin{sageexample}
-  sage: load("ore_algebra.py", "guessing.py");  #### TO BE FIXED ####
+  sage: load("ore_algebra.py", "guessing.py", "ore_operator.py");  #### TO BE FIXED ####
   sage: R.<x> = PolynomialRing(ZZ); A.<Dx> = OreAlgebra(R)
@@
 
-\subsection{gcrd and lclm?}
+\subsection{gcrd and lclm}
 
--- options for selecting algorithms / remainder sequence
+A univariate Ore algebra over a field is a left Euclidean domain, which means that it is possible to perform 
+left division with remainder. Building upon this, the greatest common right divisor (GCRD) and the least common 
+left multiple (LCLM) of two Ore polynomials can be computed. The package provides a number of methods to carry 
+out these tasks
 
--- cite max's paper
+\begin{center}
+  \begin{tabular}{|l|p{.55\hsize}|}
+    \hline
+    Method name & short description \\\hline
+    \verb|quo_rem(B)| & returns the left quotient and the left remainder of $A$ and $B$.\\
+    \verb|gcrd(B)| & returns the greatest common right divisor of $A$ and $B$.\\
+    \verb|xgcrd(B)| & returns the greatest common right divisor of $A$ and $B$ and the according B\'ezout coefficients.\\
+    \verb|lclm(B)| & returns the least common left multiple of $A$ and $B$.\\\hline
+  \end{tabular}
+\end{center}
 
--- available for all (univariate) algebras
+All these methods are also available for Ore operators living in univariate Ore algebras over a base ring $R$ which 
+does not necessarily have to be a field. The operators are then implicitly assumed to live in the respective Ore 
+algebra over the quotient field $K$ of $R$. The output will be the GCRD (LCLM, quotient, remainder) in the Ore 
+algebra over $K$ but not over $R$, in which these objects might not exist or might not be computable.
 
--- conceptually the base ring is always a field, even when it's not.
+\begin{sageexample}
+  sage: A = OreAlgebra(ZZ['n'], 'Sn');                                           
+  sage: G,L1,L2 = A.random_element(2),A.random_element(7),A.random_element(5)    
+  sage: while not L1.gcrd(L2)==1: L2 = A.random_element(5)                       
+  sage: L1,L2 = L1*G,L2*G                                                        
+  sage: (L3,S,T) = L1.xgcrd(L2)                             
+  sage: (S*L1+T*L2).normalize() == L3.normalize() == G.normalize()
+  sage: LCLM = L1.lclm(L2)
+  sage: LCLM.quo_rem(L1)[1] == LCLM.quo_rem(L2)[1] == A.zero()
+\end{sageexample}
+
+There are a number of options to adjust the functionality of the methods. For example, the efficiency of computing the GCRD 
+depends on the size of the coefficients of intermediate results and there are different strategies to control this in so called 
+polynomial remainder sequences (PRS). The default is the improved PRS described in \cite{jaroschek13b}. which will usually be 
+the fastest choice.
+
+\begin{sageexample}
+  sage: A = OreAlgebra(ZZ['n'], 'Sn');                                           
+  sage: L1,L2 = A.random_element(3),A.random_element(2)    
+  sage: L1.gcrd(L2, prs=__primitivePRS__)
+  sage: L1.gcrd(L2, prs=__classicPRS__)
+\end{sageexample}
 

### 2013-06-13 ore_operator.py

@@ -10,6 +10,7 @@ from sage.rings.arith import gcd, lcm
 from sage.rings.rational_field import QQ
 from sage.rings.integer_ring import ZZ
 from sage.rings.infinity import infinity
+from sage.functions.generalized import sign
 
 class OreOperator(RingElement):
     """
@@ -522,10 +523,17 @@ class OreOperator(RingElement):
                     c = c.lc()
                 except:
                     break
-        if c.is_unit():
-            return self.parent()((~c)*num)
-        else:
-            return num
+        while c.parent() is not c.parent().base_ring():
+            try:
+                c = c.parent().base_ring()(c)
+            except:
+                pass
+        if not c.is_unit(): 
+            try:
+                c = sign(c)
+            except:
+                c=c.parent().one()
+        return self.parent()((~c)*num)
 
     def map_coefficients(self, f, new_base_ring = None):
         """
@@ -856,6 +864,7 @@ class UnivariateOreOperator(OreOperator):
         while not r[1].is_zero():
             (r2,q,alpha,beta,correct)=prs(r,additional)
             if not correct:
+                print "switch to primitve PRS"
                 prs = __primitivePRS__
             else:
                 r=r2
@@ -866,7 +875,7 @@ class UnivariateOreOperator(OreOperator):
 
         return r
 
-    def xgcd(self, other, prs=None):
+    def xgcrd(self, other, prs=None):
         """
         When called for two operators p,q, this will return their GCRD g together with 
         two operators s and t such that sp+tq=g. 
@@ -905,6 +914,7 @@ class UnivariateOreOperator(OreOperator):
         while not r[1].is_zero():  
             (r2, q, alpha, beta, correct) = prs(r, additional)
             if not correct:
+                print "switch to primitve PRS"
                 prs = __primitivePRS__
             else:
                 r = r2; bInv = ~beta
@@ -915,7 +925,7 @@ class UnivariateOreOperator(OreOperator):
 
         r = r[0]
         c = RF.base_ring().one() if prs is __classicPRS__ else ~r.content()
-        return (self.parent()(c*r), c*a11, c*a12) 
+        return (self.parent()(c*r), c*a11, c*a12) if self.order()>=other.order() else (self.parent()(c*r), c*a12, c*a11)
 
     def lclm(self, *other, **kwargs):
[…]

### 2013-06-13 ore_operator.py
bug in equality test fixed.

@@ -686,8 +686,10 @@ class UnivariateOreOperator(OreOperator):
 
     def __eq__(self, other):
 
+        if self.order() < 0:
+            return other == ZZ.zero()
         if self.order() == 0:
-            return self.constant_coefficient() == other
+            return other == self.constant_coefficient()
         elif not isinstance(other, OreOperator):
             return False
         elif self.parent() == other.parent():

### 2013-06-13 ore_operator.py
another fix of the same problem. hopefully settled now.

@@ -681,25 +681,8 @@ class UnivariateOreOperator(OreOperator):
     def __nonzero__(self):
         return self._poly.__nonzero__()
 
-    def __neq__(self, other):
-        return not (self == other)
-
-    def __eq__(self, other):
-
-        if self.order() < 0:
-            return other == ZZ.zero()
-        if self.order() == 0:
-            return other == self.constant_coefficient()
-        elif not isinstance(other, OreOperator):
-            return False
-        elif self.parent() == other.parent():
-            return self.polynomial() == other.polynomial()
-        else:
-            try:
-                A, B = canonical_coercion(self, other)
-                return A == B
-            except:
-                return False
+    def __cmp__(self, other):
+        return cmp(self.polynomial(), other.polynomial())
 
     def _is_atomic(self):
         return self._poly._is_atomic()

### 2013-06-14 doc/conf.py
added files for creating an spkg file.
To produce an spkg file from the current *.py files, go to the
spkg directory and type "make". This produces a file ore_algebra-0.1.spkg.

To install the package into your Sage distribution, type

  sage -i ore_algebra-0.1.spkg

After that, from within Sage, the package can be loaded via

  sage: from ore_algebra import *

--

removed redundant *.py files in doc directory.

To build documentation, go to doc and type:

  sage -sh
  cp ../*.py .
  make html
  exit

(So the py files are still needed, but there is no point to have them
under version control.)

@@ -1,196 +1,0 @@
-# -*- coding: utf-8 -*-
-#
-# holonomic documentation build configuration file, created by
-# sphinx-quickstart on Sat Feb  2 16:17:05 2013.
-#
-# This file is execfile()d with the current directory set to its containing dir.
-#
-# Note that not all possible configuration values are present in this
-# autogenerated file.
-#
-# All configuration values have a default; values that are commented out
-# serve to show the default.
-
-import sys, os
-
-# If extensions (or modules to document with autodoc) are in another directory,
-# add these directories to sys.path here. If the directory is relative to the
-# documentation root, use os.path.abspath to make it absolute, like shown here.
-sys.path.append(os.path.abspath('.'))
-
-# -- General configuration -----------------------------------------------------
-
-# Add any Sphinx extension module names here, as strings. They can be extensions
-# coming with Sphinx (named 'sphinx.ext.*') or your custom ones.
-extensions = ['sphinx.ext.autodoc', 'sphinx.ext.doctest', 'sphinx.ext.todo', 'sphinx.ext.mathjax']
-
-# Add any paths that contain templates here, relative to this directory.
-templates_path = ['_templates']
-
-# The suffix of source filenames.
-source_suffix = '.rst'
-
-# The encoding of source files.
-#source_encoding = 'utf-8'
-
-# The master toctree document.
-master_doc = 'index'
-
-# General information about the project.
-project = u'ore_algebra'
-copyright = u'2013, Manuel Kauers, Maximilian Jaroschek, Fredrik Johansson'
-
-# The version info for the project you're documenting, acts as replacement for
-# |version| and |release|, also used in various other places throughout the
-# built documents.
-#
-# The short X.Y version.
-version = '0.1'
-# The full version, including alpha/beta/rc tags.
-release = '0.1'
-
-# The language for content autogenerated by Sphinx. Refer to documentation
-# for a list of supported languages.
-#language = None
-
-# There are two options for replacing |today|: either, you set today to some
-# non-false value, then it is used:
-#today = ''
-# Else, today_fmt is used as the format for a strftime call.
-#today_fmt = '%B %d, %Y'
-
-# List of documents that shouldn't be included in the build.
-#unused_docs = []
-
-# List of directories, relative to source directory, that shouldn't be searched
-# for source files.
-exclude_trees = ['_build']
-
-# The reST default role (used for this markup: `text`) to use for all documents.
-#default_role = None
-
[…]

### 2013-06-14 doc/dfinite_function.py

@@ -1,431 +1,0 @@
-
-"""
-dfinite_function
-================
-
-"""
-
-from sage.rings.ring import Algebra
-
-class DFiniteFunctionRing(Algebra):
-    """
-    A Ring of Dfinite objects (functions or sequences)
-    """
-    _no_generic_basering_coercion = False
-    
-    def __init__(self, base_ring, name=None, element_class=None, category=None):
-        """
-        """
-        raise NotImplementedError
-        
-    def _element_constructor_(self, x=None, check=True, is_gen = False, construct=False, **kwds):
-        r"""
-        Convert ``x`` into this algebra, possibly non-canonically.
-        """
-        raise NotImplementedError
-
-    def is_integral_domain(self, proof = True):
-        """
-        """
-        return False
-
-    def is_noetherian(self):
-        """
-        """
-        return False
-            
-    def construction(self):
-        """
-        """
-        raise NotImplementedError
-
-    def _coerce_map_from_(self, P):
-        """
-        """
-        raise NotImplementedError
-
-    def _sage_input_(self, sib, coerced):
-        r"""
-        Produce an expression which will reproduce this value when
-        evaluated.
-        """
-        raise NotImplementedError
-
-    def _is_valid_homomorphism_(self, codomain, im_gens):
-        """
-        """
-        raise NotImplementedError
-
-    def __hash__(self):
-        # should be faster than just relying on the string representation
-        try:
-            return self._cached_hash
-        except AttributeError:
-            pass
-        h = self._cached_hash = hash((self.base_ring(),self.variable_name()))
-        return h
-
-    def _repr_(self):
-        try:
-            return self._cached_repr
-        except AttributeError:
-            pass
-        raise NotImplementedError
-
-    def _latex_(self):
-        """
-        """
-        raise NotImplementedError
-
-    def base_extend(self, R):
-        """
-        """
-        raise NotImplementedError
-
-    def ore_algebra(self):
-        """
-        """
-        raise NotImplementedError
-                
-    def characteristic(self):
-        """
-        Return the characteristic of this Ore algebra, which is the
-        same as that of its base ring.
-        """
-        return self.base_ring().characteristic()
-
-    def is_finite(self):
-        """
-        Return False since polynomial rings are not finite (unless the base
-        ring is 0.)
-        """
-        R = self.base_ring()
-        if R.is_finite() and R.order() == 1:
-            return True
-        return False
-
-    def is_exact(self):
-        """
-        """
[…]

### 2013-06-14 doc/generalized_series.py

@@ -1,1490 +1,0 @@
-
-"""
-generalized_series
-==================
-
-"""
-
-from sage.rings.power_series_ring import PowerSeriesRing
-from sage.rings.polynomial.polynomial_ring import is_PolynomialRing
-from sage.rings.integer_ring import ZZ
-from sage.rings.rational_field import QQ
-from sage.rings.real_mpfr import RR
-from sage.functions.log import log as LOG
-from sage.functions.log import exp as EXP
-from sage.rings.number_field.number_field import is_NumberField
-from sage.structure.element import Element, RingElement, canonical_coercion
-from sage.structure.parent import Parent
-from sage.rings.infinity import infinity
-from sage.rings.arith import gcd, lcm
-from sage.rings.qqbar import QQbar
-
-import re
-
-def GeneralizedSeriesMonoid(base, x, type="continuous"):
-    r"""
-    Creates a monoid of generalized series objects.
-
-    INPUT:
-
-    - ``base`` -- constant field, may be either ``QQ`` or a number field. 
-    - ``x`` -- name of the variable, must not contain the substring ``"log"``.
-    - ``type`` (optional) -- either ``"continuous"`` or ``"discrete"``.
-
-    If the type is ``"continuous"``, the domain contains series objects of the form
-
-    `\exp(\int_0^x \frac{p(t^{-1/r})}t dt)*q(x^{1/r},\log(x))`
-
-    where
-
-    * `r` is a positive integer (the object's "ramification")
-    * `p` is in `K[x]` (the object's "exponential part")
-    * `q` is in `K[[x]][y]` with `x\nmid q` unless `q` is zero (the object's "tail")
-    * `K` is the base ring.
-
-    Any two such objects can be multiplied and differentiated.
-    Objects whose exponential parts differ by an integer ("similar" series) can also be added. 
-
-    If the type is ``"discrete"``, the domain contains series objects of the form
-
-    `(x/e)^{x u/v}\rho^x\exp\bigl(c_1 x^{1/(m*v)} +...+ c_{v-1} x^{1-1/(m*v)}\bigr)x^\alpha p(x^{-1/(m*v)},\log(x))`
-
-    where
-
-    * `e` is Euler's constant (2.71...)
-    * `v` is a positive integer (the object's "ramification")
-    * `u` is an integer; the term `(x/e)^(v/u)` is called the "superexponential part" of the solution
-    * `\rho` is an element of an algebraic extension of the coefficient field `K`
[…]

### 2013-06-14 doc/guessing.py

@@ -1,892 +1,0 @@
-
-"""
-guessing
-========
-
-"""
-
-######### development mode ###########
-
-try:
-    if sys.modules.has_key('ore_algebra'):
-        del sys.modules['ore_algebra']
-except:
-    pass
-
-#######################################
-
-from sage.rings.integer_ring import ZZ
-from sage.rings.rational_field import QQ
-from sage.rings.finite_rings.all import GF, is_FiniteField
-from sage.matrix.constructor import Matrix, matrix
-from sage.rings.arith import xgcd
-from sage.parallel.decorate import parallel
-from sage.rings.polynomial.polynomial_ring import is_PolynomialRing
-
-from ore_algebra import *
-
-from datetime import datetime
-import nullspace
-import math
-
-def guess_rec(data, n, S, **kwargs):
-    """
-    Shortcut for ``guess`` applied with an Ore algebra of shift operators in `S` over `K[n]`
-    where `K` is the parent of ``data[0]``.
-
-    See the docstring of ``guess`` for further information.    
-    """
-    R = data[0].parent()[n]; x = R.gen()
-    return guess(data, OreAlgebra(R, (S, {n:n+R.one()}, {})), **kwargs)
-
-def guess_deq(data, x, D, **kwargs):
-    """
-    Shortcut for ``guess`` applied with an Ore algebra of differential operators in `D` over `K[x]`
-    where `K` is the parent of ``data[0]``.
-
-    See the docstring of ``guess`` for further information.    
-    """
-    R = data[0].parent()[x]; x = R.gen()
-    return guess(data, OreAlgebra(R, (D, {}, {x:R.one()})), **kwargs)
-
-def guess_qrec(data, qn, Q, q, **kwargs):
-    """
-    Shortcut for ``guess`` applied with an Ore algebra of `q`-recurrence operators in `Q` over `K[qn]`
-    where `K` is the parent of `q`.
-
-    See the docstring of ``guess`` for further information.    
-    """
-    R = q.parent()[qn]; x = R.gen()
-    return guess(data, OreAlgebra(R, (Q, {qn:q*qn}, {qn:R.one()})), **kwargs)
-
-def guess(data, algebra, **kwargs):
-    """
-    Searches for an element of the algebra which annihilates the given data.
-
-    INPUT:
-
-    - ``data`` -- a list of elements of the algebra's base ring's base ring `K` (or at least
[…]

### 2013-06-14 doc/nullspace.py

@@ -1,1944 +1,0 @@
-
-"""
-
-nullspace
-=========
-
-This module is about computing the right kernel of matrices with polynomial entries.
-It provides various general purpose tools for constructing **solvers**. 
-
-By a **solver**, we mean a function of the following form:
-
-  .. function:: solver(mat, degrees=[], infolevel=0)
-
-  INPUT:
-  
-  - ``mat`` -- a matrix `A` over some polynomial ring
-  - ``degrees`` -- list of nonnegative integers, providing, for each variable `x`, a bound 
-    on the degree with which `x` is going to appear in the output. This data is optional. 
-    If set to ``[]``, the solver has to figure out the degrees by itself.
-  - ``infolevel`` -- a nonnegative integer indicating the desired verbosity (default=0), or
-    a pair ``(u, v)`` where ``u`` indicates the desired verbosity and ``v`` specifies how many
-    leading spaces should be added to the printed lines (default=0).
-  
-  OUTPUT:
-
-  - list of vectors `v` such that `A*v=0`
-
-Depending on the application from which a polynomial matrix arises, the preferred way of computing its
-nullspace may be different. The purpose of this module is not to provide a single solver which is
-good for every possible input, but instead, it provides a collection of functions by which solvers 
-can be composed. 
-
-For example, gauss_ is a function for constructing a solver using fraction free Gaussian elimination.
-It works for most polynomial domains. 
-
-::
-  
-  sage: my_solver = gauss()
-  sage: A = MatrixSpace(ZZ['x'], 4, 5).random_element()
-  sage: V = my_solver(A)
-  sage: A*V[0]
-  (0, 0, 0, 0)
-    
-  sage: A = MatrixSpace(ZZ['x', 'y'], 4, 5).random_element()
-  sage: V = my_solver(A)
-  sage: A*V[0]
-  (0, 0, 0, 0)
-    
-  sage: A = MatrixSpace(GF(1093)['x', 'y'], 4, 5).random_element()
-  sage: V = my_solver(A)
-  sage: A*V[0]
-  (0, 0, 0, 0)
-
-Several other functions create solvers which reduce the nullspace problem to one or more nullspace problems
-over simpler domains which are then solved by another solver. This solver can be chosen by the user. 
-
-For example, kronecker_ is a function for constructing a solver for matrices of multivariate polynomials.
[…]

### 2013-06-14 doc/ore_algebra.py

@@ -1,1720 +1,0 @@
-
-"""
-ore_algebra
-===========
-
-
-
-"""
-
-######### development mode ###########
-
-try:
-    if sys.modules.has_key('nullspace'):
-        del sys.modules['nullspace']
-except:
-    pass
-try:
-    if sys.modules.has_key('ore_operator'):
-        del sys.modules['ore_operator']
-except:
-    pass
-try:
-    if sys.modules.has_key('ore_operator_1_1'):
-        del sys.modules['ore_operator_1_1']
-except:
-    pass
-try:
-    if sys.modules.has_key('ore_operator_mult'):
-        del sys.modules['ore_operator_mult']
-except:
-    pass
-try:
-    if sys.modules.has_key('generalized_series'):
-        del sys.modules['generalized_series']
-except:
-    pass
-
-def taylor_shift_univ_int_poly(p, i):
-    ## assuming that p is an element of ZZ['x']
-    return p(p.parent().gen() + i)
-def taylor_shift_univ_int_ratfun(q, i):
-    ## assuming that q is an element of Fract(ZZ['x'])
-    num = taylor_shift_univ_int_poly(q.numerator(), i)
-    den = taylor_shift_univ_int_poly(q.denominator(), i)
-    return q.parent()(num, den, coerce=False, reduce=False)
-def taylor_shift_univ_modp_poly(p, i):
-    ## assuming that p is an element of GF(m)['x']
-    return p(p.parent().gen() + i)
-def taylor_shift_univ_modp_ratfun(q, i):
-    ## assuming that q is an element of Fract(GF(m)['x'])
-    num = taylor_shift_univ_modp_poly(q.numerator(), i)
-    den = taylor_shift_univ_modp_poly(q.denominator(), i)
-    return q.parent()(num, den, coerce=False, reduce=False)
-# if sage version is 5.10.beta3 or later, use taylor shift of flint2
-import datetime
-try:
-    v = version()
-except:
-    v = "2013-01-01"    
-if datetime.date(int(v[-10:-6]), int(v[-5:-3]), int(v[-2:])) >= datetime.date(2013, 05, 15):
-    try:
-        load("shift.spyx") # cython code defining 'taylor_shift_univ_int_poly' and 'taylor_shift_univ_modp_poly'
-    except:
-        pass
-
-#######################################
-
-from sage.structure.element import RingElement
-from sage.rings.ring import Algebra
-from sage.rings.ring import Ring 
[…]

### 2013-06-14 doc/ore_operator.py

@@ -1,1610 +1,0 @@
-
-"""
-ore_operator
-============
-
-"""
-
-from sage.structure.element import RingElement, canonical_coercion
-from sage.rings.arith import gcd, lcm
-from sage.rings.rational_field import QQ
-from sage.rings.integer_ring import ZZ
-from sage.rings.infinity import infinity
-
-class OreOperator(RingElement):
-    """
-    An Ore operator. This is an abstract class whose instances represent elements of ``OreAlgebra``.
-
-    In addition to usual ``RingElement`` features, Ore operators provide coefficient extraction
-    functionality and the possibility of letting an operator act on another object. The latter
-    is provided through ``__call__``.
-
-    """
-
-    # constructor
-
-    def __init__(self, parent, is_gen = False, construct=False):
-        RingElement.__init__(self, parent)
-        self._is_gen = is_gen
-
-    def __copy__(self):
-        """
-        Return a "copy" of self. This is just self, since in Sage
-        operators are immutable this just returns self again.
-        """
-        return self
-
-    # action
-
-    def __call__(self, f, **kwds):
-        """
-        Lets ``self`` act on ``f`` and returns the result.
-        The meaning of the action corresponding to the generator
-        of the Ore algebra can be specified with a keyword arguments
-        whose left hand sides are the names of the generator and the
-        right hand side some callable object. If no such information
-        is provided for some generator, a default function is used.
-        The choice of the default depends on the subclass. 
-
-        The parent of ``f`` must be a ring supporting conversion
-        from the base ring of ``self``. (There is room for generalization.)
-
-        EXAMPLES::
-
-           # In differential operator algebras, generators acts as derivations
-           sage: R.<x> = QQ['x']
-           sage: A.<Dx> = OreAlgebra(R.fraction_field(), "Dx")
-           sage: (Dx^5)(x^5) # acting on base ring elements
-           120
-           sage: (x*Dx - 1)(x)
-           0
-           sage: RR = PowerSeriesRing(QQ, "x", 5)
-           sage: 1/(1-RR.gen())
-           1 + x + x^2 + x^3 + x^4 + O(x^5)
-           sage: (Dx^2 - (5*x-3)*Dx - 1)(_) # acting on something else
-           4 + 6*x + 10*x^2 + O(x^3)
-
[…]

### 2013-06-14 doc/ore_operator_1_1.py

@@ -1,3519 +1,0 @@
-
-"""
-ore_operator_1_1
-================
-
-Special classes for operators living in algebras with one generator with base rings that also have
-one generator.
-
-"""
-
-from sage.structure.element import RingElement, canonical_coercion
-from sage.rings.arith import gcd, lcm
-from sage.rings.rational_field import QQ
-from sage.rings.integer_ring import ZZ
-from sage.rings.infinity import infinity
-from sage.rings.qqbar import QQbar
-from sage.rings.polynomial.polynomial_ring_constructor import PolynomialRing
-
-from ore_operator import *
-from generalized_series import *
-from generalized_series import _generalized_series_shift_quotient, _binomial ## why not implied by the previous line?
-
-class UnivariateOreOperatorOverUnivariateRing(UnivariateOreOperator):
-    """
-    Element of an Ore algebra with a single generator and a commutative rational function field as base ring.     
-    """
-    # Overview of dependencies between degree and denominator bounding functions:
-    #
-    #   * _degree_bound()          ==> requires indicial_polynomial(1/x)
-    #
-    #   * _denominator_bound()     ==> requires dispersion() and implements Abramov's algorithm
-    #                                  Not universally correct. May have to be adapted by subclasses.
-    #
-    #   * indicial_polynomial(p)   ==> requires _coeff_list_for_indicial_polynomial when p is x or 1/x
-    #                                  ABSTRACT for other arguments p.
-    #
-    #   * dispersion(p)            ==> requires spread(p)
-    #
-    #   * _coeff_list_for_ind...   ==> ABSTRACT
-    #
-    #   * spread(p)                ==> ABSTRACT
-    #
-
-    def __init__(self, parent, *data, **kwargs):
-        super(UnivariateOreOperator, self).__init__(parent, *data, **kwargs)
-
-    def _normalize_base_ring(self):
-        """
-        Rewrites ``self`` into an operator from an algebra whose base ring is a univariate
-        polynomial ring over a field.
-
-        Returns a tuple ``(A, R, K, L)`` where
-
-         * ``L`` is the new operator
- 
-         * ``A`` is the parent of ``L``
-
-         * ``R`` is the base ring of ``A``
-
-         * ``K`` is the base ring of ``R``
-
-        """
[…]

### 2013-06-14 doc/ore_operator_mult.py

@@ -1,171 +1,0 @@
-
-"""
-ore_operator_mult
-=================
-
-"""
-
-from sage.structure.element import RingElement, canonical_coercion
-from sage.rings.arith import gcd, lcm
-from sage.rings.rational_field import QQ
-from sage.rings.integer_ring import ZZ
-from sage.rings.infinity import infinity
-
-from ore_operator import *
-
-class MultivariateOreOperator(OreOperator):
-    """
-    An Ore operator. Instances of this class represent elements of Ore algebras with more than
-    one generator.
-    """
-
-    # constructor
-
-    def __init__(self, parent, data): 
-        OreOperator.__init__(self, parent)
-        if isinstance(data, OreOperator):
-            data = data.polynomial()        
-        self.__poly = parent.associated_commutative_algebra()(data)
-
-    # action
-
-    def __call__(self, f, **kwds):
-        raise NotImplementedError
-
-    # tests
-
-    def __nonzero__(self):
-        return not self.__poly.is_zero()
-
-    def __neq__(self, other):
-        return not (self == other)
-
-    def __eq__(self, other):
-
-        if len(self.exponents()) == 1:
-            return self.constant_coefficient() == other
-        elif not isinstance(other, OreOperator):
-            return False
-        elif self.parent() is other.parent():
-            return self.__poly == other.__poly
-        else:
-            try:
-                A, B = canonical_coercion(self, other)
-                return A == B
-            except:
-                return False
-
-    def _is_atomic(self):
-        return self.__poly._is_atomic()
-       
-    def is_gen(self):
-        return self.__poly.is_gen()
-
-    # conversion
-
-    def polynomial(self):
-        return self.__poly            
-
-    def change_ring(self, R):
-        """
-        Return a copy of this operator but with coefficients in R, if at
-        all possible.
-
-        EXAMPLES::
-
-          sage: R.<x> = QQ['x']
-          sage: A.<Dx> = OreAlgebra(R, 'Dx')
-          sage: op = Dx^2 + 5*x*Dx + 1
-          sage: op.parent()
-          Univariate Ore algebra in Dx over Univariate Polynomial Ring in x over Rational Field
-          sage: op = op.change_ring(R.fraction_field())
-          sage: op.parent()
-          Univariate Ore algebra in Dx over Fraction Field of Univariate Polynomial Ring in x over Rational Field
-        
-        """
-        if R == self.base_ring():
[…]

### 2013-06-14 spkg/setup.py

@@ -1,0 +1,12 @@
+
+from distutils.core import setup
+
+setup(
+    name = "ore_algebra",
+    version = "1.0",
+    author = "Manuel Kauers, Maximilian Jaroschek, Fredrik Johansson",
+    author_email = "manuel@kauers.de",
+    licence = "GPL",
+    packages = ["ore_algebra"]
+    )
+    

### 2013-06-14 paper/main.tex
minor changes in doc.
for the paper, Makefile and main.tex changed to proper loading of package.

@@
 \begin{sageexample}
-  sage: load("ore_algebra.py", "guessing.py", "ore_operator.py");  #### TO BE FIXED ####
+  sage: #load("ore_algebra.py", "guessing.py", "ore_operator.py");  #### TO BE FIXED ####
+  sage: from ore_algebra import *
   sage: R.<x> = PolynomialRing(ZZ); A.<Dx> = OreAlgebra(R)
@@
   sage: L1,L2 = A.random_element(3),A.random_element(2)    
-  sage: L1.gcrd(L2, prs=__primitivePRS__)
-  sage: L1.gcrd(L2, prs=__classicPRS__)
+  sage: #L1.gcrd(L2, prs=__primitivePRS__)
+  sage: #L1.gcrd(L2, prs=__classicPRS__)
 \end{sageexample}

### 2013-06-14 __init__.py

@@ -9,6 +9,7 @@
 #  http://www.gnu.org/licenses/                                             #
 #############################################################################
 
+
 from ore_algebra import OreAlgebra
 from guessing import guess, guess_raw
 

### 2013-06-14 doc/conf.py

@@ -1,0 +1,196 @@
+# -*- coding: utf-8 -*-
+#
+# ore_algebra documentation build configuration file, created by
+# sphinx-quickstart on Sat Feb  2 16:17:05 2013.
+#
+# This file is execfile()d with the current directory set to its containing dir.
+#
+# Note that not all possible configuration values are present in this
+# autogenerated file.
+#
+# All configuration values have a default; values that are commented out
+# serve to show the default.
+
+import sys, os
+
+# If extensions (or modules to document with autodoc) are in another directory,
+# add these directories to sys.path here. If the directory is relative to the
+# documentation root, use os.path.abspath to make it absolute, like shown here.
+sys.path.append(os.path.abspath('.'))
+
+# -- General configuration -----------------------------------------------------
+
+# Add any Sphinx extension module names here, as strings. They can be extensions
+# coming with Sphinx (named 'sphinx.ext.*') or your custom ones.
+extensions = ['sphinx.ext.autodoc', 'sphinx.ext.doctest', 'sphinx.ext.todo', 'sphinx.ext.mathjax']
+
+# Add any paths that contain templates here, relative to this directory.
+templates_path = ['_templates']
+
+# The suffix of source filenames.
+source_suffix = '.rst'
+
+# The encoding of source files.
+#source_encoding = 'utf-8'
+
+# The master toctree document.
+master_doc = 'index'
+
+# General information about the project.
+project = u'ore_algebra'
+copyright = u'2013, Manuel Kauers, Maximilian Jaroschek, Fredrik Johansson'
+
+# The version info for the project you're documenting, acts as replacement for
+# |version| and |release|, also used in various other places throughout the
+# built documents.
+#
+# The short X.Y version.
+version = '0.1'
+# The full version, including alpha/beta/rc tags.
+release = '0.1'
+
+# The language for content autogenerated by Sphinx. Refer to documentation
+# for a list of supported languages.
+#language = None
+
+# There are two options for replacing |today|: either, you set today to some
+# non-false value, then it is used:
+#today = ''
+# Else, today_fmt is used as the format for a strftime call.
+#today_fmt = '%B %d, %Y'
+
+# List of documents that shouldn't be included in the build.
+#unused_docs = []
+
+# List of directories, relative to source directory, that shouldn't be searched
+# for source files.
+exclude_trees = ['_build']
+
+# The reST default role (used for this markup: `text`) to use for all documents.
+#default_role = None
+
[…]

### 2013-06-14 ore_algebra.py

@@ -3,6 +3,34 @@
 ore_algebra
 ===========
 
+The ore_algebra package provides functionality for doing computations with Ore polynomials.
+
+Ore polynomials are operators which can be used to describe special functions or combinatorial
+sequences. Typical examples are linear differential operators with polynomial coefficients.
+
+Ore polynomials are elements of Ore algebras. Ore algebras are ring objects created by the
+function ``OreAlgebra`` as described below. 
+
+Depending on the particular parent algebra, Ore polynomials may support different functionality.
+For example, for Ore polynomials representing recurrence operators, there is a method for 
+computing interlacing operatos, an operation which does not make sense for differential operators.
+
+The typical user will only need two functions defined in the package: 
+
+ * ``OreAlgebra`` -- for creating a new Ore algebra object. 
+ * ``guess`` -- for fitting an Ore polynomial to a given set of data.
+
+Ore polynomials are created using ``OreAlgebra`` objects, and most of the functionality for
+doing calculations with Ore polynomials is available in the methods attached to them. 
+
+For examples and further information, see the docstring of ``OreAlgebra`` below, or the 
+tutorial paper /Ore Polynomials in Sage/ by the authors. 
+
+
+AUTHOR:
+
+ - Manuel Kauers, Maximilian Jaroschek, Fredrik Johansson (2013-06-15)
+
 """
 
 #############################################################################
@@ -18,31 +46,13 @@ ore_algebra
 
 ######### development mode ###########
 
-try:
-    if sys.modules.has_key('nullspace'):
-        del sys.modules['nullspace']
-except:
-    pass
-try:
-    if sys.modules.has_key('ore_operator'):
-        del sys.modules['ore_operator']
-except:
-    pass
-try:
-    if sys.modules.has_key('ore_operator_1_1'):
-        del sys.modules['ore_operator_1_1']
-except:
-    pass
-try:
-    if sys.modules.has_key('ore_operator_mult'):
-        del sys.modules['ore_operator_mult']
-except:
-    pass
-try:
-    if sys.modules.has_key('generalized_series'):
-        del sys.modules['generalized_series']
-except:
-    pass
+if True:
+    # let load("ore_algebra") trigger reload of the modules in the list below
+    for mod in ['nullspace', 'ore_operator', 'ore_operator_1_1', 'ore_operator_mult', 'generalized_series']:
+        try:
+            del sys.modules[mod]
+        except:
+            pass
 
 def taylor_shift_univ_int_poly(p, i):
     ## assuming that p is an element of ZZ['x']

### 2013-06-14 ore_operator_1_1.py

@@ -473,7 +473,7 @@ class UnivariateOreOperatorOverUnivariateRing(UnivariateOreOperator):
         The output is not unique. With low probability, the leading coefficient degree in the output
         may not be minimal. 
 
-        EXAMPLES:
+        EXAMPLES::
 
           sage: R.<n> = ZZ['n']
           sage: A.<Sn> = OreAlgebra(R, 'Sn')

### 2013-06-14 ore_operator_mult.py

@@ -47,23 +47,8 @@ class MultivariateOreOperator(OreOperator):
     def __nonzero__(self):
         return not self.__poly.is_zero()
 
-    def __neq__(self, other):
-        return not (self == other)
-
-    def __eq__(self, other):
-
-        if len(self.exponents()) == 1:
-            return self.constant_coefficient() == other
-        elif not isinstance(other, OreOperator):
-            return False
-        elif self.parent() is other.parent():
-            return self.__poly == other.__poly
-        else:
-            try:
-                A, B = canonical_coercion(self, other)
-                return A == B
-            except:
-                return False
+    def __cmp__(self, other):
+        return cmp(self.__poly, other.__poly)
 
     def _is_atomic(self):
         return self.__poly._is_atomic()

### 2013-06-14 paper/main.tex
Added introduction to section 4 of the paper. Changed the way a polynomial remainder sequence is chosen.

--- 2013-06-13 paper/main.tex
@@
 
+Besides basic arithmetic for Ore operators, the package provides a wide range of methods to create, manipulate and solve several different kinds of functional equations. 
+Some of these methods are accessible in any Ore algebra while others are tied specifically to e.g.\ recurrence operators or differential operators.
+In these predefined algebras, operators can act on suitable objects like functions or sequences in a natural way:
+
+\begin{sageexample}
+  sage: R.<x> = PolynomialRing(QQ)
+  sage: A = OreAlgebra(R, 'Dx'); Dx=A.gen() 
+  sage: p,q = R.random_element(3),R.random_element(1)
+  sage: L = p*Dx - p.derivative()
+  sage: L(p)
+  sage: L(q)
+\end{sageexample}
+
+It is also possible to define actions for other Ore algebras or to redefine the action in the predefined ones.
+
+\begin{sageexample}
+  sage: R.<x> = PolynomialRing(QQ)
+  sage: A = OreAlgebra(R, ('S', {x:x+1}, {x:1})); S=A.gen()
+  sage: B = OreAlgebra(R, 'Dx'); Dx=B.gen()
+  sage: S(x-1,action=lambda x: 2*x)
+  sage: Dx(x,action=lambda x: x)
+  sage: Dx(x)
+\end{sageexample}
+
+We give a small overview of the functionality of some selected methods.
+
 \subsection{gcrd and lclm}
@@
 left multiple (LCLM) of two Ore polynomials can be computed. The package provides a number of methods to carry 
-out these tasks
+out these tasks.
 
@@
     \verb|xgcrd(B)| & returns the greatest common right divisor of $A$ and $B$ and the according B\'ezout coefficients.\\
-    \verb|lclm(B)| & returns the least common left multiple of $A$ and $B$.\\\hline
+    \verb|lclm(B)| & returns the least common left multiple of $A$ and $B$.\\
+    \verb|xlclm(B)| & returns the least common left multiple $L$ of $A$ and $B$ and the left quotients of $L$ and $A$ and of $L$ and $B$.\\\hline
   \end{tabular}
@@
 
-All these methods are also available for Ore operators living in univariate Ore algebras over a base ring $R$ which 
+All these methods are also available for Ore operators living in univariate Ore algebras over a base ring $\set R$ which 
 does not necessarily have to be a field. The operators are then implicitly assumed to live in the respective Ore 
-algebra over the quotient field $K$ of $R$. The output will be the GCRD (LCLM, quotient, remainder) in the Ore 
-algebra over $K$ but not over $R$, in which these objects might not exist or might not be computable.
+algebra over the quotient field $\set K$ of $\set R$. The output will be the GCRD (LCLM, quotient, remainder) in the Ore 
+algebra over $\set K$ but not over $\set R$, in which these objects might not exist or might not be computable.
 
@@
   sage: L1,L2 = A.random_element(3),A.random_element(2)    
-  sage: L1.gcrd(L2, prs=__primitivePRS__)
-  sage: L1.gcrd(L2, prs=__classicPRS__)
+  sage: L1.gcrd(L2, prs="primitive")
+  sage: L1.gcrd(L2, prs="monic")
 \end{sageexample}

### 2013-06-14 paper/main.tex

@@
 \begin{sageexample}
-  sage: load("ore_algebra.py", "guessing.py", "ore_operator.py");  #### TO BE FIXED ####
+  sage: from ore_algebra import *
   sage: R.<x> = PolynomialRing(ZZ); A.<Dx> = OreAlgebra(R)
@@
 
-\section{Guessing}\label{sec:5}
-
-[[CITE RUBEY.]]
-
-Guessing is one of the most popular features of software packages like gfun.  It
-is a technique for detecting patterns in sequences of numbers, which is often
-used in combinatorics. The basic idea is simple. If we are given a finite array
-of numbers, though of as the first terms of an infinite sequence, we may want to
-know whether this sequence satisfies a recurrence. The algorithm behind a
-guessing engine searches for small equations matching the given data.
-Generically, no such equations exist, so if some are found, it is fair to
-``guess'' that they are in fact valid equations for the whole infinite sequence.
-=======
 \subsection{Guessing}
@@
 equations for the whole infinite sequence.
->>>>>>> manuel/master
 

### 2013-06-15 generalized_series.py
minor bugfixes in code and docstrings.
resultant moved from UnivariateOreOperatorOverUnivariateRing to UnivariateOreOperator

@@ -648,20 +648,20 @@ class ContinuousGeneralizedSeries(RingElement):
                 if alpha.is_zero():
                     pass
                 elif alpha.is_one():
-                    E_rep += "*" + x
+                    E_rep += "\\cdot " + x
                 elif alpha in QQ:
-                    E_rep += "*" + x + "^{" + str(alpha) + "}" # prefer 1/3 over \frac{1}{3} in exponents
+                    E_rep += "\\cdot " + x + "^{" + str(alpha) + "}" # prefer 1/3 over \frac{1}{3} in exponents
                 else:
-                    E_rep += "*" + x + "^{" + alpha._latex_() + "}"
+                    E_rep += "\\cdot " + x + "^{" + alpha._latex_() + "}"
 
             if T.is_one():
                 rep = E_rep
             elif (-T).is_one():
                 rep = '-' + E_rep
             elif T._is_atomic():
-                rep = E_rep + "*" + T_rep
+                rep = E_rep + "\\cdot " + T_rep
             else:
-                rep = E_rep + r"*\Bigl(" + T_rep + r"\Bigr)"
+                rep = E_rep + r"\cdot\Bigl(" + T_rep + r"\Bigr)"
 
         if r > 1:
             # x^{1/3}^{17} --> x^{17/3}

### 2013-06-15 ore_operator.py

@@ -376,7 +376,7 @@ class OreOperator(RingElement):
 
     def constant_coefficient(self):
         """
-        Return the leading coefficient of this operator. 
+        Return the coefficient of `\partial^0` of this operator. 
         """
         raise NotImplementedError
 
@@ -839,7 +839,7 @@ class UnivariateOreOperator(OreOperator):
             return self.parent().one()
         elif self.parent() is not other.parent():
             A, B = canonical_coercion(self, other)
-            return A.gcrd(B)
+            return A.gcrd(B, **kwargs)
 
         prs = kwargs["prs"] if kwargs.has_key("prs") else None
 
@@ -854,7 +854,7 @@ class UnivariateOreOperator(OreOperator):
                    "classic" : __classicPRS__,
                    "subresultant" : __subresultantPRS__,
                    "monic" : __monicPRS__,
-        }
+                   }
 
         try:
             prs = prslist[prs]
@@ -868,16 +868,13 @@ class UnivariateOreOperator(OreOperator):
         while not r[1].is_zero():
             (r2,q,alpha,beta,correct)=prs(r,additional)
             if not correct:
-                print "switch to primitve PRS"
+                print "switching to primitve PRS"
                 prs = __primitivePRS__
             else:
                 r=r2
         r=r[0]
 
-        if not prs==__classicPRS__:
-            r = r.primitive_part()
-
-        return r
+        return r.normalize()
 
     def xgcrd(self, other, prs=None):
         """
@@ -936,7 +933,7 @@ class UnivariateOreOperator(OreOperator):
         while not r[1].is_zero():  
             (r2, q, alpha, beta, correct) = prs(r, additional)
             if not correct:
-                print "switch to primitve PRS"
+                print "switching to primitve PRS"
                 prs = __primitivePRS__
             else:
                 r = r2; bInv = ~beta
@@ -1151,6 +1148,63 @@ class UnivariateOreOperator(OreOperator):
             L0 = L.change_ring(K)
         
         return (L, L0 // A, L0 // B)
+
+    def resultant(self, other):
+        """
+        Returns the resultant between this operator and ``other``. 
+
+        INPUT:
+        
[…]

### 2013-06-15 ore_operator_1_1.py

@@ -77,7 +77,7 @@ class UnivariateOreOperatorOverUnivariateRing(UnivariateOreOperator):
             R = R.ring()
 
         if not K.is_field():
-            R.change_ring(K.fraction_field())
+            R = R.change_ring(K.fraction_field())
 
         L = L.change_ring(R)
         return L.parent(), R, K, L
@@ -550,6 +550,8 @@ class UnivariateOreOperatorOverUnivariateRing(UnivariateOreOperator):
         where `F_k` denotes the Fibonacci sequence. The rational function `m` does not appear
         in the closed form, it can be regarded as a certificate.         
 
+        The method returns the empty list if and only if no nontrivial solutions exist. 
+
         This function may not be implemented for every algebra.
 
         EXAMPLES::
@@ -637,62 +639,6 @@ class UnivariateOreOperatorOverUnivariateRing(UnivariateOreOperator):
             sol[i] = (M, rat)
 
         return filter(lambda p: p is not None, sol)
-
-    def resultant(self, other):
-        """
-        Returns the resultant between this operator and ``other``. 
-
-        INPUT:
-        
-        - ``other`` -- some operator that lives in the same algebra as ``self``.
-        
-        OUTPUT:
-
-        The resultant between ``self`` and ``other``, which is defined as the determinant of the
-        `(n+m) x (n+m)` matrix `[ A, D*A, ..., D^(m-1)*A, B, D*B, ..., D^(n-1)*B ]` constructed
-        from the coefficient vectors of the operators obtained from ``self`` and ``other`` by
-        multiplying them by powers of the parent's generator. 
-
-        EXAMPLES::
-
-           sage: R.<x> = ZZ['x']
-           sage: A.<Dx> = OreAlgebra(R, 'Dx')
-           sage: L1 = (5*x+3)*Dx^3 + (7*x+4)*Dx^2 + (3*x+2)*Dx + (4*x-1)
-           sage: L2 = (8*x-7)*Dx^2 + (x+9)*Dx + (2*x-3)
-           sage: L1.resultant(L2)
-           2934*x^5 - 8200*x^4 + 32161*x^3 - 83588*x^2 - 67505*x + 42514
-           sage: L2.resultant(L1)
-           -2934*x^5 + 8200*x^4 - 32161*x^3 + 83588*x^2 + 67505*x - 42514
-
-           sage: R.<x> = ZZ['x']
[…]

### 2013-06-15 paper/main.tex
paper: more stuff added to section 4; gaps closed.

--- 2013-06-14 paper/main.tex
@@
-\documentclass{amsart}
+\documentclass[11pt]{amsart}
 \usepackage{amsmath}
@@
 computer algebra systems, for example in the Maple packages
-OreTools~\cite{abramov03}, gfun~\cite{salvy94} or mgfun~\cite{chyzak98a}, in the
-Mathematica packages by Mallinger~\cite{mallinger96} and
+OreTools~\cite{abramov03}, gfun~\cite{salvy94} or mgfun~\cite{chyzak98a}, or in
+the Mathematica packages by Mallinger~\cite{mallinger96} and
 Koutschan~\cite{koutschan10c,koutschan09}.
@@
 who want to try it out are invited to download the current latest version from 
-[URL] and are encouraged to send us bug reports and other comments. We hope 
+\begin{center}
+  URL
+\end{center}
+and are encouraged to send us bug reports and other comments. We hope 
 that the community will find the code useful.
@@
 \begin{sageexample}
-  sage: #load("ore_algebra.py", "guessing.py", "ore_operator.py");  #### TO BE FIXED ####
   sage: from ore_algebra import *
@@
 
-More details on the construction of Ore algebras are given in the following section.
-Also Ore algebras with several generators are already supported. However, so far we offer
-no functionality beyond addition and multiplication for these. More functionality 
-is available for algebras with one generator. We describe some of it in 
-Section~\ref{sec:3}. Even more functionality has been implemented for special 
-algebras like the ones created above, which are most frequently used in practice. 
-Section~\ref{sec:4} describes some of these features.
+More details on the construction of Ore algebras are given in the following
+section. The construction and manipulation of elements of Ore algebras is
+discussed in Section~\ref{sec:3}.
 
-/// disclaimer: we don't explain algorithms here, but only the software ///
+The package also supports Ore algebras with several generators already. However,
+so far we offer no functionality beyond addition and multiplication for
+these. Much more functionality is available for algebras with one generator. We
+describe some of it in Section~\ref{sec:4}.
 
@@
 \begin{center}
-  \begin{tabular}{|l|p{.55\hsize}|}
+  \begin{tabular}{|l|p{.5\hsize}|}
     \hline
@@
 
-\smallskip
+\smallskip\goodbreak
 
@@
   sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
-  sage: Ac = A.associated_commutative_algebra(); Ac
+  sage: Ac = A.associated_commutative_algebra()
+  sage: Ac
   sage: A(Ac.random_element())
@@
 
-In accordance with the default Sage coercion model, such conversions take place automatically (if possible) 
+In accordance with the Sage coercion model, such conversions take place automatically (if possible) 
 when operators from different algebras are added or multiplied. Note that the result of such an operation
@@
 \begin{sageexample}
-  sage: A = OreAlgebra(ZZ['t']['x'], 'Dx'); 
+  sage: A = OreAlgebra(ZZ['t']['x'], 'Dx')
   sage: B = OreAlgebra(QQ['x'].fraction_field(), 'Dx')
@@
 
-\subsection{gcrd and lclm}
+Besides basic arithmetic for Ore operators, the package provides a wide range of
+methods to create, manipulate and solve several different kinds of operators.
+Some of these methods are accessible in any Ore algebra while others are tied
+specifically to e.g.\ recurrence operators or differential operators.
 
-A univariate Ore algebra over a field is a left Euclidean domain, which means that it is possible to perform 
-left division with remainder. Building upon this, the greatest common right divisor (GCRD) and the least common 
-left multiple (LCLM) of two Ore polynomials can be computed. The package provides a number of methods to carry 
-out these tasks
+In this section, we give an overview of some of the functionality provided by
+the package. Because of space limitation, only some of the available methods
+can be discussed here. For further information, we refer to the documentation.
+
+\subsection{Methods for General Algebras}
+
+A univariate Ore algebra over a field is a left Euclidean domain, which means
+that it is possible to perform left division with remainder. Building upon this,
+the greatest common right divisor (GCRD) and the least common left multiple
+(LCLM) of two Ore polynomials can be computed. The package provides a number of
+methods to carry out these tasks.
+
+\begin{center}
+  \begin{tabular}{|l|p{.65\hsize}|}
+    \hline
+    Method name & short description \\\hline
+    \verb|A.quo_rem(B)| & returns the left quotient and the left remainder of $A$ and $B$.\\
+    \verb|A.gcrd(B)| & returns the greatest common right divisor of $A$ and $B$.\\
+    \verb|A.xgcrd(B)| & returns the greatest common right divisor of $A$ and $B$ and the according B\'ezout coefficients.\\
+    \verb|A.lclm(B)| & returns the least common left multiple of $A$ and $B$.\\
+    \verb|A.xlclm(B)| & returns the least common left multiple $L$ of $A$ and $B$ and the left quotients of $L$ and $A$ and of $L$ and $B$.\\
+    \verb|A.resultant(B)| & returns the resultant of $A$ and $B$ (see \cite{li96} for further information).\\
+    \hline
+  \end{tabular}
+\end{center}
+
+All these methods are also available for Ore operators living in univariate Ore
+algebras over a base ring $R$ which does not necessarily have to be a
+field. The operators are then implicitly assumed to live in the respective Ore
+algebra over the quotient field $K$ of~$R$. The output will be the
+GCRD (LCLM, quotient, remainder) in the Ore algebra over $K$ but not over
+$R$, in which these objects might not exist or might not be computable.
+
+\begin{sageexample}
+  sage: A = OreAlgebra(ZZ['n'], 'Sn')
+  sage: G = A.random_element(2)
+  sage: L1, L2 = A.random_element(7), A.random_element(5)
+  sage: while L1.gcrd(L2) != 1: L2 = A.random_element(5)                       
+  sage: L1, L2 = L1*G, L2*G                                                        
+  sage: L1.gcrd(L2) == G.normalize()
+  sage: L3, S, T = L1.xgcrd(L2)                             
+  sage: S*L1 + T*L2 == L3
+  sage: LCLM = L1.lclm(L2)
+  sage: LCLM % L1 == LCLM % L2 == 0
+  sage: LCLM.order() == L1.order() + L2.order() - G.order()
+\end{sageexample}
+
+The GCRD is only unique up to multiplication (from the left) by an element from
+the base ring. The method \verb|normalize| called in line~6 of the listing above
+multiplies a given operator from the left by some element from the base ring
+such as to produce a canonical representative of the set of all the operators
+that can be obtained from the given one by left multiplication of a base ring
+element. This facilitates the comparison of output.
+
+There are a number of options to adjust the functionality of the methods. For
+example, the efficiency of computing the GCRD depends on the size of the
+coefficients of intermediate results and there are different strategies to
+control this in so called polynomial remainder sequences (PRS). The default is
+the improved PRS described in~\cite{jaroschek13b}. which will usually be the
+fastest choice.
+
+\begin{sageexample}
+  sage: A = OreAlgebra(ZZ['n'], 'Sn')
+  sage: L1, L2 = A.random_element(3), A.random_element(2)
+  sage: algs = ["improved", "classic", "monic", "subresultant"]
+  sage: [L1.gcrd(L2, prs=a) for a in algs]
+\end{sageexample}
+
+If $L_1,L_2$ are operators, then the solutions of their GCRD are precisely the
+common solutions of $L_1$ and~$L_2$. The LCLM, on the other hand, is the minimal
+order operator whose solution space contains all the solutions of $L_1$ as well
+as all the solutions of~$L_1$. Because of this property, GCRD and LCLM are
+useful tools for constructing operators with prescribed solutions. For example,
+here is how to construct a differential operator which has the solutions $x^5$
+and $\exp(x)$ starting from the obvious operators annihilating $x^5$
+and~$\exp(x)$, respectively.
+
+\begin{sageexample}
+  sage: R.<x> = ZZ[]; A.<Dx> = OreAlgebra(R)
+  sage: L = (Dx - 1).lclm(x*Dx - 5)
+  sage: L
+  sage: L(x^5)
+  sage: L(exp(x)).full_simplify() 
+\end{sageexample}
+
+Observe how in the last two lines we apply the operator~$L$ to other
+objects. Such applications are not defined for every algebra and in general have
+to be specified by the user through an optional argument:
+
+\begin{sageexample}
+  sage: A.<Qqn> = OreAlgebra(ZZ['q']['qn'])
+  sage: var('q', 'n', 'x')
+  sage: (Qqn^2+Qqn+1)(q^n, action=lambda expr: expr.substitute(n=n+1))
+  sage: (Qqn^2+Qqn+1)(x, action=lambda expr: expr.substitute(x=q*x))
+\end{sageexample}
+
+Thanks to the LCLM operation discussed above, we have the property that when $f$
+and $g$ are two objects which are annihilated by some operators $L_1,L_2$
+belonging to some Ore algebra~$A$ then this algebra contains also an operator
+which annihilates their sum~$f+g$. Thus the class of solutions of operators of
+$A$ is closed under addition. It turns out that similar closure properties hold
+for other operations. The following table lists some of the corresponding
+methods. Methods for some further closure properties will appear further below.
 
@@
 Method name & short description \\\hline
-    \verb|quo_rem(B)| & returns the left quotient and the left remainder of $A$ and $B$.\\
-    \verb|gcrd(B)| & returns the greatest common right divisor of $A$ and $B$.\\
-    \verb|xgcrd(B)| & returns the greatest common right divisor of $A$ and $B$ and the according B\'ezout coefficients.\\
-    \verb|lclm(B)| & returns the least common left multiple of $A$ and $B$.\\\hline
+\verb|lclm()| & computes an annihilating operator for $f+g$ from annihilating operators for $f$ and $g$ \\
+\verb|symmetric_product()| & computes an annihilating operator for $fg$ from annihilating operators for $f$ and $g$ \\
+\verb|symmetric_power()| & computes an annihilating operator for $f^n$ from an annihilating operator for $f$ and a given positive integer~$n$\\ 
+\verb|annihilator_of_associate()| & computes an annihilating operator for $M(f)$ from an annihilating operator for $f$ and a given operator~$M$\\ 
+\verb|annihilator_of_polynomial()| & computes an annihilating operator for the object $p(f,\partial f,\partial^2 f,\dots)$ from an annihilating operator for $f$ and a given multivariate polynomial~$p$.
+\\
+\hline
 \end{tabular}
@@
 
-All these methods are also available for Ore operators living in univariate Ore algebras over a base ring $R$ which 
-does not necessarily have to be a field. The operators are then implicitly assumed to live in the respective Ore 
-algebra over the quotient field $K$ of $R$. The output will be the GCRD (LCLM, quotient, remainder) in the Ore 
-algebra over $K$ but not over $R$, in which these objects might not exist or might not be computable.
+As an example application, let us prove Cassini's identity for Fibonacci
+numbers: 
+\[
+  F_{n+1}^2 - F_nF_{n+2}=(-1)^n.
+\]
+The idea is to derive, using commands
+from the table above, a recurrence satisfied by the left hand side and then show
+that this recurrence is also valid for the right hand side.
 
 \begin{sageexample}
-  sage: A = OreAlgebra(ZZ['n'], 'Sn');                                           
-  sage: G,L1,L2 = A.random_element(2),A.random_element(7),A.random_element(5)    
-  sage: while not L1.gcrd(L2)==1: L2 = A.random_element(5)                       
-  sage: L1,L2 = L1*G,L2*G                                                        
-  sage: (L3,S,T) = L1.xgcrd(L2)                             
-  sage: (S*L1+T*L2).normalize() == L3.normalize() == G.normalize()
-  sage: LCLM = L1.lclm(L2)
-  sage: LCLM.quo_rem(L1)[1] == LCLM.quo_rem(L2)[1] == A.zero()
+  sage: A.<Sn> = OreAlgebra(ZZ['n'])
+  sage: fib = Sn^2 - Sn - 1
+  sage: R.<x0,x1,x2> = ZZ['n']['x0','x1','x2']
+  sage: fib.annihilator_of_polynomial(x1^2 - x0*x2) 
 \end{sageexample}
 
-There are a number of options to adjust the functionality of the methods. For example, the efficiency of computing the GCRD 
-depends on the size of the coefficients of intermediate results and there are different strategies to control this in so called 
-polynomial remainder sequences (PRS). The default is the improved PRS described in \cite{jaroschek13b}. which will usually be 
-the fastest choice.
+As this operator obviously annihilates $(-1)^n$, the proof is complete after
+checking that the identity holds for $n=0$. Another way of carrying out the
+same computation using the other commands would be as follows.
 
 \begin{sageexample}
-  sage: A = OreAlgebra(ZZ['n'], 'Sn');                                           
-  sage: L1,L2 = A.random_element(3),A.random_element(2)    
-  sage: #L1.gcrd(L2, prs=__primitivePRS__)
-  sage: #L1.gcrd(L2, prs=__classicPRS__)
+  sage: A.<Sn> = OreAlgebra(ZZ['n'])
+  sage: fib = Sn^2 - Sn - 1
+  sage: L1 = fib.annihilator_of_associate(Sn).symmetric_power(2)
+  sage: L2 = fib.annihilator_of_associate(Sn^2).symmetric_product(fib)
+  sage: L1.lclm(L2)
 \end{sageexample}
 
-\subsection{closure properties?}
+Observe that the resulting operator again annihilates~$(-1)^n$, but its order is
+higher than the operator obtained before, so we need to check more initial
+values to complete the proof. For larger computations, the command
+\verb|annihilator_of_polynomial| would also consume less computation time than
+the step-by-step approach.
 
--- table mapping function names to arithmetic
+\subsection{Methods for Special Algebras}
 
--- example for how to prove some identity with the methods provided.
-
--- merge with previous section?
-
-\subsection{conversions?}
-
--- re2de and all its relatives
-
-\subsection{solutions?}
-
--- polynomial, rational, generalized series solutions
-
--- available for differential and shift algebras
-
-\subsection{action?}
-
--- should be mentioned somewhere, although there is not much to say about it.
-
-\subsection{Differential Operators}
-
-Useful methods available for differential operators:
+For the elements of some of the most important algebras, additional methods are
+available. The following table lists some of the additional methods available
+for differential operators, i.e., elements of an Ore algebra of the
+form~$R[x]\<Dx>$ or $K(x)\<Dx>$.
 
@@
     \verb|annihilator_of_composition()| & converts an annihilator for $f(x)$ to one for $f(a(x))$ where $a(x)$ is algebraic over the base ring\\
+    \verb|desingularize()| & computes a left multiple of this operator with polynomial coefficients and 
+           lowest possible leading coefficient degree\\
+    \verb|associate_solutions(p)| & applied to an operator~$P$, this computes, if possible, an operator $M$
+    and a rational function $m$ such that $DM=p+mP$ (see \cite{abramov99} for further information)\\
+    \verb|polynomial_solutions()| & computes the polynomial solutions of this operator\\
+    \verb|rational_solutions()| & computes the rational function solutions of this operator\\
     \verb|power_series_solutions()| & computes power series solutions of this operator\\
@@
 
-We compute an annihilator for the error function $\frac{2}{\sqrt{\pi}} \int_0^x e^{-t^2} dt$,
-starting from the differential equation for $e^x$, and produce the
-recurrence for the Taylor series coefficients at the origin:
+As an example application, we compute an annihilator for the error function
+$\frac{2}{\sqrt{\pi}} \int_0^x \exp(-t^2) dt$, starting from the differential
+equation for $\exp(x)$, and produce the recurrence for the Taylor series
+coefficients at the origin. Finally, we compute the series solutions of the
+differential operator at infinity.
 
@@
   sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R, 'Dx')                            
-  sage: (Dx - 1).annihilator_of_composition(-x**2)
-  Dx + 2*x
-  sage: (Dx + 2*x).annihilator_of_integral()                                     
-  Dx^2 + 2*x*Dx
-  sage: (Dx^2 + 2*x*Dx).to_S(OreAlgebra(ZZ['n'], 'Sn'))                          
-  (n^2 + 3*n + 2)*Sn^2 + 2*n
-  sage: (Dx^2 + 2*x*Dx).power_series_solutions(10)                               
+  sage: (Dx - 1).annihilator_of_composition(-x^2)
+  sage: L = (Dx + 2*x).annihilator_of_integral()                                     
+  sage: L
+  sage: L.to_S(OreAlgebra(ZZ['n'], 'Sn'))                          
+  sage: L.power_series_solutions(10)                               
   [x - 1/3*x^3 + 1/10*x^5 - 1/42*x^7 + 1/216*x^9 + O(x^10), 1 + O(x^10)]
+  sage: L.annihilator_of_composition(1/x).generalized_series_solutions()
 \end{sageexample}
 
-\subsection{Computation of Sequence Terms}
+The last output implies that the operator annihilating $\int_0^x \exp(-t^2) dt$
+admits also a solution which behaves for $x\to\infty$ like $\frac1x\exp(-x^2)$.
 
-(... a few introductory words ... available for shift and q-shift algebras ...)
+The next example illustrates the methods for finding rational and polynomial
+solutions of an operator~$L$.  These methods accept as an optional parameter an
+inhomogeneous part consisting of a list (or tuple) of base ring elements,
+$(f_1,\dots,f_r)$. They return as output a list of tuples $(g,c_1,\dots,c_r)$
+with $L(g)=c_1f_1+\cdots+c_rf_r$ where $g$ is a polynomial or rational function
+and $c_1,\dots,c_r$ are constants, i.e., elements of the base ring's base
+ring. The tuples form a basis of the solution space.
 
-Useful methods available for recurrence operators:
+In the example session below, we start from two polynomials $p,q$, then compute an 
+operator $L$ having $p$ and $q$ as solutions, and then recover $p$ and $q$ from~$L$.
+Note that for consistency also the solutions of homogeneous equations are returned as 
+tuples. At the end we give an example for solving an inhomogeneous equation.
+
+\begin{sageexample}
+  sage: R.<x> = ZZ[]
+  sage: p = x^2 + 3*x + 8; q = x^3 - 7*x + 5
+  sage: A.<Dx> = OreAlgebra(R)
+  sage: L = (p*Dx - p.derivative()).lclm(q*Dx - q.derivative())
+  sage: L
+  sage: L.polynomial_solutions()
+  sage: M = (2*x+3)*Dx^2 + (4*x+5)*Dx + (6*x+7)
+  sage: sol = M.polynomial_solutions([1,x,x^2,x^3])
+  sage: sol
+  sage: map(lambda s: M(s[0]) == s[1]+s[2]*x+s[3]*x^2+s[4]*x^3, sol)
+\end{sageexample}
+
+The functions \verb|polynomial_solutions| and \verb|rational_solutions| are not
+only defined for differential operators but also for recurrence operators, i.e.,
+elements of an Ore algebra of the form $R[x]\<Sx>$ or~$K(x)\<Sx>$. Some other
+methods defined for recurrence operators are listed in the following table.
 
@@
     \verb|to_T()| & converts to a differential operator in terms of the Euler derivative \\
-    \verb|annihilator_of_sum()| & converts an annihilator for $f(n)$ to one for $\sum_{k=0}^n f(k)$\\
-    \verb|annihilator_of_composition()| & converts an annihilator for $f(n)$ to one for $f(\lfloor un+v \rfloor)$ where $u, v \in \mathbb{Q}$ \\
+    \verb|annihilator_of_sum()| & converts an annihilator for $f(n)$ to one for the sum $\sum_{k=0}^n f(k)$\\
+    \verb|annihilator_of_composition()| & converts an annihilator for $f(n)$ to one for $f(\lfloor un+v \rfloor)$ where $u, v \in \set Q$ \\
     \verb|annihilator_of_interlacing()| & interlaces two or more sequences\\
-    \verb|generalized_series_solutions()| & computes asymptotic expansions of sequences annihilated by the operator\\ \hline
+    \verb|desingularize()| & computes a left multiple of this operator with polynomial coefficients and 
+           lowest possible leading coefficient degree\\
+    \verb|associate_solutions(p)| & applied to an operator~$P$, this computes, if possible, an operator $M$
+    and a rational function $m$ such that $(S-1)M=p+mP$ (see \cite{abramov99} for further information)\\
+    \verb|polynomial_solutions()| & computes the polynomial solutions of this operator\\
+    \verb|rational_solutions()| & computes the rational function solutions of this operator\\
+    \verb|generalized_series_solutions()| & computes asymptotic expansions of sequences annihilated by the operator\\ 
+    \verb|to_list()| & computes terms of a sequence annihilated by the operator\\\hline
   \end{tabular}
@@
 
-We compute an annihilator for the sequence
+As an example application, we compute an annihilator for the sequence
 $c(n) = \sum_{k=0}^n 1 / k!$:
@@
 \begin{sageexample}
-  sage: R.<n> = ZZ[]; RS.<Sn> = OreAlgebra(R)
+  sage: R.<n> = ZZ[]; A.<Sn> = OreAlgebra(R)
   sage: inverse_factorials = (n + 1) * Sn - 1
@@
 
-... add a figure with timings? ...
-
-\subsection{The $q$-Case}
-
-J and Q
-
-\section{Guessing}\label{sec:5}
-
-[[CITE RUBEY.]]
-
-Guessing is one of the most popular features of software packages like gfun.  It
-is a technique for detecting patterns in sequences of numbers, which is often
-used in combinatorics. The basic idea is simple. If we are given a finite array
-of numbers, though of as the first terms of an infinite sequence, we may want to
-know whether this sequence satisfies a recurrence. The algorithm behind a
-guessing engine searches for small equations matching the given data.
-Generically, no such equations exist, so if some are found, it is fair to
-``guess'' that they are in fact valid equations for the whole infinite sequence.
-=======
 \subsection{Guessing}
 
-Guessing is, in some sense, the reverse operation of \verb|to_list|. It is one
-of the most popular features of packages like gfun, and there are even some
-special purpose packages dedicated to this technique~\cite{kauers09a,hebisch11}.
-The basic idea is simple. If we are given a finite array of numbers, thought of
-as the first terms of an infinite sequence, we may want to know whether this
-sequence satisfies a recurrence. The algorithm behind a guessing engine searches
-for small equations matching the given data.  Generically, no such equations
-exist, so if some are found, it is fair to ``guess'' that they are in fact valid
-equations for the whole infinite sequence.
->>>>>>> manuel/master
+Guessing is, in some sense, the reverse operation of \verb|to_list| for
+recurrence operators, or \verb|power_series_solutions| for differential
+operators.  It is of the most popular features of packages like gfun, and there
+are even some special purpose packages dedicated to this
+technique~\cite{kauers09a,hebisch11}. The basic idea is simple. If we are given
+a finite array of numbers, thought of as the first terms of an infinite
+sequence, we may want to know whether this sequence satisfies a recurrence. The
+algorithm behind a guessing engine searches for small equations matching the
+given data.  Generically, no such equations exist, so if some are found, it is
+fair to ``guess'' that they are in fact valid equations for the whole infinite
+sequence.
 
@@
 The method tests a sequence of points $(r_1,d_1)$, $(r_2,d_2)$, \dots\ right
-below this second parabola. Success at a point $(r_i,d_i)$ means that some
+below this second hyperbola. Success at a point $(r_i,d_i)$ means that some
 evidence for an operator of order $\leq r_i$ and degree $\leq d_i$ has been found. 
@@
 
-\hangindent=-6.3cm\hangafter=-14\leavevmode
+\hangindent=-6.3cm\hangafter=-13\leavevmode
 \smash{\raise-5.3cm\rlap{\kern\hsize\kern-5.1cm\begin{tikzpicture}[scale=.45]
@@
 
-Finally, there is an option \verb|infolevel| which, when set to a positive integer, causes the method
-to print some output about the progress of the computation. Higher values lead to more detailed output.
-
-\section{Concluding Remarks}
-
-...
-
  \bibliographystyle{plain}

### 2013-06-16 doc/conf.py
minor changes in sphinx setup.

@@ -91,7 +91,7 @@ pygments_style = 'sphinx'
 
 # The theme to use for HTML and HTML Help pages.  Major themes that come with
 # Sphinx are currently 'default' and 'sphinxdoc'.
-html_theme = 'default'
+html_theme = 'pyramid'
 
 # Theme options are theme-specific and customize the look and feel of a theme
 # further.  For a list of options available for each theme, see the

### 2013-06-16 ore_operator.py
companion_matrix moved back to UnivariateOreOperator

@@ -1205,6 +1205,42 @@ class UnivariateOreOperator(OreOperator):
         from sage.matrix.constructor import matrix      
         return s.factorial(other.leading_coefficient(), n) * matrix(Alg.base_ring().fraction_field(), mat).det()
 
+    def companion_matrix(self):
+        r"""
+        Returns the companion matrix of ``self``.
+
+        If `r` is the order of ``self`` and `y` is annihilated by ``self``, then the companion matrix
+        as computed by this method is an `r\times r` matrix `M` such that 
+        `[\partial y,\partial^2 y,\dots,\partial^r y] = M [y,\partial y,\dots,\partial^{r-1}y]^T`.
+
+        In the shift case, if `c_i` is a sequence annihilated by ``self``, then also
+        `[c_{i+1}, c_{i+2}, \ldots, c_{i+r}]^T = M(i) [c_i, c_{i+1}, \ldots, c_{i+r-1}]^T`
+
+        EXAMPLES::
+
+            sage: R.<n> = QQ['n']
+            sage: A.<Sn> = OreAlgebra(R, 'Sn')
+            sage: M = ((-n-4)*Sn**2 + (5+2*n)*Sn + (3+3*n)).companion_matrix()
+            sage: M
+            [                0                 1]
+            [(3*n + 3)/(n + 4) (2*n + 5)/(n + 4)]
+            sage: initial = Matrix([[1],[1]])
+            sage: [prod(M(k) for k in range(n, -1, -1)) * initial for n in range(10)]
+            [
+            [1]  [2]  [4]  [ 9]  [21]  [ 51]  [127]  [323]  [ 835]  [2188]
+            [2], [4], [9], [21], [51], [127], [323], [835], [2188], [5798]
+            ]
+
+        """
+        from sage.matrix.constructor import Matrix
+        ring = self.base_ring().fraction_field()
+        r = self.order()
+        M = Matrix(ring, r, r)
+        for i in range(r-1):
+            M[i, i+1] = 1
+        for j in range(r):
+            M[r - 1, j] = self[j] / (-self[r])
+        return M
 
     def symmetric_product(self, other, solver=None):
         """

### 2013-06-16 ore_operator_1_1.py

@@ -1567,38 +1567,6 @@ class UnivariateRecurrenceOperatorOverUnivariateRing(UnivariateOreOperatorOverUn
         """
         return _rec2list(self, init, n, start, append, padd, ZZ)
 
-    def companion_matrix(self):
-        r"""
-        If ``self`` is an operator of order `r`, returns an `r` by `r` matrix
-        `M` such that for any sequence `c_i` annihilated by ``self``,
-        `[c_{i+1}, c_{i+2}, \ldots, c_{i+r}]^T = M(i) [c_i, c_{i+1}, \ldots, c_{i+r-1}]^T`
-
-        EXAMPLES::
-
-            sage: R.<n> = QQ['n']
-            sage: A.<Sn> = OreAlgebra(R, 'Sn')
-            sage: M = ((-n-4)*Sn**2 + (5+2*n)*Sn + (3+3*n)).companion_matrix()
-            sage: M
-            [                0                 1]
-            [(3*n + 3)/(n + 4) (2*n + 5)/(n + 4)]
-            sage: initial = Matrix([[1],[1]])
-            sage: [prod(M(k) for k in range(n, -1, -1)) * initial for n in range(10)]
-            [
-            [1]  [2]  [4]  [ 9]  [21]  [ 51]  [127]  [323]  [ 835]  [2188]
-            [2], [4], [9], [21], [51], [127], [323], [835], [2188], [5798]
-            ]
-
-        """
-        from sage.matrix.constructor import Matrix
-        ring = self.base_ring().fraction_field()
-        r = self.order()
-        M = Matrix(ring, r, r)
-        for i in range(r-1):
-            M[i, i+1] = 1
-        for j in range(r):
-            M[r - 1, j] = self[j] / (-self[r])
-        return M
-
     def forward_matrix_bsplit(self, n, start=0):
         r"""
         Uses division-free binary splitting to compute a product of ``n``

### 2013-06-16 paper/main.tex
paper: minor changes

@@
 and has not yet been incorporated into the official Sage distribution. Readers
-who want to try it out are invited to download the current latest version from 
+who want to try it out are invited to download the current version from 
 \begin{center}
-  URL
+  \verb|http://www.risc.jku.at/research/combinat/software/ore_algebra|
 \end{center}
-and are encouraged to send us bug reports and other comments. We hope 
+and are encouraged to send us bug reports or other comments. We hope 
 that the community will find the code useful.
 
-Creating an Ore algebra object is easy. The following instructions show how to 
-load the code and then create an Ore algebra $A$ of linear differential operators
-and an Ore algebra $B$ of recurrence operators. Observe the correct application
-of the respective commutation rules in both cases.
+The following instructions show how to load the code and then create an Ore
+algebra~$A$ of linear differential operators and an Ore algebra~$B$ of
+recurrence operators. Observe the correct application of the respective
+commutation rules in both cases.
 
@@
 The package also supports Ore algebras with several generators already. However,
-so far we offer no functionality beyond addition and multiplication for
-these. Much more functionality is available for algebras with one generator. We
-describe some of it in Section~\ref{sec:4}.
+so far we offer hardly more functionality than addition and multiplication for
+these. Much more functionality is available for algebras with one generator. 
+Some of it is described in Section~\ref{sec:4}. We plan to add more code for the
+multivariate case in the future.
 
@@
 the standard derivation $d/dt$ acting on the generator $t$ of the base ring, 
-a generator called $Sn$ represents the standard shift operator sending the
+a generator called $\mathrm{Sn}$ represents the standard shift operator sending the
 generator $n$ of the base ring to $n+1$.
@@
 of maps $\sigma,\delta$. Different generators commute with each other; 
-noncommutativity only takes place ``between'' generators and base ring elements.
+noncommutativity only takes place between generators and base ring elements.
 
@@
 mapped to themselves, so that \verb|{}| in the definition of $A$ in the example
-above is equivalent to \verb|{x:x,y:y,z:z}|.  Likewise, in the dictionaries
+above is equivalent to \verb|{x:x,y:y,z:z}|. In the dictionaries
 specifying~$\delta$, omitted generators are understood to be mapped to zero.
@@
 that behaves like before, and in addition an $Sx$ which acts like the standard
-shift on~$x$.
+shift on~$x$ and leaves the other generators fixed.
 
@@
   sage: D*x, Sx*x
+  sage: D*y, Sx*y
+  sage: D*z, Sx*z
 \end{sageexample}
@@
 Some of these methods are accessible in any Ore algebra while others are tied
-specifically to e.g.\ recurrence operators or differential operators.
+specifically to, e.g., recurrence operators or differential operators.
 
-In this section, we give an overview of some of the functionality provided by
+In this section, we give an overview of the functionality provided by
 the package. Because of space limitation, only some of the available methods
@@
     \verb|A.xlclm(B)| & returns the least common left multiple $L$ of $A$ and $B$ and the left quotients of $L$ and $A$ and of $L$ and $B$.\\
-    \verb|A.resultant(B)| & returns the resultant of $A$ and $B$ (see \cite{li96} for further information).\\
+    \verb|A.resultant(B)| & returns the resultant of $A$ and $B$ (see \cite{li96} for its definition and
+    properties).\\
     \hline
@@
 multiplies a given operator from the left by some element from the base ring
-such as to produce a canonical representative of the set of all the operators
-that can be obtained from the given one by left multiplication of a base ring
+such as to produce a canonical representative from the class of all the operators
+that can be obtained from each other by left multiplication of a base ring
 element. This facilitates the comparison of output.
 
-There are a number of options to adjust the functionality of the methods. For
-example, the efficiency of computing the GCRD depends on the size of the
-coefficients of intermediate results and there are different strategies to
-control this in so called polynomial remainder sequences (PRS). The default is
-the improved PRS described in~\cite{jaroschek13b}. which will usually be the
-fastest choice.
+The efficiency of computing the GCRD depends on the size of the
+coefficients of intermediate results, and there are different strategies to
+control this growth via so called polynomial remainder sequences (PRS). The default is
+the improved PRS described in~\cite{jaroschek13b}, which will usually be the
+fastest choice. Other strategies can be selected by the option \verb|prs|.
 
@@
 common solutions of $L_1$ and~$L_2$. The LCLM, on the other hand, is the minimal
-order operator whose solution space contains all the solutions of $L_1$ as well
-as all the solutions of~$L_1$. Because of this property, GCRD and LCLM are
+order operator whose solution space contains all the solutions of $L_1$ and
+all the solutions of~$L_1$. Because of this property, GCRD and LCLM are
 useful tools for constructing operators with prescribed solutions. For example,
 here is how to construct a differential operator which has the solutions $x^5$
-and $\exp(x)$ starting from the obvious operators annihilating $x^5$
+and $\exp(x)$, starting from the obvious operators annihilating $x^5$
 and~$\exp(x)$, respectively.
@@
 belonging to some Ore algebra~$A$ then this algebra contains also an operator
-which annihilates their sum~$f+g$. Thus the class of solutions of operators of
+which annihilates their sum~$f+g$. In other words, the class of solutions of operators of
 $A$ is closed under addition. It turns out that similar closure properties hold
 for other operations. The following table lists some of the corresponding
-methods. Methods for some further closure properties will appear further below.
+methods. Methods for more special closure properties will appear further below.
 
@@
 \]
-The idea is to derive, using commands
-from the table above, a recurrence satisfied by the left hand side and then show
-that this recurrence is also valid for the right hand side.
+The idea is to derive, using commands from the table above, a recurrence
+satisfied by the left hand side, and then show that this recurrence is also
+valid for the right hand side.
 
@@
 
-For the elements of some of the most important algebras, additional methods are
-available. The following table lists some of the additional methods available
-for differential operators, i.e., elements of an Ore algebra of the
+For the elements of some of the most important algebras, additional methods have
+been implemented. The following table lists some of the additional methods
+available for differential operators, i.e., elements of an Ore algebra of the
 form~$R[x]\<Dx>$ or $K(x)\<Dx>$.
@@
 The last output implies that the operator annihilating $\int_0^x \exp(-t^2) dt$
-admits also a solution which behaves for $x\to\infty$ like $\frac1x\exp(-x^2)$.
+also admits a solution which behaves for $x\to\infty$ like $\frac1x\exp(-x^2)$.
 
@@
 and $c_1,\dots,c_r$ are constants, i.e., elements of the base ring's base
-ring. The tuples form a basis of the solution space.
+ring. The tuples form a vector space basis of the solution space.
 
 In the example session below, we start from two polynomials $p,q$, then compute an 
-operator $L$ having $p$ and $q$ as solutions, and then recover $p$ and $q$ from~$L$.
+operator~$L$ having $p$ and $q$ as solutions, and then recover $p$ and $q$ from~$L$.
 Note that for consistency also the solutions of homogeneous equations are returned as 
@@
 
-In some cases (for example when the base ring is $\mathbb{Z}$ or
-$\mathbb{Z}[x]$), isolated values of a sequence can be
+In some cases, for example when the base ring is $\mathbb{Z}$ or
+$\mathbb{Z}[x]$, isolated values of a sequence can be
 computed asymptotically faster for large $n$ than by listing all
@@
 The \verb|forward_matrix_bsplit| method, called with argument~$n$,
-returns a matrix $P$ and a denominator $Q$ such that $P / Q$ multiplied a
+returns a matrix $P$ and a polynomial $Q$ such that $P / Q$ multiplied by a
 column vector of initial values $c_0, c_1, \ldots$
@@
 Guessing is, in some sense, the reverse operation of \verb|to_list| for
-recurrence operators, or \verb|power_series_solutions| for differential
+recurrence operators, or of \verb|power_series_solutions| for differential
 operators.  It is of the most popular features of packages like gfun, and there
 are even some special purpose packages dedicated to this
-technique~\cite{kauers09a,hebisch11}. The basic idea is simple. If we are given
+technique~\cite{kauers09a,hebisch11}. The basic idea is simple. Given
 a finite array of numbers, thought of as the first terms of an infinite
-sequence, we may want to know whether this sequence satisfies a recurrence. The
+sequence, we want to know whether this sequence satisfies a recurrence. The
 algorithm behind a guessing engine searches for small equations matching the
@@
 
-A number of options are available for governing the search for relations.  In
+Several options are available for customizing the search for relations.  In
 order to explain them, we first need to give some details on the underlying
@@
 The figure on the right illustrates the typical situation for guessing problems
-that are not too small and not too artificial. The light gray region indicates
+that are not too small and not too artificial. The gray region indicates
 the area which is not accessible with the given amount of data. Only the points
@@
 more terms than needed. If the data array contains more terms, superfluous ones are ignored in the
-interested of better performance. We must always have $0\leq e\leq c\leq \infty$. The default setting
+interest of better performance. We must always have $0\leq e\leq c\leq \infty$. The default setting
 is $e=0$, $c=\infty$.

### 2013-06-18 ore_operator_1_1.py
markup fix

@@ -2878,7 +2878,7 @@ class UnivariateQDifferentialOperatorOverUnivariateRing(UnivariateOreOperatorOve
         return self*self.parent().gen()
 
     def power_series_solutions(self, n=5):
-        """
+        r"""
         Computes the first few terms of the power series solutions of this operator.
 
         The method raises an error if Sage does not know how to factor univariate polynomials
@@ -2892,7 +2892,7 @@ class UnivariateQDifferentialOperatorOverUnivariateRing(UnivariateOreOperatorOve
 
         OUTPUT:
 
-        A list of power series of the form `x^\alpha + ...` with pairwise distinct
+        A list of power series of the form `x^{\alpha} + ...` with pairwise distinct
         exponents `\alpha` and coefficients in the base ring's base ring's fraction field.
         All expansions are computed up to order `k` where `k` is obtained by adding the
         maximal `\alpha` to the maximum of `n` and the order of ``self``.         

### 2013-06-18 paper/main.tex
minor change in paper and one docstring.

@@
 
-Not every ring is suitable as base ring of an Ore algebra. Base rings must
-themselves be polynomial rings (univariate or multivariate), or fraction 
-fields of polynomial rings. Their base rings in turn may be either $\set Z$,
-$\set Q$, a prime field $GF(p)$, or a number field $\set Q(\alpha)$, or
---- recursively --- some ring which itself would be suitable as base ring
-of an Ore algebra.
+In theory, any integral domain can serve as base ring of an Ore algebra. Not so
+in our implementation. Here, base rings must themselves be polynomial rings
+(univariate or multivariate), or fraction fields of polynomial rings. Their base
+rings in turn may be either $\set Z$, $\set Q$, a prime field $GF(p)$, or a
+number field $\set Q(\alpha)$, or --- recursively --- some ring which itself
+would be suitable as base ring of an Ore algebra.
 

### 2013-06-18 ore_operator_1_1.py

@@ -1576,12 +1576,12 @@ class UnivariateRecurrenceOperatorOverUnivariateRing(UnivariateOreOperatorOverUn
         allows rapidly computing `c_n, \ldots, c_{n+r-1}` (or just `c_n`)
         without generating all the intermediate values.
 
-        INPUT::
+        INPUT:
 
         - ``n`` -- desired number of terms to move forward
         - ``start`` (optional) -- starting index. Defaults to zero.
 
-        OUTPUT::
+        OUTPUT:
 
         A pair `(M, Q)` where `M` is an `r` by `r` matrix and `Q`
         is a scalar, such that `M / Q` is the product of the companion

### 2013-06-18 paper/main.tex
version of gpl added to copyright notice, as suggested by Fredrik.
development version switched off.

THIS IS VERSION 0.1

@@
 
- \bibliographystyle{plain}
- \bibliography{main}
+% \bibliographystyle{plain}
+% \bibliography{main}
+
+\begin{thebibliography}{10}
+
+\bibitem{abramov03}
+Sergei~A. Abramov, Ha~Q. Le, and Ziming Li.
+\newblock {OreTools:} a computer algebra library for univariate ore polynomial
+  rings.
+\newblock Technical Report CS-2003-12, University of Waterloo, 2003.
+
+\bibitem{abramov99}
+Sergei~A. Abramov and Mark van Hoeij.
+\newblock Integration of solutions of linear functional equations.
+\newblock {\em Integral transforms and Special Functions}, 9:3--12, 1999.
+
+\bibitem{bronstein96}
+Manuel Bronstein and Marko Petkov{\v s}ek.
+\newblock An introduction to pseudo-linear algebra.
+\newblock {\em Theoretical Computer Science}, 157(1):3--33, 1996.
+
+\bibitem{chyzak98a}
+Fr{\'e}d{\'e}ric Chyzak.
+\newblock {\em Fonctions holonomes en calcul formel}.
+\newblock PhD thesis, INRIA Rocquencourt, 1998.
+
+\bibitem{chyzak98}
+Fr{\'e}d{\'e}ric Chyzak and Bruno Salvy.
+\newblock Non-commutative elimination in {O}re algebras proves multivariate
+  identities.
+\newblock {\em Journal of Symbolic Computation}, 26:187--227, 1998.
+
+\bibitem{hebisch11}
+Waldemar Hebisch and Martin Rubey.
+\newblock Extended {R}ate, more {GFUN}.
+\newblock {\em Journal of Symbolic Computation}, 46(8):889--903, 2011.
+
+\bibitem{jaroschek13b}
+Maximilian Jaroschek.
+\newblock Improved polynomial remainder sequences for {Ore} polynomials.
+\newblock {\em Journal of Symbolic Computation}, 2013.
+\newblock to appear.
+
+\bibitem{jaroschek13a}
+Maximilian Jaroschek, Manuel Kauers, Shaoshi Chen, and Michael~F. Singer.
+\newblock Desingularization explains order-degree curves for {Ore} operators.
+\newblock In Manuel Kauers, editor, {\em Proceedings of ISSAC'13}, 2013.
+\newblock to appear.
+
+\bibitem{kauers09a}
+Manuel Kauers.
+\newblock Guessing handbook.
+\newblock Technical Report 09-07, RISC-Linz, 2009.
+
+\bibitem{kauers13}
+Manuel Kauers.
+\newblock The holonomic toolkit.
+\newblock In Johannes Bluemlein and Carsten Schneider, editors, {\em Computer
+  Algebra in Quantum Field Theory: Integration, Summation and Special
+  Functions}, Texts and Monographs in Symbolic Computation. Springer, 2013.
+
+\bibitem{koutschan09}
+Christoph Koutschan.
+\newblock {\em Advanced Applications of the Holonomic Systems Approach}.
+\newblock PhD thesis, RISC-Linz, Johannes Kepler Universit{\"a}t Linz, 2009.
+
+\bibitem{koutschan10c}
+Christoph Koutschan.
+\newblock {HolonomicFunctions (User's Guide)}.
+\newblock Technical Report 10-01, RISC Report Series, University of Linz,
+  Austria, January 2010.
+
+\bibitem{li96}
+Ziming Li.
+\newblock {\em A Subresultant Theory for Linear Differential, Linear
+  Difference, and {O}re Polynomials, with Applications}.
+\newblock PhD thesis, RISC-Linz, 1996.
+
+\bibitem{mallinger96}
+Christian Mallinger.
+\newblock Algorithmic manipulations and transformations of univariate holonomic
+  functions and sequences.
+\newblock Master's thesis, J. Kepler University, Linz, August 1996.
+
+\bibitem{ore33}
+{\O}ystein Ore.
+\newblock Theory of non-commutative polynomials.
+\newblock {\em Annals of Mathematics}, 34:480--508, 1933.
+
+\bibitem{salvy94}
+Bruno Salvy and Paul Zimmermann.
+\newblock {G}fun: a {M}aple package for the manipulation of generating and
+  holonomic functions in one variable.
+\newblock {\em ACM Transactions on Mathematical Software}, 20(2):163--177,
+  1994.
+
+\bibitem{sage}
+W.\thinspace{}A. Stein et~al.
+\newblock {\em {S}age {M}athematics {S}oftware (Version 5.9)}.
+\newblock The Sage Development Team, 2013.
+\newblock {\tt http://www.sagemath.org}.
+
+\end{thebibliography}
 

### 2013-06-18 generalized_series.py

@@ -11,6 +11,7 @@ generalized_series
 #                     Fredrik Johansson (fjohanss@risc.jku.at).             #
 #                                                                           #
 #  Distributed under the terms of the GNU General Public License (GPL)      #
+#  either version 2, or (at your option) any later version                  #
 #                                                                           #
 #  http://www.gnu.org/licenses/                                             #
 #############################################################################

### 2013-06-18 guessing.py

@@ -11,6 +11,7 @@ guessing
 #                     Fredrik Johansson (fjohanss@risc.jku.at).             #
 #                                                                           #
 #  Distributed under the terms of the GNU General Public License (GPL)      #
+#  either version 2, or (at your option) any later version                  #
 #                                                                           #
 #  http://www.gnu.org/licenses/                                             #
 #############################################################################

### 2013-06-18 ore_algebra.py

@@ -39,6 +39,7 @@ AUTHOR:
 #                     Fredrik Johansson (fjohanss@risc.jku.at).             #
 #                                                                           #
 #  Distributed under the terms of the GNU General Public License (GPL)      #
+#  either version 2, or (at your option) any later version                  #
 #                                                                           #
 #  http://www.gnu.org/licenses/                                             #
 #############################################################################
@@ -46,7 +47,8 @@ AUTHOR:
 
 ######### development mode ###########
 
-if True:
+if False:
+
     # let load("ore_algebra") trigger reload of the modules in the list below
     for mod in ['nullspace', 'ore_operator', 'ore_operator_1_1', 'ore_operator_mult', 'generalized_series']:
         try:
@@ -54,11 +56,27 @@ if True:
         except:
             pass
 
+    # if sage version is 5.10.beta3 or later, use taylor shift of flint2
+    import datetime
+    try:
+        v = version()
+    except:
+        v = "2013-01-01"    
+
+    if datetime.date(int(v[-10:-6]), int(v[-5:-3]), int(v[-2:])) >= datetime.date(2013, 05, 15):
+        try:
+            load("shift.spyx") # cython code defining 'taylor_shift_univ_int_poly' and 'taylor_shift_univ_modp_poly'
+        except:
+            pass
+
+#######################################
+
+
 def taylor_shift_univ_int_poly(p, i):
     ## assuming that p is an element of ZZ['x']
     return p(p.parent().gen() + i)
 def taylor_shift_univ_int_ratfun(q, i):
-    ## assuming that q is an element of Fract(ZZ['x'])
+    ## assuming that q is an element of Frac(ZZ['x'])
     num = taylor_shift_univ_int_poly(q.numerator(), i)
     den = taylor_shift_univ_int_poly(q.denominator(), i)
     return q.parent()(num, den, coerce=False, reduce=False)
@@ -70,21 +88,6 @@ def taylor_shift_univ_modp_ratfun(q, i):
     num = taylor_shift_univ_modp_poly(q.numerator(), i)
     den = taylor_shift_univ_modp_poly(q.denominator(), i)
     return q.parent()(num, den, coerce=False, reduce=False)
-
-# if sage version is 5.10.beta3 or later, use taylor shift of flint2
-import datetime
-try:
-    v = version()
-except:
-    v = "2013-01-01"    
-
[…]

### 2013-06-18 ore_operator.py

@@ -11,6 +11,7 @@ ore_operator
 #                     Fredrik Johansson (fjohanss@risc.jku.at).             #
 #                                                                           #
 #  Distributed under the terms of the GNU General Public License (GPL)      #
+#  either version 2, or (at your option) any later version                  #
 #                                                                           #
 #  http://www.gnu.org/licenses/                                             #
 #############################################################################

### 2013-06-18 ore_operator_1_1.py

@@ -14,6 +14,7 @@ one generator.
 #                     Fredrik Johansson (fjohanss@risc.jku.at).             #
 #                                                                           #
 #  Distributed under the terms of the GNU General Public License (GPL)      #
+#  either version 2, or (at your option) any later version                  #
 #                                                                           #
 #  http://www.gnu.org/licenses/                                             #
 #############################################################################

### 2013-06-18 ore_operator_mult.py

@@ -11,6 +11,7 @@ ore_operator_mult
 #                     Fredrik Johansson (fjohanss@risc.jku.at).             #
 #                                                                           #
 #  Distributed under the terms of the GNU General Public License (GPL)      #
+#  either version 2, or (at your option) any later version                  #
 #                                                                           #
 #  http://www.gnu.org/licenses/                                             #
 #############################################################################

### 2013-06-18 ore_algebra.py
development mode switched on again.

@@ -47,7 +47,7 @@ AUTHOR:
 
 ######### development mode ###########
 
-if False:
+if True:
 
     # let load("ore_algebra") trigger reload of the modules in the list below
     for mod in ['nullspace', 'ore_operator', 'ore_operator_1_1', 'ore_operator_mult', 'generalized_series']:

### 2014-02-27 ore_operator.py
renaming some variables in the PRS methods

@@ -1679,9 +1679,9 @@ def __improvedPRS__(r,additional):
     Computes one division step in the improved polynomial remainder sequence.
     """
 
-    d0 = r[0].order()
-    d1 = r[1].order()
-    orddiff = d0-d1
+    d1 = r[0].order()
+    d0 = r[1].order()
+    orddiff = d1-d0
 
     R = r[0].parent()
     Rbase = R.base_ring()
@@ -1693,7 +1693,7 @@ def __improvedPRS__(r,additional):
         beta = (-Rbase.one())**(orddiff+1)*sigma.factorial(sigma(phi,1),orddiff)
     else:
         (d2,oldalpha,k,essentialPart,phi) = (additional.pop(),additional.pop(),additional.pop(),additional.pop(),additional.pop())
-        phi = oldalpha / sigma.factorial(sigma(phi,1),d2-d1-1)
+        phi = oldalpha / sigma.factorial(sigma(phi,1),d2-d0-1)
         beta = oldalpha.parent()(((-Rbase.one())**(orddiff+1)*sigma.factorial(sigma(phi,1),orddiff)*k))
         essentialPart = sigma(essentialPart,-orddiff)
 
@@ -1715,9 +1715,9 @@ def __subresultantPRS__(r,additional):
     Computes one division step in the subresultant polynomial remainder sequence.
     """
 
-    d0 = r[0].order()
-    d1 = r[1].order()
-    orddiff = d0-d1
+    d1 = r[0].order()
+    d0 = r[1].order()
+    orddiff = d1-d0
 
     R = r[0].parent()
     Rbase = R.base_ring()
@@ -1728,7 +1728,7 @@ def __subresultantPRS__(r,additional):
         beta = (-Rbase.one())*sigma.factorial(sigma(phi,1),orddiff)
     else:
         (d2,phi) = (additional.pop(),additional.pop())
-        orddiff2 = d2-d0
+        orddiff2 = d2-d1
         phi = sigma.factorial(-r[0].leading_coefficient(),orddiff2) / sigma.factorial(sigma(phi,1),orddiff2-1)
         beta = (-Rbase.one())*sigma.factorial(sigma(phi,1),orddiff)*r[0].leading_coefficient()
 

### 2014-02-27 paper/main.tex

@@
-\documentclass[11pt]{amsart}
+%\documentclass[11pt]{amsart}
+\documentclass{llncs}
 \usepackage{amsmath}
@@
 
- \author[Manuel Kauers, Maximilian Jaroschek, Fredrik Johansson]
-   {Manuel Kauers\,$^\ast$, Maximilian Jaroschek\,$^\ast$, Fredrik Johansson\,$^\ast$}
- \address{Manuel Kauers, Research Institute for Symbolic Computation (RISC), J. Kepler University Linz, Austria}
- \email{mkauers@risc.uni-linz.ac.at}
- \address{Maximilian Jaroschek, Research Institute for Symbolic Computation (RISC), J. Kepler University Linz, Austria}
- \email{mjarosch@risc.uni-linz.ac.at}
- \address{Fredrik Johansson, Research Institute for Symbolic Computation (RISC), J. Kepler University Linz, Austria}
- \email{fjohanss@risc.uni-linz.ac.at}
- \thanks{$^\ast$ Supported by the Austrian FWF grant Y464-N18.}
+ \title{Ore Polynomials in Sage\thanks{All three authors were supported by the Austrian FWF grant Y464-N18.}}
+ \titlerunning{Ore Polynomials in Sage}
 
- \title{Ore Polynomials in Sage}
+ \author{Manuel Kauers\and Maximilian Jaroschek\and Fredrik Johansson}
+ \authorrunning{Manuel Kauers\and Maximilian Jaroschek\and Fredrik Johansson}
+
+\institute{Research Institute for Symbolic Computation (RISC)\\ 
+  Johannes Kepler University (JKU)\\
+  4040 Linz, Austria\\
+  \email{[mkauers|mjarosch|fjohanss]@risc.jku.at}}
+
+ \maketitle
 
@@
 We present a Sage implementation of Ore algebras. The main features for the most
-common instances include basic arithmetic and actions; gcrd and lclm; D-finite
+common instances include basic arithmetic and actions; GCRD and LCLM; D-finite
 closure properties; natural transformations between related algebras; guessing;
@@
  \end{abstract}
-
- \maketitle
 
@@
 computer algebra systems, for example in the Maple packages
-OreTools~\cite{abramov03}, gfun~\cite{salvy94} or mgfun~\cite{chyzak98a}, or in
+OreTools~\cite{abramov03}, gfun~\cite{salvy94} or Mgfun~\cite{chyzak98a}, or in
 the Mathematica packages by Mallinger~\cite{mallinger96} and
@@
 here is a way to define an algebra $A$ over $K$ with two generators, a $D$ 
-that behaves like before, and in addition an $Sx$ which acts like the standard
+that behaves like before, and in addition an $\mathrm{Sx}$ which acts like the standard
 shift on~$x$ and leaves the other generators fixed.
@@
 coefficients of intermediate results, and there are different strategies to
-control this growth via so called polynomial remainder sequences (PRS). The default is
+control this growth via so-called polynomial remainder sequences (PRS). The default is
 the improved PRS described in~\cite{jaroschek13b}, which will usually be the
@@
 order operator whose solution space contains all the solutions of $L_1$ and
-all the solutions of~$L_1$. Because of this property, GCRD and LCLM are
+all the solutions of~$L_2$. Because of this property, GCRD and LCLM are
 useful tools for constructing operators with prescribed solutions. For example,
@@
 recurrence operators, or of \verb|power_series_solutions| for differential
-operators.  It is of the most popular features of packages like gfun, and there
+operators.  It is one of the most popular features of packages like gfun, and there
 are even some special purpose packages dedicated to this
@@
 Sergei~A. Abramov, Ha~Q. Le, and Ziming Li.
-\newblock {OreTools:} a computer algebra library for univariate ore polynomial
+\newblock {OreTools:} a computer algebra library for univariate Ore polynomial
   rings.
@@
 \newblock Improved polynomial remainder sequences for {Ore} polynomials.
-\newblock {\em Journal of Symbolic Computation}, 2013.
-\newblock to appear.
+\newblock {\em Journal of Symbolic Computation}, 58:64--76, 2013.
 
@@
 \newblock Desingularization explains order-degree curves for {Ore} operators.
-\newblock In Manuel Kauers, editor, {\em Proceedings of ISSAC'13}, 2013.
-\newblock to appear.
+\newblock In Manuel Kauers, editor, {\em Proceedings of ISSAC'13}, pp. 157--164, 2013.
 
@@
 \newblock The holonomic toolkit.
-\newblock In Johannes Bluemlein and Carsten Schneider, editors, {\em Computer
+\newblock In Johannes Bl\"umlein and Carsten Schneider, editors, {\em Computer
   Algebra in Quantum Field Theory: Integration, Summation and Special

### 2019-04-25 papers/issac19demo/main.tex
software demo abstract

\documentclass[11pt]{article}
\usepackage{sigsam, amsmath, sagetex, tikz}

% PAGE LIMIT: FOUR PAGES!

% leave as is
\issue{Vol.xx, No.xx, Issue xxx, month 201x}
\articlehead{ISSAC 2019 abstracts}
\titlehead{Multivariate Ore Polynomials in Sage}
\authorhead{Manuel Kauers, Marc Mezzarobba}
%\titlehead{Title of your paper}
%\authorhead{Author's Name}
\setcounter{page}{1}

\def\<#1>{\langle#1\rangle}
\let\Bold\mathbb
\let\set\mathbb

\begin{document}

\title{Multivariate Ore Polynomials in Sage}

\author{%
  Manuel Kauers\footnote{Supported by the Austrian FWF grants P31571-N32, F5004} \
  and Marc Mezzarobba\footnote{??}\\[\medskipamount]
  Institute for Algebra $\cdot$ Johannes Kepler University $\cdot$ Linz, Austria\\
  CNRS, LIB6 $\cdot$ Sorbonne Universit\'e $\cdot$ Paris, France\\[\medskipamount]
  \url{manuel@kauers.de} $\cdot$ \url{marc@mezzarobba.net}
}

\date{}

\maketitle

\begin{abstract}
  We present the latest update of the ore\_algebra package for Sage.
  The main new feature in this release is the support of operators in several variables.
\end{abstract}

\section{Introduction}

Ore polynomials are important features of computer algebra systems because they provide
functionality for working with differential and recurrence operators that describe
so-called D-finite functions and sequences~\cite{stanley80,kauers10j,kauers13}. Maple has
the OreTools package~\cite{abramov03} and gfun~\cite{salvy94}, for Mathematica we have
the package of Koutschan~\cite{koutschan10c},
and there is the ore\_algebra package by Kauers, Jaroschek and Johansson~\cite{kauers14b} for
Sage~\cite{zimmermann18}. This software demo abstract is about the most recent version of the
latter package.

Earlier versions of the ore\_algebra package already provided functionality for basic arithmetic
and actions; GCRD and LCLM; D-finite closure properties; natural transformations between related
algebras; guessing; desingularization; solvers for polynomials, rational functions and (generalized)
power series. Mezzarobba~\cite{mezzarobba16} added a subpackage ore\_algebra.analytic for arbitrary-precision
evaluation of D-finite functions, which is now fully integrated into the main version of the package.
Hofstadler~\cite{hofstadler19} contributed a set of convenience functions which support a better integration
with symbolic expressions.

Almost all the features so far were limited to the univariate case. The main contribution in the
latest version is an extension to the multivariate case. While basic arithmetic of Ore polynomials
in several variables has been supported since the beginning, we now also offer implementations of
D-finite closure properties and creative telescoping as well as a (rudimentary) implementation of
Gr\"obner bases.

\section{Basic Examples}

After loading the package and creating the base ring $\set Z[x,y]$, the Ore algebra of differential
operators for $x$ and $y$ can be created as follows. 
\begin{sageexample}
  sage: from ore_algebra import *
  sage: R.<x,y> = PolynomialRing(ZZ)
  sage: A.<Dx,Dy> = OreAlgebra(R)
  sage: A
\end{sageexample}
The usual arithmetic respects the required commutation rules. 
\begin{sageexample}
  sage: (Dx + Dy)*(y + x)
  sage: (x + y)*(Dx + Dy)
\end{sageexample}
Differential operators can be applied to objects which Sage knows to differentiate and to multiply
by base ring elements:
\begin{sageexample}
  sage: (Dx + Dy)(sin(x*y))
\end{sageexample}
Using $Sx,Sy$ instead of $Dx,Dy$ creates an Ore algebra with shift operators instead of differential operators.
Of course, also combinations like $\set Z[x,y]\<Sx,Dy>$ and other generators besides shift and derivation are possible.
The available operators are the same as in the univariate case, see~\cite{kauers14b} for more details.

In the univariate case, methods for executing closure properties were directly attached to the annihilating operators.
In the multivariate case, where several annihilating operators are needed to describe a D-finite object, we must work
with left ideals of Ore algebras. They can be constructed as follows.
\begin{sageexample}
  sage: I = A.ideal([Dy^2, Dx*Dy, Dx^2 - y*Dy + 1])
  sage: I
\end{sageexample}
Note that the base ring is always converted to a field.

Various ideal theoretic operations are available as methods. 
\begin{sageexample}
  sage: I.eliminate([Dy])
  sage: Dx^3 + Dx in I # (membership)
  sage: J = A.ideal([-Dy^2 + x*Dx - 1, Dx*Dy, Dx^2])
  sage: I == J
  sage: K = I.intersection(J) # closure property `addition`
  sage: K <= J # (inclusion)
  sage: K = I.symmetric_product(I) # closure property `multiplication`
  sage: K.vector_space_basis() # terms under the staircase
  sage: I.annihilator_of_associate(Dx + Dy) # closure property `ore action`
  sage: I.annihilator_of_composition(x=x^2,y=1-x^2) # closure property `composition`
\end{sageexample}
Most of these operations rely on a generic FGLM implementation which uses the packages own linear system solvers.
There is also a direct implementation of Buchberger's algorithm for computing Gr\"obner bases, which however is
not highly optimized. 
\begin{sageexample}
  sage: I.groebner_basis()
\end{sageexample}
All operations discussed so far are generalizations of the univariate counterparts that already existed
in earlier versions. A feature that is inherently multivariate is creative telescoping~\cite{zeilberger91,chyzak00,chyzak14}.
Here, the input is a left ideal $I\subseteq A$, say $A=\set Q(x,y)\<Dx,Dy>$ and the task is to find
the ideal consisting of all operators $P\in\set Q(x)\<Dx>$ such that there exists $Q\in A$ with $P-D_yQ\in I$.
The operators $P$ are called \emph{telescopers,} and the corresponding operators $Q$ are called their \emph{certificates.}
Creative telescoping is a central operation for evaluating definite sums and integrals~\cite{chyzak14}.
We cannot explain here in detail the relevance of creative telescoping for these applications, but the use of the
package is easily demonstrated:
\begin{sageexample}
  sage: A.ideal([Dx - 2*x*y^2, Dy - 2*x^2*y]).ct(Dy)
\end{sageexample}
The output is a pair $([P_1,P_2,\dots],[Q_1,Q_2,\dots])$ whose first component is a basis of the ideal of telescopers
and whose second component contains the certificates corresponding to the basis elements.
The example above originates from the integration problem $\int_y\exp(x^2y^2)dy$. The following example is the essence
of a proof of the binomial theorem $\sum_k\binom nk=2^n$:
\begin{sageexample}
  sage: B.<Sx,Sy> = OreAlgebra(R)
  sage: B.ideal([(y+1)*Sy + (y-x), (x-y+1)*Sx - (x+1)]).ct(Sy - 1)
\end{sageexample}
Finally, also the guessing features have been extended to the multivariate setting. For example, annihilating operators
for the binomial coefficients can be found as follows.
\begin{sageexample}
  sage: data = [[binomial(n,k) for k in range(20)] for n in range(20)]
  sage: guess(data, B, point_filter = lambda n,k : k<=n).groebner_basis()
\end{sageexample}
The option \verb|point_filter| informs the guessing engine which part of the input array contains interesting data.
A number of further options is specified in the documentation.

\section{Some Slightly Larger Examples}

The examples shown above need virtually no computation time. In order to get some idea about the performance of our
implementation, we consider a set of 19 creative telescoping problems that appeared in a study of restricted lattice
walks~\cite{bostan16b}.
Each of the 19 problems starts from a certain rational function $r\in\set Q(u,v,t)$ and the task consists of two
applications of creative telescoping. Writing $I$ for the ideal of annihilating operators of $r$ in $\set Q(u,v,t)\<Du,Dv,Dt>$,
we first have to compute the ideal $T_v$ in $\set Q(u,t)\<Du,Dt>$ of all telescopers of $I$ with respect to~$Dv$ and then
the ideal $T_{u,v}$ in $\set Q(t)\<Dt>$ of telescopers of $T_v$ with respect to~$Du$. The smallest rational function in
the collection is
\[
\tfrac{-u^2 v^2+u^2+v^2-1}{t u^3 v^2-t u^3 v+t u^2 v^3-2 t u^2 v^2+2 t u^2 v-t u^2-t u v^3+2 t u v^2-2 t
  u v+t u-t v^2+t v-u^2 v^2+u^2 v+u v^2-u v},
\]
and the largest is about twice as long. The complete data is available in ore\_algebra.examples.smallsteps.

In the table below, we compare our performance to that of Koutschan's package~\cite{koutschan10c}. We do not claim that a
performance comparison will look similarly for any creative telescoping problem, but we think the experiment is at least
an indication that our implementation is not too bad.

\begin{center}
  % timings taken on picard.algebra.uni-linz.ac.at
  % with Sage 8.7 and ore_algebra version 7ab0f7e821cde38e1ace6ae0844bc34e726aaf5a
  % with Mathematica 12 and HolonomicFunctions.m version 1.7.3
  % on 2019-04-25
  \begin{tabular}{c|ccccccccccccccccccc}
    case & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9  \\\hline
    ore\_algebra & 601ms & 668ms & 705ms & 862ms & 4.82s & 10.3s & 5.22s & 10.7s & 11.1s \\
    HolonomicFunctions.m & 1.55s & 1.59s & 1.86s & 2.09s & 25.2s & 51.9s & 26.8s & 55.1s & 51.5s
  \end{tabular}

  \begin{tabular}{c|ccccccccccccccccccc}
    case & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 \\\hline
    ore\_algebra & 11.9s & 4.46s & 8.57s & 11.1s & 11.7s & 4.21s & 8.27s & 18.9s & 36.2s & 2.05s \\
    HolonomicFunctions.m & 63.4s & 19.4s & 35.4s & 51.3s & 62.3s & 19.1s & 34.7s & 198s & 445s & 5.77s
  \end{tabular}
\end{center}

\small
\bibliographystyle{plain}
\bibliography{bib}

\end{document}

### 2019-04-25 src/ore_algebra/examples/smallsteps.py

@@ -5,15 +5,16 @@ of walks with small steps in the quarter plane" by Bostan, Chyzak, van Hoeij,
 Kauers, and Pech (2017).
 
 For the rational functions rat[1], ..., rat[19], the task consists in computing
-a telescoper wrt Du and Dv. The variables x and y may be set to 1. 
+a telescoper wrt Du and Dv. Interesting settings are when x and y are both set
+to 1 or one of them is set to 0 and the other is left symbolic. 
 
 Example: 
 
    sage: from ore_algebra.examples.smallsteps import rat
    sage: from ore_algebra import OreAlgebra
-   sage: A.<Du,Dv,Dt> = OreAlgebra(rat[1](x=1,y=1).parent(), 'Du', 'Dv', 'Dt')
+   sage: A.<Du,Dv,Dt> = OreAlgebra(ZZ['u','v','t'].fraction_field(), 'Du', 'Dv', 'Dt')
    
-   sage: q = rat[1](x=1,y=1)
+   sage: q = A.base_ring()(rat[1](x=1,y=1))
    sage: ct1 = A.ideal([q*D - D(q) for D in Du,Dv,Dt]).ct(Dv)[0]
    sage: ct2 = ct1[0].parent().ideal(ct1).ct(Du)[0]
 

### 2019-05-09 papers/issac19demo/main.tex
link to github
small changes to the software demo abstract
* Sage → SageMath
* official affiliation and funding sources for MM
* formatting changes
* explicitly refer to v0.4
* installation instructions
* reword a couple of sentences, mention some other improvements in 0.4,
  work that went directly into Sage
* smallsteps → ssw (cf. previous commit)
* thank reviewers of Sage patches

@@
 \documentclass[11pt]{article}
-\usepackage{sigsam, amsmath, sagetex, tikz}
+\usepackage[fleqn]{amsmath}
+\usepackage{sigsam, sagetex, tikz}
+\usepackage{hyperref}
+\usepackage{booktabs}
 
@@
 \articlehead{ISSAC 2019 abstracts}
-\titlehead{Multivariate Ore Polynomials in Sage}
+\titlehead{Multivariate Ore Polynomials in SageMath}
 \authorhead{Manuel Kauers, Marc Mezzarobba}
@@
 
+% \setkeys{FV}{vspace=0pt}
+\makeatletter
+\let\FV@ListVSpace\relax
+\makeatother
+
+\date{}
+
 \begin{document}
 
-\title{Multivariate Ore Polynomials in Sage}
+\title{Multivariate Ore Polynomials in SageMath}
 
@@
   Manuel Kauers\footnote{Supported by the Austrian FWF grants P31571-N32, F5004} \
-  and Marc Mezzarobba\footnote{??}\\[\medskipamount]
+  and Marc Mezzarobba%
+  \footnote{Supported in part by ANR grant ANR-14-CE25-0018-01 (FastRelax).}%
+  \\[\medskipamount]
   Institute for Algebra $\cdot$ Johannes Kepler University $\cdot$ Linz, Austria\\
-  CNRS, LIB6 $\cdot$ Sorbonne Universit\'e $\cdot$ Paris, France\\[\medskipamount]
-  \url{manuel@kauers.de} $\cdot$ \url{marc@mezzarobba.net}
+  Sorbonne Université, CNRS, Laboratoire d'informatique de Paris~6, LIP6,
+  F-75005 Paris, France\\[\medskipamount]
+  \href{mailto:manuel@kauers.de}{\nolinkurl{manuel@kauers.de}} $\cdot$
+  \href{mailto:marc@mezzarobba.net}{\nolinkurl{marc@mezzarobba.net}}
 }
-
-\date{}
 
@@
 \begin{abstract}
-  We present the latest update of the ore\_algebra package for Sage.
+  We present the latest update of the ore\_algebra package for SageMath.
   The main new feature in this release is the support of operators in several variables.
@@
 Ore polynomials are important features of computer algebra systems because they provide
-functionality for working with differential and recurrence operators that describe
-so-called D-finite functions and sequences~\cite{stanley80,kauers10j,kauers13}. Maple has
-the OreTools package~\cite{abramov03} and gfun~\cite{salvy94}, for Mathematica we have
-the package of Koutschan~\cite{koutschan10c},
-and there is the ore\_algebra package by Kauers, Jaroschek and Johansson~\cite{kauers14b} for
-Sage~\cite{zimmermann18}. This software demo abstract is about the most recent version of the
-latter package.
+functionality for manipulating differential and recurrence operators that describe
+so-called D-finite functions and sequences~\cite{stanley80,kauers10j,kauers13}.
+Packages for working with Ore polynomials include, among others,
+OreTools~\cite{abramov03} and gfun~\cite{salvy94} for Maple,
+the HolonomicFunctions package of Koutschan~\cite{koutschan10c} for Mathematica,
+and ore\_algebra package by Kauers, Jaroschek and Johansson~\cite{kauers14b} for
+SageMath (Sage)~\cite{zimmermann18}. This software demo abstract is about the most recent version of the
+latter package, which works with Sage~8.7.
 
@@
 algebras; guessing; desingularization; solvers for polynomials, rational functions and (generalized)
-power series. Mezzarobba~\cite{mezzarobba16} added a subpackage ore\_algebra.analytic for arbitrary-precision
-evaluation of D-finite functions, which is now fully integrated into the main version of the package.
+power series. Mezzarobba~\cite{mezzarobba16} added a subpackage ore\_algebra.analytic for symbolic-numeric
+computations with D-finite analytic functions, which is now fully integrated into the main version of the package.
 Hofstadler~\cite{hofstadler19} contributed a set of convenience functions which support a better integration
@@
 
-Almost all the features so far were limited to the univariate case. The main contribution in the
-latest version is an extension to the multivariate case. While basic arithmetic of Ore polynomials
-in several variables has been supported since the beginning, we now also offer implementations of
-D-finite closure properties and creative telescoping as well as a (rudimentary) implementation of
-Gr\"obner bases.
+While basic arithmetic of Ore polynomials in several variables has been supported since the
+beginning, everything else so far was limited to the univariate case.
+The main new feature in versions~0.3--0.4 of the package is an extension of the core
+functionality to multivariate operators, with efficient support of D-finite closure properties and
+creative telescoping as well as a rudimentary implementation of Gr\"obner bases.
+Other novelties include substantially improved numerical evaluation code compared to the status
+described in~\cite{mezzarobba16}, and preliminary support for Python~3.
 
 \section{Basic Examples}
+
+With a typical Sage setup, ore\_algebra~v0.4 can be installed using the shell command
+\begin{verbatim}
+  $ sage -pip install git+https://github.com/mkauers/ore_algebra.git@0.4
+\end{verbatim}
+(Omit the \verb!@0.4! suffix for the latest development version.)
 
@@
 Using $Sx,Sy$ instead of $Dx,Dy$ creates an Ore algebra with shift operators instead of differential operators.
-Of course, also combinations like $\set Z[x,y]\<Sx,Dy>$ and other generators besides shift and derivation are possible.
+Of course, combinations like $\set Z[x,y]\<Sx,Dy>$ and other generators besides shift and derivation are also possible.
 The available operators are the same as in the univariate case, see~\cite{kauers14b} for more details.
@@
 In the multivariate case, where several annihilating operators are needed to describe a D-finite object, we must work
-with left ideals of Ore algebras. They can be constructed as follows.
+with left ideals of Ore algebras. Ideals can be constructed as follows.
 \begin{sageexample}
@@
 \end{sageexample}
+
 All operations discussed so far are generalizations of the univariate counterparts that already existed
@@
 \end{sageexample}
+
 Finally, also the guessing features have been extended to the multivariate setting. For example, annihilating operators
@@
 \[
-\tfrac{-u^2 v^2+u^2+v^2-1}{t u^3 v^2-t u^3 v+t u^2 v^3-2 t u^2 v^2+2 t u^2 v-t u^2-t u v^3+2 t u v^2-2 t
+\frac{{-}u^2 v^2{+}u^2{+}v^2{-}1}{t u^3 v^2{-}t u^3 v{+}t u^2 v^3{-}2 t u^2 v^2{+}2 t u^2 v{-}t u^2{-}t u v^3{+}2 t u v^2{-}2 t
   u v+t u-t v^2+t v-u^2 v^2+u^2 v+u v^2-u v},
 \]
-and the largest is about twice as long. The complete data is available in ore\_algebra.examples.smallsteps.
+and the largest is about twice as long. The complete data is available in ore\_algebra.examples.ssw.
 
@@
 an indication that our implementation is not too bad.
+Note in passing that reaching an acceptable performance level required a number of improvements to
+the implementation of multivariate polynomials and rational functions in Sage itself.
+Although ore\_algebra is developed as an external Sage package, a significant part of the
+development effort reported here will thus directly benefit other Sage users.
 
@@
   % on 2019-04-25
-  \begin{tabular}{c|ccccccccccccccccccc}
-    case & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9  \\\hline
-    ore\_algebra & 601ms & 668ms & 705ms & 862ms & 4.82s & 10.3s & 5.22s & 10.7s & 11.1s \\
-    HolonomicFunctions.m & 1.55s & 1.59s & 1.86s & 2.09s & 25.2s & 51.9s & 26.8s & 55.1s & 51.5s
-  \end{tabular}
-
-  \begin{tabular}{c|ccccccccccccccccccc}
-    case & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 \\\hline
-    ore\_algebra & 11.9s & 4.46s & 8.57s & 11.1s & 11.7s & 4.21s & 8.27s & 18.9s & 36.2s & 2.05s \\
-    HolonomicFunctions.m & 63.4s & 19.4s & 35.4s & 51.3s & 62.3s & 19.1s & 34.7s & 198s & 445s & 5.77s
+  \begin{tabular}{rccccccccccccccccccc}
+    \toprule
+    case & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 \\\midrule
+    ore\_algebra & 601ms & 668ms & 705ms & 862ms & 4.82s & 10.3s & 5.22s & 10.7s & 11.1s & 11.9s \\
+    HolonomicFunctions.m & 1.55s & 1.59s & 1.86s & 2.09s & 25.2s & 51.9s & 26.8s & 55.1s & 51.5s &
+    63.4s \\
+    \midrule
+    case & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 \\\midrule
+    ore\_algebra & 4.46s & 8.57s & 11.1s & 11.7s & 4.21s & 8.27s & 18.9s & 36.2s & 2.05s \\
+    HolonomicFunctions.m & 19.4s & 35.4s & 51.3s & 62.3s & 19.1s & 34.7s & 198s & 445s & 5.77s \\
+    \bottomrule
   \end{tabular}
 \end{center}
+
+\paragraph{Acknowledgements.}
+
+We thank Bruno Grenet, Ben Hutz, Julian Rüth, and Travis Scrimshaw for reviewing the Sage patches
+mentioned above.
 

### 2019-05-09 src/ore_algebra/examples/smallsteps.py
merge examples.smallsteps into examples.ssw

@@ -1,67 +1,0 @@
-
-r"""
-Examples from the paper "Hypergeometric expressions for generating functions
-of walks with small steps in the quarter plane" by Bostan, Chyzak, van Hoeij,
-Kauers, and Pech (2017).
-
-For the rational functions rat[1], ..., rat[19], the task consists in computing
-a telescoper wrt Du and Dv. Interesting settings are when x and y are both set
-to 1 or one of them is set to 0 and the other is left symbolic. 
-
-Example: 
-
-   sage: from ore_algebra.examples.smallsteps import rat
-   sage: from ore_algebra import OreAlgebra
-   sage: A.<Du,Dv,Dt> = OreAlgebra(ZZ['u','v','t'].fraction_field(), 'Du', 'Dv', 'Dt')
-   
-   sage: q = A.base_ring()(rat[1](x=1,y=1))
-   sage: ct1 = A.ideal([q*D - D(q) for D in Du,Dv,Dt]).ct(Dv)[0]
-   sage: ct2 = ct1[0].parent().ideal(ct1).ct(Du)[0]
-
-
-"""
-
-from sage.rings.all import PolynomialRing, ZZ
-
-Pols, (u, v, x, y, t) = PolynomialRing(ZZ, ('u', 'v', 'x', 'y', 't')).objgens()
-
-rat = {}
-
-# indexed from 1 on in order to be consistent with the paper
[…]

### 2019-05-09 src/ore_algebra/examples/ssw.py

@@ -2,19 +2,30 @@
 r"""
 Small-step walks
 
-Operators and connection problems related to work by Alin Bostan, Frédéric
-Chyzak, Mark van Hoeij, Manuel Kauers and Lucien Pech on the combinatorics of
-walks with small steps in the quarter plane. (A. Bostan and M. Kauers, Automatic
-Classification of Restricted Lattice Walks, FPSAC'09; A. Bostan, F. Chyzak,
-M. van Hoeij, M. Kauers and L. Pech, Hypergeometric expressions for generating
-functions of walks with small steps in the quarter plane, arXiv:1606.02982
-[math.CO], <URL: http://specfun.inria.fr/chyzak/ssw/index.html>)::
+Examples from the paper "Hypergeometric expressions for generating functions
+of walks with small steps in the quarter plane" by Alin Bostan, Frédéric
+Chyzak, Mark van Hoeij, Manuel Kauers and Lucien Pech (European Journal of
+Combinatorics, 2017; arXiv:1606.02982 [math.CO],
+<URL: http://specfun.inria.fr/chyzak/ssw/index.html>)::
 
-    sage: from ore_algebra import *
-    sage: Pols.<t> = QQ[]
-    sage: Dops.<Dt> = OreAlgebra(Pols)
+For the rational functions rat[1], ..., rat[19], the task consists in computing
+a telescoper wrt Du and Dv. Interesting settings are when x and y are both set
+to 1 or one of them is set to 0 and the other is left symbolic. For example::
+
[…]

### 2019-05-09 src/ore_algebra/guessing.py
better handling of data with lots of zeros in univariate guessing

@@ -22,7 +22,6 @@ TESTS::
 
 from __future__ import absolute_import, division, print_function
 
-
 ######### development mode ###########
 """
 try:
@@ -161,16 +160,48 @@ def guess(data, algebra, **kwargs):
 
     A = algebra; R = A.base_ring(); K = R.base_ring(); x = R.gen()
 
+    if A.ngens() > 1 or R.ngens() > 1:
+        return guess_mult(data, algebra, **kwargs)
+    
     if type(data) == str:
         with open(data, 'r') as f:
             data = [ K(line) for line in f ]
 
-    if A.ngens() > 1 or R.ngens() > 1:
-        return guess_mult(data, algebra, **kwargs)
-    
-    elif R.is_field():
-        return guess(data, A.change_ring(R.ring()), **kwargs)
+    if (data[0].is_zero() or data[1].is_zero()) and (A.is_C() or A.is_S()):
+        
+        if all( d.is_zero() for d in data ):
+            return A.one()
+        for i in range(len(data)):
+            if not data[i].is_zero():
+                a = i
+                break
+        m = None; b = ZZ(a)
+        for i in range(b + 1, len(data)):
+            if not data[i].is_zero():
+                m = i - b if m is None else m.gcd(i - b)
+                b = i
+                if m.is_one():
+                    break
 
+        if m is not None and m > 1 and m > a:
+            if 'infolevel' in kwargs and kwargs['infolevel'] >= 1:
+                print("Recognized that only 1 out of " + str(m) + " terms is nonzero; removing zeros...")
+            eq = guess([data[m*k + a] for k in range((len(data) - a)/m)], A, **kwargs)
+            if 'infolevel' in kwargs and kwargs['infolevel'] >= 1:
+                print("Adjusting equation to restore deleted zeros...")
+            x = R.gen()
+            if A.is_S():
+                ops = [A.one()]*m; ops[a] = eq
+                return ops[0].annihilator_of_interlacing(*(ops[1:]))
+            else:
+                eq = eq.polynomial().map_coefficients(lambda p: p(x**m))
+                if a != 0:
+                    eq = eq(x**(-a)*eq.parent().gen())*x**(a*eq.degree())
+                return A(eq)
+                
+    if R.is_field():
+        return guess(data, A.change_ring(R.ring()), **kwargs)
+                
     elif A.is_F() is not False:
         # reduce to shift case; note that this does not alter order or degrees
         if 'infolevel' in kwargs and kwargs['infolevel'] >= 1:

### 2019-05-09 src/ore_algebra/examples/ssw.py
ssw: test more of the 4*19 examples

@@ -41,10 +41,21 @@ An example provided by Bruno Salvy, also related to small-step walks::
     sage: ssw.aux_dop.numerical_solution(ssw.aux_ini, [0,1]) # (1.5 s), TODO: double-check result
     [10.662510694...]
 
+TESTS::
+
+    sage: for i in [2, 5, 12, 18]: # long time (1 min 40 s)
+    ....:     for xx, yy in [(0, 1), (1, 0)]:
+    ....:         ssw.test_ct(i, xx, yy)
+
+    sage: for i in [1..19]: # not tested (9 min 31 s)
+    ....:     for xx in [0, 1]:
+    ....:         for yy in [0, 1]:
+    ....:             ssw.test_ct(i, xx, yy)
+
 """
 
-from sage.rings.all import Integer, PolynomialRing, ZZ, QQ
-from ore_algebra import DifferentialOperators
+from sage.rings.all import Frac, Integer, PolynomialRing, ZZ, QQ
+from ore_algebra import DifferentialOperators, OreAlgebra
 
 Pols, (u, v, x, y, t) = PolynomialRing(ZZ, ('u', 'v', 'x', 'y', 't')).objgens()
 DiffOps, t, Dt = DifferentialOperators(QQ, 't')
@@ -1414,6 +1425,16 @@ aux_dop = (
     - 15345679965249398832*t**7
     - 3938256369400303296*t**6)*Dt**7
 )
+
+def test_ct(i, xx, yy, infolevel=0):
+    A, (Du,Dv,Dt) = OreAlgebra(Frac(ZZ['t,u,v']), 'Du', 'Dv', 'Dt').objgens()
+    q = A.base_ring()(rat[i](x=xx, y=yy))
+    ct1, _ = A.ideal([q*D - D(q) for D in Du,Dv,Dt]).ct(Dv)
+    ct2, _ = ct1[0].parent().ideal(ct1).ct(Du, infolevel=infolevel)
+    assert len(ct2) == 1
+    tel = ct2[0]
+    cst = tel//dop[i,xx,yy]
+    assert cst == 1 or cst == -1
 
 aux_ini = [ Integer(59)/Integer(36), Integer(547)/Integer(2160),
         Integer(1)/Integer(2)*Integer(229)/Integer(252),

### 2019-05-10 src/ore_algebra/guessing.py
fix guessing fix: allow data to contain Python ints

@@ -167,17 +167,17 @@ def guess(data, algebra, **kwargs):
         with open(data, 'r') as f:
             data = [ K(line) for line in f ]
 
-    if (data[0].is_zero() or data[1].is_zero()) and (A.is_C() or A.is_S()):
+    if (data[0] == 0 or data[1] == 0) and (A.is_C() or A.is_S()):
         
-        if all( d.is_zero() for d in data ):
+        if all( d == 0 for d in data ):
             return A.one()
         for i in range(len(data)):
-            if not data[i].is_zero():
+            if data[i] != 0:
                 a = i
                 break
         m = None; b = ZZ(a)
         for i in range(b + 1, len(data)):
-            if not data[i].is_zero():
+            if data[i] != 0:
                 m = i - b if m is None else m.gcd(i - b)
                 b = i
                 if m.is_one():

### 2019-05-10 src/ore_algebra/analytic/__init__.py
add copyright statements to analytic/*

@@ -459,6 +459,16 @@ To run the test suite of the ``ore_algebra.analytic`` subpackage, run::
     src$ PYTHONPATH="$PWD" sage -t ore_algebra/analytic/
 """
 
+# Copyright 2015, 2016, 2017, 2018, 2019 Marc Mezzarobba
+# Copyright 2015, 2016, 2017, 2018, 2019 Centre national de la recherche scientifique
+# Copyright 2015, 2016, 2017, 2018 Université Pierre et Marie Curie
+# Copyright 2019 Sorbonne Université
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 from __future__ import absolute_import
 
 from . import analytic_continuation as ancont

### 2019-05-10 src/ore_algebra/analytic/analytic_continuation.py

@@ -3,6 +3,16 @@
 Evaluation of univariate D-finite functions by numerical analytic continuation
 """
 
+# Copyright 2015, 2016, 2017, 2018, 2019 Marc Mezzarobba
+# Copyright 2015, 2016, 2017, 2018, 2019 Centre national de la recherche scientifique
+# Copyright 2015, 2016, 2017, 2018 Université Pierre et Marie Curie
+# Copyright 2019 Sorbonne Université
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 import logging
 
 import sage.rings.all as rings

### 2019-05-10 src/ore_algebra/analytic/binary_splitting.py

@@ -210,6 +210,16 @@ Miscellaneous examples::
     sage: logger.setLevel(logging.WARNING)
 """
 
+# Copyright 2015, 2016, 2017, 2018, 2019 Marc Mezzarobba
+# Copyright 2015, 2016, 2017, 2018, 2019 Centre national de la recherche scientifique
+# Copyright 2015, 2016, 2017, 2018 Université Pierre et Marie Curie
+# Copyright 2019 Sorbonne Université
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 from __future__ import print_function
 from six.moves import range
 

### 2019-05-10 src/ore_algebra/analytic/binary_splitting_arb.pyx

@@ -3,6 +3,15 @@ r"""
 Lower-level reimplementation of key subroutines of binary_splitting
 """
 
+# Copyright 2018 Marc Mezzarobba
+# Copyright 2018 Centre national de la recherche scientifique
+# Copyright 2018 Université Pierre et Marie Curie
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 from __future__ import print_function
 
 from sage.libs.arb.types cimport *

### 2019-05-10 src/ore_algebra/analytic/bounds.py

@@ -3,6 +3,16 @@ r"""
 Error bounds
 """
 
+# Copyright 2015, 2016, 2017, 2018, 2019 Marc Mezzarobba
+# Copyright 2015, 2016, 2017, 2018, 2019 Centre national de la recherche scientifique
+# Copyright 2015, 2016, 2017, 2018 Université Pierre et Marie Curie
+# Copyright 2019 Sorbonne Université
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 from __future__ import division, print_function
 from six.moves import range
 

### 2019-05-10 src/ore_algebra/analytic/context.py

@@ -3,6 +3,15 @@ r"""
 Analytic continuation contexts
 """
 
+# Copyright 2018 Marc Mezzarobba
+# Copyright 2018 Centre national de la recherche scientifique
+# Copyright 2018 Université Pierre et Marie Curie
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 class Context(object):
 
     def __init__(self, dop=None, path=None, eps=None,

### 2019-05-10 src/ore_algebra/analytic/function.py

@@ -50,6 +50,16 @@ TESTS::
 
 """
 
+# Copyright 2017, 2018, 2019 Marc Mezzarobba
+# Copyright 2017, 2018, 2019 Centre national de la recherche scientifique
+# Copyright 2017, 2018 Université Pierre et Marie Curie
+# Copyright 2019 Sorbonne Université
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 from six import iteritems
 from six.moves import range
 

### 2019-05-10 src/ore_algebra/analytic/monodromy.py

@@ -35,6 +35,16 @@ TESTS::
     True
 """
 
+# Copyright 2017, 2018, 2019 Marc Mezzarobba
+# Copyright 2017, 2018, 2019 Centre national de la recherche scientifique
+# Copyright 2017, 2018 Université Pierre et Marie Curie
+# Copyright 2019 Sorbonne Université
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 import logging
 
 import sage.matrix.special as matrix

### 2019-05-10 src/ore_algebra/analytic/naive_sum.py

@@ -3,6 +3,16 @@
 Evaluation of convergent D-finite series by direct summation
 """
 
+# Copyright 2015, 2016, 2017, 2018, 2019 Marc Mezzarobba
+# Copyright 2015, 2016, 2017, 2018, 2019 Centre national de la recherche scientifique
+# Copyright 2015, 2016, 2017, 2018 Université Pierre et Marie Curie
+# Copyright 2019 Sorbonne Université
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 # TODO:
 # - support summing a given number of terms rather than until a target accuracy
 # is reached?

### 2019-05-10 src/ore_algebra/analytic/naive_sum_c.pyx

@@ -3,6 +3,15 @@ r"""
 Lower-level reimplementation of key subroutines of naive_sum
 """
 
+# Copyright 2018 Marc Mezzarobba
+# Copyright 2018 Centre national de la recherche scientifique
+# Copyright 2018 Université Pierre et Marie Curie
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 from __future__ import print_function
 
 from libc.stdlib cimport malloc, free

### 2019-05-10 src/ore_algebra/analytic/path.py

@@ -3,6 +3,16 @@
 Analytic continuation paths
 """
 
+# Copyright 2015, 2016, 2017, 2018, 2019 Marc Mezzarobba
+# Copyright 2015, 2016, 2017, 2018, 2019 Centre national de la recherche scientifique
+# Copyright 2015, 2016, 2017, 2018 Université Pierre et Marie Curie
+# Copyright 2019 Sorbonne Université
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 import logging
 
 import sage.plot.all as plot

### 2019-05-10 src/ore_algebra/analytic/safe_cmp.py

@@ -3,6 +3,15 @@ r"""
 Safe comparisons
 """
 
+# Copyright 2015 Marc Mezzarobba
+# Copyright 2015 Centre national de la recherche scientifique
+# Copyright 2015 Université Pierre et Marie Curie
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 __all__ = ["safe_lt", "safe_le", "safe_gt", "safe_ge", "safe_eq", "safe_ne"]
 
 import logging

### 2019-05-10 src/ore_algebra/analytic/ui.py

@@ -4,6 +4,15 @@ Some convenience functions for features not yet easily accessible from methods
 of differential operators.
 """
 
+# Copyright 2015, 2016, 2017, 2018 Marc Mezzarobba
+# Copyright 2015, 2016, 2017, 2018 Centre national de la recherche scientifique
+# Copyright 2015, 2016, 2017, 2018 Université Pierre et Marie Curie
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 from ..ore_algebra import DifferentialOperators
 from . import analytic_continuation as ancont
 from . import polynomial_approximation as polapprox

### 2019-05-10 src/ore_algebra/analytic/utilities.py

@@ -3,6 +3,15 @@
 Miscellaneous utilities
 """
 
+# Copyright 2015, 2016, 2017, 2018 Marc Mezzarobba
+# Copyright 2015, 2016, 2017, 2018 Centre national de la recherche scientifique
+# Copyright 2015, 2016, 2017, 2018 Université Pierre et Marie Curie
+#
+# Distributed under the terms of the GNU General Public License (GPL) either
+# version 2, or (at your option) any later version
+#
+# http://www.gnu.org/licenses/
+
 import sage.rings.complex_arb
 import sage.rings.real_arb
 

### 2019-05-14 src/ore_algebra/analytic/differential_operator.py
extend_scalars: implement extensions by several elements

@@ -13,7 +13,6 @@ Custom differential operators
 # http://www.gnu.org/licenses/
 
 from sage.arith.all import lcm
-from sage.categories.pushout import pushout
 from sage.misc.cachefunc import cached_method
 from sage.rings.all import CIF, QQbar, QQ, ZZ
 from sage.rings.complex_arb import ComplexBallField
@@ -22,6 +21,7 @@ from sage.rings.infinity import infinity
 from sage.rings.number_field.number_field import is_NumberField
 from sage.rings.polynomial.polynomial_ring_constructor import PolynomialRing
 from sage.structure.coerce_exceptions import CoercionException
+from sage.structure.element import coercion_model
 
 from ..ore_algebra import OreAlgebra
 from ..ore_operator_1_1 import UnivariateDifferentialOperatorOverUnivariateRing
@@ -154,34 +154,35 @@ class PlainDifferentialOperator(UnivariateDifferentialOperatorOverUnivariateRing
             mag = IR.zero()
         return int(est.ceil().upper()), int(mag.ceil().upper())
 
-    def extend_scalars(self, pt):
+    def extend_scalars(self, *pts):
         r"""
-        Extend the ground field so that the new field contains pt.
+        Extend the ground field so that the new field contains pts.
         """
         Dops = self.parent()
         Pols = Dops.base_ring()
         Scalars = Pols.base_ring()
-        if Scalars.has_coerce_map_from(pt.parent()):
-            return self, pt
+        if all(Scalars.has_coerce_map_from(pt.parent()) for pt in pts):
+            return (self,) + pts
         gen = Scalars.gen()
         try:
             # Largely redundant with the other branch, but may do a better job
             # in some cases, e.g. pushout(QQ, QQ(α)), where as_enf_elts() would
             # invent new generator names.
-            NF = pushout(Scalars, pt.parent())
+            NF = coercion_model.common_parent(Scalars, **pts)
             if not is_NumberField(NF):
                 raise CoercionException
             gen1 = NF.coerce(gen)
-            pt1 = NF.coerce(pt)
+            pts1 = tuple(NF.coerce(pt) for pt in pts)
         except (CoercionException, TypeError):
-            NF, (gen1, pt1) = as_embedded_number_field_elements([gen,pt])
+            NF, val1 = as_embedded_number_field_elements((gen,)+pts)
+            gen1, pts1 = val1[0], tuple(val1[1:])
[…]

### 2019-05-14 src/ore_algebra/analytic/binary_splitting.py
binsplit: algebraic points and exponents, ctd

@@ -116,8 +116,7 @@ Whittaker functions with irrational exponents::
     [[-3.829367993175840...] + [+/-...]*I  [7.857756823216...] + [+/-...]*I]
     [[-1.135875563239369...] + [+/-...]*I  [1.426170676718...] + [+/-...]*I]
 
-Various mixes of algebraic exponents and evaluation points (TODO: add more
-complicated combinations, after improving the code...)::
+Various mixes of algebraic exponents and evaluation points::
 
     sage: ((x*Dx)^3-2-x).numerical_transition_matrix([0,1], 1e-5, algorithm="binsplit")
     [ [0.75335...] + [-0.09777...]*I  [0.75335...] + [0.09777...]*I  [1.1080...] + [+/- ...]*I]
@@ -145,6 +144,9 @@ complicated combinations, after improving the code...)::
     sage: (((x-i)*Dx)^2+1-x).numerical_transition_matrix([i, sqrt(2)], algorithm="binsplit")
     [ [0.69261048...] + [0.16794895...]*I  [2.70326173...] + [-0.78385034...]*I]
     [[0.2516164...] + [-0.10376355...]*I   [0.68776459...] + [1.07708951...]*I]
+
+    sage: (Dx - i).numerical_solution([1], [0, sqrt(2)], algorithm="binsplit")
+    [0.1559436947653...] + [0.9877659459927...]*I
 
 An interesting “real-world” example, where one of the local exponents is
 irrational, but in the same extension as the singularity itself::
@@ -240,7 +242,6 @@ import sage.rings.polynomial.polynomial_ring as polyring
 import sage.rings.polynomial.polynomial_ring_constructor as polyringconstr
 
 from sage.arith.all import gcd
-from sage.categories.pushout import pushout
 from sage.matrix.constructor import matrix
 from sage.matrix.matrix_space import MatrixSpace
 from sage.modules.free_module_element import vector
@@ -768,8 +769,9 @@ class MatrixRec(object):
             except CoercionException:
                 # Not great, but allows us to handle a few combination of
                 # algebraic points that we couldn't otherwise...
-                dop, dz = dop.extend_scalars(dz)
+                dop, dz, shift = dop.extend_scalars(dz, shift)
                 deq_Scalars = dop.base_ring().base_ring()
+                E = dz.parent()
                 self._init_generic(deq_Scalars, shift, E, dz)
         self.dz = dz
 
@@ -860,7 +862,7 @@ class MatrixRec(object):
[…]

### 2019-05-14 src/ore_algebra/analytic/local_solutions.py

@@ -20,7 +20,6 @@ import collections, logging
 from itertools import chain
 
 from sage.arith.all import gcd, lcm
-from sage.categories.pushout import pushout
 from sage.misc.cachefunc import cached_method
 from sage.misc.lazy_attribute import lazy_attribute
 from sage.modules.free_module_element import vector
@@ -46,7 +45,7 @@ logger = logging.getLogger(__name__)
 ##############################################################################
 
 def bw_shift_rec(dop, shift=ZZ.zero(), clear_denominators=False):
-    Scalars = pushout(dop.base_ring().base_ring(), shift.parent())
+    Scalars = utilities.mypushout(dop.base_ring().base_ring(), shift.parent())
     if dop.parent().is_D():
         dop = DifferentialOperator(dop) # compatibility bugware
         rop = dop._my_to_S()
@@ -470,7 +469,7 @@ def exponent_shifts(dop, leftmost):
     return shifts
 
 def log_series(ini, bwrec, order):
-    Coeffs = pushout(bwrec.base_ring.base_ring(), ini.universe)
+    Coeffs = utilities.mypushout(bwrec.base_ring.base_ring(), ini.universe)
     log_prec = sum(len(v) for v in ini.shift.values())
     precomp_len = max(1, bwrec.order) # hack for recurrences of order zero
     bwrec_nplus = collections.deque(

### 2019-05-14 src/ore_algebra/analytic/path.py

@@ -427,7 +427,7 @@ class EvaluationPoint(object):
 
     def jet(self, Intervals):
         base_ring = (Intervals if self.is_numeric
-                     else pushout(self.pt.parent(), Intervals))
+                     else mypushout(self.pt.parent(), Intervals))
         Pol = PolynomialRing(base_ring, 'delta')
         return Pol([self.pt, 1]).truncate(self.jet_order)
 

### 2019-05-14 src/ore_algebra/analytic/utilities.py

@@ -20,7 +20,7 @@ from sage.misc.cachefunc import cached_function
 from sage.misc.misc import cputime
 from sage.rings.all import ZZ, QQ, QQbar, CIF
 from sage.rings.number_field.number_field import (NumberField,
-        NumberField_quadratic)
+        NumberField_quadratic, is_NumberField)
 from sage.rings.polynomial.polynomial_ring_constructor import PolynomialRing
 from sage.rings.qqbar import number_field_elements_from_algebraics
 
@@ -129,6 +129,17 @@ def invert_order_element(alg):
         modulus = Order.gen(1).minpoly()
         den, num, _ = pol.xgcd(modulus)  # hopefully fraction-free!
         return Order(num), ZZ(den)
+
+def mypushout(X, Y):
+    if X.has_coerce_map_from(Y):
+        return X
+    elif Y.has_coerce_map_from(X):
+        return Y
+    else:
+        Z = pushout(X, Y)
+        assert (is_NumberField(Z) if is_NumberField(X) and is_NumberField(Y)
+                                  else True)
+        return Z
 
 ######################################################################
 # Sage features

### 2019-05-14 src/ore_algebra/analytic/local_solutions.py
binsplit: fix denominator removal bug
in the computation of local recurrence operators at singularities of
non-integer exponent

@@ -53,7 +53,9 @@ def bw_shift_rec(dop, shift=ZZ.zero(), clear_denominators=False):
         Pols_n = PolynomialRing(dop.base_ring().base_ring(), 'n')
         rop = dop.to_S(ore_algebra.OreAlgebra(Pols_n, 'Sn'))
     Pols_n, n = rop.base_ring().change_ring(Scalars).objgen()
-    rop = rop.change_ring(Pols_n)
+    Rops = ore_algebra.OreAlgebra(Pols_n, 'Sn')
+    ordrec = rop.order()
+    rop = Rops([p(n-ordrec+shift) for p in rop])
     if clear_denominators:
         den = lcm([p.denominator() for p in rop])
         rop = den*rop
@@ -72,9 +74,7 @@ def bw_shift_rec(dop, shift=ZZ.zero(), clear_denominators=False):
         g = None
     if g is not None:
         rop = (1/g)*rop
-    ordrec = rop.order()
-    coeff = [rop[ordrec-k](n-ordrec+shift)
-             for k in range(ordrec+1)]
+    coeff = [rop[ordrec-k] for k in range(ordrec+1)]
     return BwShiftRec(coeff)
 
 class BwShiftRec(object):

### 2019-05-14 src/ore_algebra/examples/iint.py

@@ -47,6 +47,8 @@ Thanks to Jakob Ablinger and Clemens Raab for the data.
     sage: myini = [0, 0, 0, 1/3, -2/3-i*pi/3, 11/12+2*i*pi/3-pi^2/6]
     sage: iint_value(dop, myini)
     [0.97080469562493...]
+    sage: iint_value(dop, myini, algorithm="binsplit")
+    [0.97080469...]
     sage: iint_value(dop, myini, 1e-500) # long time (4.7 s)
     [0.97080469562493...0383420...]
 

### 2019-05-14 src/ore_algebra/analytic/examples/misc.py
improve printing of expansions of local solutions
Thanks to Emre Sertöz for complaining about the issue :-)

@@ -28,7 +28,7 @@ numerical analytic continuation code::
     [4.64191240683...] + [-0.01596122801...]*I
     sage: melczer1.local_basis_expansions(rts[1])
     [1 + (1269/32*a+3105/28)*(z + 0.086...? + 0.069...*I)^4 + ...,
-     sqrt(z + 0.086...? + 0.069...*I) + (365/96*a+13/3)*(z + 0.086...? + 0.069...*I)^(3/2) - ...,
+     (z + 0.086...? + 0.069...*I)^(1/2) + (365/96*a+13/3)*(z + 0.086...? + 0.069...*I)^(3/2) - ...,
      ...]
 """
 import collections

### 2019-05-14 src/ore_algebra/analytic/local_solutions.py

@@ -19,11 +19,14 @@ import collections, logging
 
 from itertools import chain
 
+import sage.functions.log as symbolic_log
+
 from sage.arith.all import gcd, lcm
 from sage.misc.cachefunc import cached_method
 from sage.misc.lazy_attribute import lazy_attribute
 from sage.modules.free_module_element import vector
-from sage.rings.all import ZZ, QQ, QQbar, RBF, RealBallField, ComplexBallField
+from sage.rings.all import ZZ, QQ, AA, QQbar, RBF
+from sage.rings.all import RealBallField, ComplexBallField
 from sage.rings.complex_arb import ComplexBall
 from sage.rings.integer import Integer
 from sage.rings.number_field.number_field import NumberField_absolute
@@ -558,3 +561,72 @@ def _pow_trunc(a, n, ord):
         n = n >> 1
     return pow
 
+##############################################################################
+# Human-readable representations that avoid various issues with symbolic
+# expressions
+##############################################################################
+
+def simplify_exponent(e):
+    r"""
+    TESTS::
+
+        sage: from ore_algebra.examples import cbt
+        sage: lc = cbt.dop[10].leading_coefficient()
+        sage: s = sorted(lc.roots(QQbar, multiplicities=False), key=abs)[0]
+        sage: cbt.dop[10].local_basis_monomials(s)
+        [1,
+        z - 0.2651878342412026?,
+        (z - 0.2651878342412026?)^2,
+        (z - 0.2651878342412026?)^3,
+        (z - 0.2651878342412026?)^4,
+        (z - 0.2651878342412026?)^4.260514654474679?,
+        (z - 0.2651878342412026?)^5,
+        (z - 0.2651878342412026?)^6,
+        (z - 0.2651878342412026?)^7,
+        (z - 0.2651878342412026?)^8,
+        (z - 0.2651878342412026?)^9]
+        sage: cbt.dop[10].local_basis_expansions(s, 1)
+        [1, 0, 0, 0, 0, (z - 0.2651878342412026?)^4.260514654474679?, 0, 0, 0,
+        0, 0]
+    """
+    for dom in ZZ, QQ, AA, QQbar:
+        try:
+            return dom(e)
+        except (TypeError, ValueError):
+            pass
+    return e
+
+class LogMonomial(object):
+
+    def __init__(self, dx, expo, shift, k):
+        self.dx = dx
+        self.expo = simplify_exponent(expo)
+        self.shift = shift
+        self.n = self.expo + shift
+        self.k = k
+
[…]

### 2019-05-14 src/ore_algebra/ore_operator_1_1.py

@@ -2253,11 +2253,12 @@ class UnivariateDifferentialOperatorOverUnivariateRing(UnivariateOreOperatorOver
             [1, sqrt(x), x]
         """
         from .analytic.differential_operator import DifferentialOperator
+        from .analytic.local_solutions import simplify_exponent
         from .analytic.path import Point
         dop = DifferentialOperator(self)
         struct = Point(point, dop).local_basis_structure()
         x = SR(dop.base_ring().gen()) - point
-        return [x**_simplify_exponent(sol.valuation)
+        return [x**simplify_exponent(sol.valuation)
                     *symbolic_log.log(x, hold=True)**sol.log_power
                     /sol.log_power.factorial()
                 for sol in struct]
@@ -2301,9 +2302,9 @@ class UnivariateDifferentialOperatorOverUnivariateRing(UnivariateOreOperatorOver
 
             sage: from ore_algebra.examples import ssw
             sage: ssw.dop[1,0,0].local_basis_expansions(0)
-            [t^(-4) + 24*log(t)/t^2 - 48*log(t) - 96*t^2*log(t) - 88*t^2,
-            t^(-2),
-            1 + 2*t^2]
+            [t^(-4) + 24*t^(-2)*log(t) - 48*log(t) - 96*t^2*log(t) - 88*t^2,
+             t^(-2),
+             1 + 2*t^2]
 
             sage: dop = (x^2*(x^2-34*x+1)*Dx^3 + 3*x*(2*x^2-51*x+1)*Dx^2
             ....:     + (7*x^2-112*x+1)*Dx + (x-5))
@@ -2317,10 +2318,8 @@ class UnivariateDifferentialOperatorOverUnivariateRing(UnivariateOreOperatorOver
             sage: basis = dop.local_basis_expansions(roots[1][0], order=3)
             sage: basis
             [1 - (-239/12*a+169/6)*(x - 0.02943725152285942?)^2,
-            sqrt(x - 0.02943725152285942?)
-            - (-203/32*a+9)*(x - 0.02943725152285942?)^(3/2)
-            + (-24031/160*a+1087523/5120)*(x - 0.02943725152285942?)^(5/2),
-            x - 0.02943725152285942? - (-55/6*a+13)*(x - 0.02943725152285942?)^2]
[…]

### 2019-05-14 src/ore_algebra/analytic/examples/misc.py
update timings

@@ -14,7 +14,7 @@ equation::
     sage: Pols.<z> = QQ[]
     sage: a = AA.polynomial_root(54*z**3+324*z**2-4265*z+432, RIF(0.1, 0.11))
     sage: roots = salvy1_pol(z=a).univariate_polynomial().roots(QQbar)
-    sage: val = salvy1_dop.numerical_solution([0, 0, 0, 0, 0, 1/2], [0, a]) # long time (6 s)
+    sage: val = salvy1_dop.numerical_solution([0, 0, 0, 0, 0, 1/2], [0, a]) # long time (5.2 s)
     sage: CBF100 = ComplexBallField(100)
     sage: [r for (r, _) in roots if CBF100(r) in val]                       # long time
     [0.0108963334211605...]

### 2019-05-14 src/ore_algebra/analytic/function.py

@@ -509,7 +509,7 @@ def _tests():
 
         sage: f = DFiniteFunction((x^2 + 1)*Dx^2 + 2*x*Dx, [0, 1])
 
-        sage: [f(10^i) for i in range(-3, 4)] # long time (3 s)
+        sage: [f(10^i) for i in range(-3, 4)] # long time (2.5 s)
         [[0.0009999996666...], [0.0099996666866...], [0.0996686524911...],
         [0.7853981633974...], [1.4711276743037...], [1.5607966601082...],
         [1.5697963271282...]]

### 2019-05-14 src/ore_algebra/examples/cbt.py

@@ -71,7 +71,7 @@ basis (1 + O(z^11), z + O(z^11), ...) to its coefficients in the basis::
 
 ::
 
-    sage: mat = dop.numerical_transition_matrix([0,s], 1e-30) # long time (3.9 s)
+    sage: mat = dop.numerical_transition_matrix([0,s], 1e-30) # long time (3.6 s)
 
 We are interested in the fifth row::
 
@@ -176,7 +176,7 @@ can automate the process for other values of k::
 
 We obtain::
 
-    sage: for k, dop in cbt.dop.items(): # long time (11.2 s)
+    sage: for k, dop in cbt.dop.items(): # long time (10 s)
     ....:     print("{}\t{}".format(k, asy(dop, 50)))
     2  [0.5613226189564568270393235883810334361992061196...]
     3  [0.6049645385653542644762421366594344081594004532...]

### 2019-05-14 src/ore_algebra/examples/fcc.py

@@ -44,7 +44,7 @@ In this case, the segment [0, 1] contains a singularity of the operator::
 We can nevertheless evaluate the solution of interest using an integration path
 that passes above the singular point::
 
-    sage: fcc.dop5.numerical_solution([0, 0, 0, 0, 1, 0], [0, 1/5+i/2, 1], 1e-60) # (1.7 s)
+    sage: fcc.dop5.numerical_solution([0, 0, 0, 0, 1, 0], [0, 1/5+i/2, 1], 1e-60) # (1.3 s)
     [1.04885235135491485162956376369999275945402550465206640313845...] + [+/-...]*I
 
 In the six-dimensional case, Koutschan gives the following operator::
@@ -53,14 +53,14 @@ In the six-dimensional case, Koutschan gives the following operator::
     (27122036833024*z^43 + ... + 2276991208061142220800000000000000000*z^6)*Dz^8
     + ... - 2428790621931885035520000000000000000
     sage: ini = [0, 0, 0, 0, 0, 1, 0, 0]
-    sage: fcc.dop6.numerical_solution(ini, [0, 3/2 + i, 1], 1e-60) # long time (14 s)
+    sage: fcc.dop6.numerical_solution(ini, [0, 3/2 + i, 1], 1e-60) # long time (9.3 s)
     [1.02774910062749883985936367927396850209243990900114872425...] + [...]*I
 
 TESTS:
 
 An experimental feature::
 
-    sage: fcc.dop5.numerical_solution([0, 0, 0, 0, 1, 0], [0, 1], assume_analytic=True) # (1.5 s)
+    sage: fcc.dop5.numerical_solution([0, 0, 0, 0, 1, 0], [0, 1], assume_analytic=True) # (1.1 s)
     [1.0488523513549...] + [+/- ...]*I
 """
 

### 2019-05-14 src/ore_algebra/examples/iint.py

@@ -49,7 +49,7 @@ Thanks to Jakob Ablinger and Clemens Raab for the data.
     [0.97080469562493...]
     sage: iint_value(dop, myini, algorithm="binsplit")
     [0.97080469...]
-    sage: iint_value(dop, myini, 1e-500) # long time (4.7 s)
+    sage: iint_value(dop, myini, 1e-500) # long time (3.4 s)
     [0.97080469562493...0383420...]
 
 Known exact values::
@@ -91,7 +91,7 @@ Known exact values::
     sage: ref[23] = 4/9*(psi1_a - 4*zeta(2))
     sage: ref[24] = sqrt(2)*(2/3*zeta(2) - 2*Li_b - ln(2)^2)
 
-    sage: for k in sorted(word.keys()): # long time (21 s)
+    sage: for k in sorted(word.keys()): # long time (16 s)
     ....:     dop = diffop(word[k])
     ....:     val = iint_value(dop, ini[k])
     ....:     ok = "ok" if k in ref and val in RBF(ref[k]).add_error(1e-10) else ""

### 2019-05-14 src/ore_algebra/examples/periods.py

@@ -33,10 +33,10 @@ TESTS::
     [[0.02222358650439156...] [-0.3691184858549074...]   [1.978544499665723...]  [-2.464594622238230...]   [5.583417676436902...]]
 
     sage: from ore_algebra.examples.periods import dop_140118_4, dop_140118_6_4_2
-    sage: dop_140118_4.numerical_transition_matrix([0,1], 1e-200, assume_analytic=True, squash_intervals=True)[0,0] # long time (4.7 s)
+    sage: dop_140118_4.numerical_transition_matrix([0,1], 1e-200, assume_analytic=True, squash_intervals=True)[0,0] # long time (3.4 s)
     [0.715789877802957863238550457253731250968474707759494604384565678079788875083797614717578209783780445614354477440408242665736695668023525540188926510250252867846105078425273565096207343428573617619055...]
 
-    sage: dop_140118_6_4_2.numerical_transition_matrix([0,0.6+0.2*I,1], 1e-150, assume_analytic=True)[0,0] # long time (60 s)
+    sage: dop_140118_6_4_2.numerical_transition_matrix([0,0.6+0.2*I,1], 1e-150, assume_analytic=True)[0,0] # long time (46 s)
     [0.869416096203083843226211954657998870957512910057684233369962964198185589472701159476011286381945095930579858841819223890538...] +
     [0.111837076188766007428795107583040967018817636196931759352101655227626469869765071421524590125612423060982791983438726224032...]*I
 """

### 2019-05-14 src/ore_algebra/examples/ssw.py

@@ -49,7 +49,7 @@ TESTS::
 
 Plus an example provided by Bruno Salvy, also related to small-step walks::
 
-    sage: ssw.aux_dop.numerical_solution(ssw.aux_ini, [0,1]) # (1.5 s), TODO: double-check result
+    sage: ssw.aux_dop.numerical_solution(ssw.aux_ini, [0,1]) # TODO: double-check result
     [10.662510694...]
 
 """

### 2019-05-14 src/ore_algebra/guessing.py

@@ -144,8 +144,8 @@ def guess(data, algebra, **kwargs):
     EXAMPLES::
 
       sage: from ore_algebra import *
-      sage: rec = guess([(2*i+1)^15 * (1 + 2^i + 3^i)^2 for i in range(1000)], OreAlgebra(ZZ['n'], 'Sn'))
-      sage: rec.order(), rec.degree()
+      sage: rec = guess([(2*i+1)^15 * (1 + 2^i + 3^i)^2 for i in range(1000)], OreAlgebra(ZZ['n'], 'Sn')) # long time (5 s)
+      sage: rec.order(), rec.degree() # long time
       (6, 90)
       sage: R.<t> = QQ['t']
       sage: rec = guess([1/(i+t) + t^i for i in range(100)], OreAlgebra(R['n'], 'Sn'))

### 2019-05-14 src/ore_algebra/ore_operator_1_1.py

@@ -3037,7 +3037,7 @@ class UnivariateRecurrenceOperatorOverUnivariateRing(UnivariateOreOperatorOverUn
             sage: Rxks = OreAlgebra(Rxk, 'Sk')
             sage: V = QQ
             sage: Vks = OreAlgebra(V['k'], 'Sk')
-            sage: for i in range(1000): # long time (4 s)
+            sage: for i in range(1000): # long time (2.5 s)
             ....:     A = Rxks.random_element(randrange(1,4))
             ....:     r = A.order()
             ....:     v = V.random_element()

### 2019-05-15 papers/icms2016/main.tex
add a copy of my ICMS paper to the repo

\documentclass[runningheads,a4paper]{llncs}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage[pdfborder={0 0 0}]{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{url}
\usepackage{tikz}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}
\begin{document}

\mainmatter

%%%%%%%%%% Start TeXmacs macros
\newcommand{\Omicron}{\mathrm{O}}
\let\tmem\emph
\let\mathbbm\mathbb
\newcommand{\tminput}[2]{\trivlist{\item{#1#2}}}
\newcommand{\tmop}[1]{\ensuremath{\operatorname{#1}}}
\newcommand{\tmsession}[3]{{\topsep=.5ex \partopsep=.5ex\ttfamily\footnotesize#3\par\vspace{1ex}}}
\newcommand{\tmtexttt}[1]{{\ttfamily{#1}}}
\newcommand{\tmunfoldedio}[3]{\trivlist{\item{#1#2}\item{#3}}}
%%%%%%%%%% End TeXmacs macros

\title{Rigorous Multiple-Precision Evaluation of D-Finite Functions in SageMath}
\titlerunning{Multiple-Precision D-Finite Functions in SageMath}
\author{Marc Mezzarobba%
\thanks{Supported in part by ANR grant ANR-14-CE25-0018-01 (FastRelax).}
}
\authorrunning{Mezzarobba}
\institute{
  CNRS, LIP6, Universit\'e Pierre et Marie Curie%
  \thanks{
    Sorbonne Universit\'es, UPMC Univ Paris 06, CNRS, LIP6 UMR 7606,
    4~place Jussieu 75005 Paris.
  },
  Paris, France\\
  \email{marc@mezzarobba.net},\\
  \texttt{http://marc.mezzarobba.net/}
}
\maketitle

\begin{abstract}
  We present a new open source implementation in the SageMath computer algebra
  system of algorithms for the numerical solution of linear ODEs with
  polynomial coefficients. Our code supports regular singular connection
  problems and provides rigorous error bounds.
  \let\thefootnote\relax
  \footnotetext{This article is in the public domain.
  In jurisdictions where this is not possible, any entity is granted the
  perpetual right to use this work for any purpose, without any conditions
  other than those required by law.}
  \begin{tikzpicture}[remember picture,overlay]
    \node [yshift=6cm, xshift=1em, anchor=north] at (current page.south)
    [text width=\textwidth, font=\footnotesize, align=justify] {
    Extended abstract for a talk given at the 5th~International Congress on Mathematical Software (ICMS~2016).
    Accepted for publication in the proceedings, but withdrawn due to a disagreement with Springer about the above public domain notice.
    };
  \end{tikzpicture}
\end{abstract}

\section{Introduction}

Many special functions satisfy differential equations
\begin{equation}
  p_r (x) f^{(r)} (x) + \cdots + p_1 (x) f^{\prime} (x) + p_0 (x) f (x) = 0
  \label{eq:deq}
\end{equation}
whose coefficients~$p_i$ depend polynomially on the variable~$x$. In virtually
all cases, such special functions can be defined by
complementing~(\ref{eq:deq}) either with simple initial values $f (0),
f^{\prime} (0), \ldots$ or with constraints on the asymptotic behavior of~$f
(x)$ as $x$~approaches a singular point. For example, the error function
satisfies
\begin{equation}
  \tmop{erf}^{\prime\prime} (x) + 2 x \tmop{erf}^{\prime} (x) = 0,
  \hspace{2em} \tmop{erf} (0) = 0, \hspace{1em} \tmop{erf}^{\prime} (0) =
  \frac{2}{\sqrt{\pi}}, \label{eq:erf}
\end{equation}
while the modified Bessel function~$K_0$ is the solution of
\begin{equation}
  xK_0^{\prime\prime} (x) + K_0^{\prime} (x) - xy (x) = 0 \hspace{1em}
  \text{s.t.} \hspace{1em} K_0 (x) = - \log (x / 2) - \gamma + \Omicron_{x
  \rightarrow 0} (x) . \label{eq:Bessel}
\end{equation}
This observation has led to the idea of developing algorithms that deal with
these functions in a uniform way, using the ODE~(\ref{eq:deq}) as a data
structure~{\cite{Lanczos1956,BenoitChyzakDarrasseGerholdMezzarobbaSalvy2010}}.

In this context, solutions of linear ODEs with polynomial coefficients are
called {\tmem{D-finite}} (or {\tmem{holonomic}}) functions. These names
originate from combinatorics, where D-finite power series arise naturally as
generating functions~{\cite{Stanley1999,FlajoletSedgewick2009}}. While
classical special functions typically satisfy ODEs of order 2~to~4 with simple
polynomial coefficients, D-finite generating functions are usually given by
more general and sometimes very large equations. Other notable sources of
general D-finite functions include mathematical physics and number theory.
Explicit examples in theoretical works on meromorphic ODEs also tend to have
polynomial coefficients.

Packages for manipulating D-finite power series
symbolically exist for major computer algebra systems, and include in
particular gfun~{\cite{SalvyZimmermann1994}} and the rest of the algolib
bundle for Maple, GeneratingFunctions~{\cite{Mallinger1996}} and the
RISCErgoSum suite for Mathematica, and
\tmtexttt{ore\_algebra}~{\cite{KauersJaroschekJohansson2015}} for
SageMath~{\cite{Sage}}.

We extended \tmtexttt{ore\_algebra} with features dedicated to D-finite
{\tmem{analytic functions}} such as special functions, starting with numerical
evaluation algorithms. The need for such features is most evident in the
context of special functions, but they turn out to be useful in a wide range
of applications, as we will see from the examples in the sequel. We refer to
our implementation as \tmtexttt{oa.analytic}, after the name of the subpackage
of \tmtexttt{ore\_algebra} where most of the code resides. The present paper
introduces the main features of \tmtexttt{oa.analytic}, with example
applications. The reader is invited to consult the package's documentation for
more information. We further refer to Hille~{\cite{Hille1976}} for more
background on complex analytic ODEs, and to Kauers~{\cite{Kauers2013}} for an
overview of the field of D-finite functions in computer algebra. Also note
that \tmtexttt{oa.analytic} is intended to supersede
NumGfun~{\cite{Mezzarobba2010}}, a Maple package with similar features
previously developed by the author\footnote{Many of the examples in this
article are adapted from earlier work on NumGfun~{\cite{Mezzarobba2011}}.}.

Since the main feature of \tmtexttt{oa.analytic} is to compute numerical
values of solutions of differential equations, it is, at its core, a numerical
ODE solver. However, it is limited to linear ODEs with polynomial coefficients, and differs from typical ODE solvers in a number of other respects. First
of all, the results it computes come with {\tmem{rigorous error bounds}},
making it a ``validated'' ODE solver. It works in {\tmem{arbitrary
precision}}, and implements specific algorithms to make evaluations at very
high precisions feasible. Last but not least, it offers extensive support for
connection problems between {\tmem{regular singular points}} of differential
operators. These distinctive features all rely on {\tmem{symbolic-numeric
algorithms}} that make use of specific properties of D-finite functions.

The development of \tmtexttt{oa.analytic} is still very much a work in
progress. At the time of writing, there is no formal release of
\tmtexttt{ore\_algebra} containing the \tmtexttt{analytic} subpackage yet. The
latest development version of \tmtexttt{oa.analytic} can be found at
{\url{http://marc.mezzarobba.net/code/ore_algebra-analytic}}. It
is distributed under the GNU General Public Licence, version~2 or later.
Comments, bug reports and feature requests are welcome.

\section{Differential Operators in \tmtexttt{ore\_algebra}}

In \tmtexttt{ore\_algebra}, differential operators are represented
algebraically as {\tmem{Ore polynomials}}. More specifically, the operators we
are interested in are elements of the ring, denoted $\mathbbm{K} [x] \langle
D_x \rangle$, of polynomials over $\mathbbm{K} [x]$ for some number field
$\mathbbm{K} \subset \mathbbm{C}$ in a variable~$D_x$ subject to the
commutation rule $D_x x = xD_x + 1$ (which expresses the equality $(xf
(x))^{\prime} = xf^{\prime} (x) + f (x)$). Elements of $\mathbbm{K} [x]
\langle D_x \rangle$ act on differentiable functions in a natural way, namely
by $x \cdot f = (\xi \mapsto \xi f (\xi))$ and $D_x \cdot f = f^{\prime}$, and
can thus be identified with differential operators.

The first thing to do in order to use \tmtexttt{ore\_algebra} in a Sage
session is to install it in the Python search path and \tmtexttt{import} it.
The rings $\mathbbm{K} [x]$ and $\mathbbm{K} [x] \langle D_x \rangle$ and
their respective generators $x$~and~$D_x$ can then be created using standard
Sage syntax as follows:

\tmsession{sage}{default}{
  \tminput{sage: }{from ore\_algebra import OreAlgebra}
  \tminput{sage: }{Pols.<x> = PolynomialRing(QQ); Dops.<Dx> =
  OreAlgebra(Pols)}
}

{\noindent}Note that \tmtexttt{ore\_algebra} supports general Ore polynomials,
but basic arithmetic with differential operators in one variable is all we are
going to need here.

After these commands, the Python variable~\tmtexttt{Dx} contains an object of
type ``Ore polynomial'' that satisfies

\tmsession{sage}{default}{\tmunfoldedio{sage: }{Dx*x}{x*Dx + 1}}

{\noindent}Most features of \tmtexttt{ore\_algebra}, including those presented
in this article, are available as methods of Ore polynomial objects. Some
functions of the \tmtexttt{analytic} subpackage can also be called directly in
advanced usage.

\section{Numerical Analytic Continuation}

Let $L = p_r (x) D_x^r + \cdots + p_0 (x) \in \mathbbm{K} [x] \langle D_x
\rangle$ be a differential operator, and assume that the origin is an ordinary
point of~$L$, i.e.\ that $p_r (0) \neq 0$. By the classical Cauchy existence
theorem~\cite{Hille1976}, the equation $L \cdot f = 0$ then admits an $r$\mbox{-}dimensional
space of complex analytic solutions defined in a neighborhood of~$0$ and
characterized by $r$~initial values at~$0$. The
\tmtexttt{numerical\_solution()} method of differential operators computes
values of D-finite functions defined in this way. Perhaps the simplest example
is the computation of $e = \exp (1)$ as the value at~$1$ of the solution of
$f^{\prime} - f = 0$ such that $f (0) = 1$:

\tmsession{sage}{default}{\tmunfoldedio{sage:
}{(Dx-1).numerical\_solution(ini=[1], path=[0,1],
eps=1e-40)}{[2.71828182845904523536028747135266249775725 +/- 7.53e-42]}}

{\noindent}The last argument, \tmtexttt{eps}, provides an indication of the
accuracy the user is interested in. The method outputs a ``mid-rad interval'',
or {\tmem{ball}}, consisting of a multiple-precision center and a
low-precision radius. In normal circumstances, the diameter of this ball will
be of the order of magnitude of \tmtexttt{eps}, but this is not guaranteed to
be the case. In contrast, barring any bug, the mathematical result is
guaranteed to lie within the bounds of the ball.

Evaluations on the complex plane are also supported. Based on~(\ref{eq:erf}),
we may for instance evaluate the error function as follows:
\enlargethispage{\baselineskip}

\tmsession{sage}{default}{\tmunfoldedio{sage: }{(Dx\^{}2 +
2*x*Dx).numerical\_solution([0, 2/sqrt(pi)], [0, i])}{[+/- 2.91e-25] +
[1.6504257587975429 +/- 2.70e-17]*I}}

{\noindent}Allowed evaluation points and initial values include rational and
floating-point numbers, algebraic numbers, and constant symbolic expressions.

The reader may have noticed that the argument used to specify the evaluation
point is called {\tmem{path}}. The main reason why
\tmtexttt{numerical\_solution()} does not simply take an evaluation
{\tmem{point}} is that in general, solutions of differential equations with
singular points are not single-valued analytic functions. What
\tmtexttt{diffop.numerical\_solution(ini, path)} really computes is the
analytic continuation along \tmtexttt{path} of the solution of
\tmtexttt{diffop} defined in the neighborhood of the starting point by the
initial conditions \tmtexttt{ini}. Consider for example the operator $L = xD_x
+ D_x$, obtained by differentiating the inhomogeneous equation $xy^{\prime}
(x) = 1$. A basis of solutions of~$L$ is $\{ 1, \log (x) \}$, corresponding
respectively to the initial values $(y (1), y^{\prime} (1)) = (1, 0)$ and $(y
(1), y^{\prime} (1)) = (0, 1)$. Analytic continuation of the latter to~$x=-1$ along a
path that passes above the singular point $x = 0$ corresponds to the standard
determination of the complex logarithm, while a path that passes below the
origin yields another branch:

\tmsession{sage}{default}{
  \tmunfoldedio{sage: }{(x*Dx\^{}2 + Dx).numerical\_solution([0, 1], [1, i,
  -1])}{[+/- 5.11e-18] + [3.1415926535897932 +/- 4.38e-17]*I}
  \tmunfoldedio{sage: }{(x*Dx\^{}2 + Dx).numerical\_solution([0, 1], [1, -i,
  -1])}{[+/- 5.11e-18] + [-3.1415926535897932 +/- 4.38e-17]*I}
}

More generally, it is possible to compute the matrix that maps a vector
\begin{equation}
  [f (x_0), f^{\prime} (x_0), \tfrac{1}{2} f^{\prime\prime} (x_0), \ldots,
  \tfrac{1}{(r - 1) !} f^{(r - 1)} (x_0)] \label{eq:ini}
\end{equation}
of initial conditions\footnote{Note the inverse factorials in our definition of
initial conditions. This convention also applies to the \tmtexttt{ini}
argument of \tmtexttt{numerical\_solution()}. It is convenient in view of the
generalization to regular singular points discussed in the next section.}
at~$x_0$ to the vector $[f (x_1), f^{\prime} (x_1), \ldots]$ of ``initial''
conditions at some other point~$x_1$ that define the analytic continuation of
the same solution~$f$ along a given path.
Following van der Hoeven~\cite{vdH1999},
we call this matrix the
{\tmem{transition matrix}} along that path.

\tmsession{sage}{default}{
  \tmunfoldedio{sage: }{(x*Dx\^{}2 + Dx).numerical\_transition\_matrix([1, 2],
  1e-10)}{[ 1.0000000000 [0.69314718056 +/-
  6.99e-14]] \\ \relax
  [ \ \ \ \ \ \ \ \ \ \ \ 0 \ [0.50000000000 +/-
  9.1e-15]]}
}

\begin{example}
  Transition matrices along loops that turn around a singular point exactly
  once are called monodromy matrices. The first step of van~Enckevort and van~Straten's
  experimental study of differential equations of Calabi-Yau
  type~{\cite{EnckevortStraten2006}} consists in computing
  high-precision approximations of all monodromy matrices starting from a
  common base point. The \tmtexttt{numerical\_transition\_matrix()} method
  makes this computation straightforward. In the case of the singularity
  at~$1$ of the operator studied in Example~1 from their article, we
  obtain\footnote{In situations such as this one where the singular point is
  regular singular, it is also possible, and typically more efficient, to
  express the monodromy matrix using the primitives described in the next
  section, as suggested by van der Hoeven~{\cite{vdH2001}}.}:
  \tmsession{sage}{default}{
    \tminput{sage: }{%
    diffop = ((x*Dx)\^{}4 - x*(65*(x*Dx)\^{}4+130*(x*Dx)\^{}3 + 105*(x*Dx)\^{}2\\
    \hbox{}\hfill + 40*x*Dx + 6) + 4*x\^{}2*(4*x*Dx+3)*(x*Dx+1)\^{}2*(4*x*Dx+5))}
    \tminput{sage: }{mat = diffop.numerical\_transition\_matrix([1/2,1-i/2,3/2,1+i/2,1/2],\\
    \hbox{}\hfill1e-1000)}
    \tmunfoldedio{sage: }{mat[0,0] \ \ \# 990 digits omitted}{[1.000...000 +/-
    6.45e-998] + [0.733...6314 +/- 8.99e-998]*I}
  }
\end{example}

The analytic continuation algorithm behind these functions is based
on work by Chudnovsky and Chudnovsky~\cite{ChudnovskyChudnovsky1990}
with some later improvements \cite{vdH1999,Mezzarobba2011}. In the
language of numerical methods for ODEs, it is a high-order Taylor series
method with a step size of the order of the local radius of convergence of
solutions. The computation of the transition matrix for each step boils down
to that of sums of power series whose coefficients are generated by linear
recurrence relations. Two algorithms for this task are currently implemented:
a naïve iterative summation and a simple binary splitting method. In both
cases, the entries of a given column are actually computed all at once, by
working with truncated power series in a manner similar to direct-mode
automatic differentiation. Once the terms of the sum have decreased below the
target accuracy, the algorithm of~{\cite{Mezzarobba-numbounds}} is used to get
a rigorous bound on the tail of the series. Both the computation of the
numerical result and that of error bounds rely on Arb~{\cite{Johansson2013}}
for controlling the round-off errors in basic arithmetic operations.

\section{Regular Singular Connection Problems}

Singular connection problems are similar to the analytic continuation problems
of the previous section, except that one or both ends of the analytic
continuation path are singular points of the operator (points where the
leading coefficient vanishes). At a singular point, the Cauchy theorem does
not apply, and solutions of the operator may diverge or display a branch
point.

Our implementation supports the situation where the singularities involved are
{\tmem{regular singular points}}. A singular point~$x_0$ of $L \in \mathbbm{K}
[x] \langle D_x \rangle$ is regular when $L$~admits a full basis of solutions
each of the form
\begin{equation}
  (x - x_0)^{\alpha}  \left( g_0 (x) + g_1 (x) \log (x - x_0) + \cdots + g_p
  (x)  \frac{(\log (x - x_0))^p}{p!} \right) \label{eq:regsingexp}
\end{equation}
where $g_0, \ldots, g_p$ are analytic functions and $\alpha$ is an algebraic number.

A basic example is that of the modified Bessel equation~(\ref{eq:Bessel}),
which possesses a regular singular point at~$x = 0$ and whose standard solutions
are defined by their asymptotic behavior there. Starting
from~(\ref{eq:Bessel}), we can compute $K_0 (1)$ by:

\tmsession{sage}{default}{
  \tminput{sage: }{ini = [-1, log(2)-euler\_gamma]}
  \tmunfoldedio{sage: }{(x*Dx\^{}2 + Dx - x).numerical\_solution(ini, [0, 1],
  1e-10)}{[0.42102443824 +/- 2.11e-12]}
}

{\noindent}This command is to be understood as follows. It is classical
that, when~$x_0$ is a regular singular point of~$L$, a solution of~$L$ is
characterized by the coefficients of the monomials $(x - x_0)^{\nu} \log (x -
x_0)^k / k!$ in its expansion~(\ref{eq:regsingexp}) about~$x_0$, where
$\nu$~ranges over the roots of the so-called indicial polynomial of~$L$
at~$x_0$ and $k$~is less than the multiplicity of the root~$\nu$. The initial
conditions accepted by \tmtexttt{numerical\_solution()} in this case are the
coefficients of these monomials (ordered in a certain way, basically by
asymptotic dominance as $x \rightarrow x_0$). This definition generalizes
the convention~(\ref{eq:ini}) used at ordinary points. We call the
corresponding basis of solutions the {\tmem{canonical local
basis}}\footnote{Note that there is another common choice, the
{\tmem{Frobenius basis}}.} at~$x_0$. A convenience method provides the
distinguished monomials.

\tmsession{sage}{default}{\tmunfoldedio{sage: }{(x*Dx\^{}2 + Dx -
x).local\_basis\_monomials(0)}{ [log(x), 1]}}

{\noindent}In the case of Equation~(\ref{eq:Bessel}), we see from the
asymptotic condition that the coefficient of~$x^0$ is $\log 2 - \gamma$ and
that of $x^0 \log x$ is $- 1$.

In calls to \tmtexttt{numerical\_transition\_matrix()}, both ends
of the path can be regular singular points. The resulting matrix expresses the
change of basis between the respective canonical local bases.

\begin{example}[Face-Centered Cubic Lattices]
  Koutschan~{\cite{Koutschan2013b}} shows that the lattice Green's function $P
  (x)$ of the four-dimensional face-centered cubic lattice is annihilated by the
  operator

  \tmsession{sage}{default}{
  \scriptsize
    \tminput{sage:\ }{%
diffop = ((-1+x)*x\^{}3*(2+x)*(3+x)*(6+x)*(8+x)*(4+
3*x)\^{}2*Dx\^{}4 \\
\hbox{}\hspace{8ex}+ 2*x\^{}2*(4+3*x)*(-3456-2304*x+3676*x\^{}2+4920*x\^{}3+
2079*x\^{}4+356*x\^{}5+21*x\^{}6)*Dx\^{}3 \\
\hbox{}\hspace{8ex}+ 6*x*(-5376-5248*x+11080*x\^{}2+25286*x\^{}3+19898*x\^{}4+
7432*x\^{}5+1286*x\^{}6+81*x\^{}7)*Dx\^{}2 \\
\hbox{}\hspace{8ex}+ 12*(-384+224*x+3716*x\^{}2+7633*x\^{}3+6734*x\^{}4+2939*x\^{}5 + 604*x\^{}6+45*x\^{}7)*Dx \\
\hbox{}\hspace{8ex}+ 12*x*(256+632*x+702*x\^{}2+382*x\^{}3+98*x\^{}4+9*x\^{}5))}
  }

  {\noindent}Koutschan then evaluates~$P (1)$ in order to obtain the return probability
  of the lattice. Both $x = 0$ and $x = 1$ are regular singular points of the
  operator. Let us examine the possible behaviors of solutions at these two
  points:

  \tmsession{sage}{default}{
    \tmunfoldedio{sage: }{diffop.local\_basis\_monomials(0)}{
    [1/6*log(x)\^{}3, 1/2*log(x)\^{}2, log(x), 1]}
    \tmunfoldedio{sage: }{diffop.local\_basis\_monomials(1)}{ [1, (x -
    1)*log(x - 1), x - 1, (x - 1)\^{}2]}
  }

  {\noindent}As it turns out, the space of analytic solutions at the origin
  has dimension~1, and all other solutions involve logarithms. It is easy to
  see from the definition of~$P$ that $P (0) = 1$, so that the function we are
  interested in is exactly the last element of the local canonical basis. At
  $x = 1$, there is a three-dimensional subspace of solutions that tend to
  zero, spanned by the last three basis elements, while the first basis
  function has a finite limit. Thus the value of~$P$ there is the upper right
  entry of the transition matrix from $0$~to~$1$, namely:

  \tmsession{sage}{default}{\tmunfoldedio{sage:
  }{diffop.numerical\_transition\_matrix([0, 1],
  1e-60)[0,3]}{
 [1.1058437979212047601829954708859 +/- 4.90e-32] + [+/- 4.55e-41]*I
  }}
\end{example}

\begin{example}[Asymptotics of Apéry Numbers]
  In his celebrated proof that $\zeta (3)$ is irrational~\cite{Apery1979}, Apéry introduces
  the two sequences
  \[ a_n = \sum_{k = 0}^n \binom{n}{k}^2  \binom{n + k}{k}^2, \hspace{2em} b_n
     = \sum_{k = 1}^n \left( \frac{a_n}{k^3} - \sum_{m = 1}^k \frac{(- 1)^m
     \binom{n}{k}^2  \binom{n + k}{k}^2}{2 m^3  \binom{n}{m}  \binom{n +
     m}{m}} \right) . \]
  The generating series $a (x) = \sum_{n = 0}^{\infty} a_n x^n$ and $b (x) =
  \sum_{n = 0}^{\infty} b_n x^n$ are solutions of the following operator:

  \tmsession{sage}{default}{
    \tminput{sage: }{diffop = (x\^{}2*(x\^{}2-34*x+1)*Dx\^{}4 +
    5*x*(2*x\^{}2-51*x+1)*Dx\^{}3\\
    \hbox{}\hfill + (25*x\^{}2-418*x+4)*Dx\^{}2 + (15*x-117)*Dx + 1)}
  }

  This operator has three regular singular points: $0$, $\alpha = (\sqrt{2}
  + 1)^4 \approx 33.971$, and $\alpha^{- 1} = (\sqrt{2} - 1)^4 \approx
  0.029$.

  \tmsession{sage}{default}{\tmunfoldedio{sage: }{s =
  diffop.leading\_coefficient().roots(AA, multiplicities=False); s}{ [0,
  0.02943725152285942?, 33.97056274847714?]}}

  {\noindent}This implies that the radius of convergence of $a (x)$ is in $\{
  \alpha, \alpha^{- 1}, \infty \}$, so that $a_n$ must be of one of the forms $\alpha^{- n
  + o (n)}$, $\alpha^{n + o (n)}$ and $e^{o (n)}$ as $n \rightarrow
  \infty$. The same holds for~$b_n$. A key step of Apéry's proof is to show
  that $a_n = \alpha^{n + o (n)}$ and $b_n = \alpha^{n + o (n)}$, but $b_n -
  \zeta (3) a_n = \alpha^{- n + o (n)}$.

  Analytic solutions at $x = 0$ are determined by the first two coefficients
  of their Taylor expansions. At $x = \alpha^{- 1}$, the canonical local basis takes the form $(f_0,
  f_1, f_2, f_3)$ where $f_0, f_2, f_3$ are analytic and
  $f_1 (x) \sim \sqrt{x - \alpha^{- 1}}$.

  \tmsession{sage}{default}{
    \tmunfoldedio{sage: }{diffop.local\_basis\_monomials(0)}{
    [1/2*log(x)\^{}2, log(x), 1, x]}
    \tmunfoldedio{sage: }{diffop.local\_basis\_monomials(s[1])}{ [1, sqrt(x -
    0.02943725152285942?), x - 0.02943725152285942?,\\
    (x - 0.02943725152285942?)\^{}2]}
  }

  {\noindent}To prove that $a_n = \alpha^{n + o (n)}$, it is enough to check
  that the coefficient $c_1$ in the decomposition $a (x) = \sum c_i f_i (x)$
  is nonzero.

  \tmsession{sage}{default}{\tminput{sage: }{mat =
  diffop.numerical\_transition\_matrix([0, sing[1]], 1e-40)}}

  {\noindent}From the initial values $a_0 = 1$, $a_1 = 5$,  we see that $c_1$
  is equal to

  \tmsession{sage}{default}{
    \tmunfoldedio{sage: }{mat[1,2] +
    5*mat[1,3]}{[4.546376247522844600239593024915161553303 +/- 9.85e-41]*I}
  }

  {\noindent}With a little more work, using the method of singularity
  analysis~{\cite{FlajoletSedgewick2009}}, this is actually enough to prove
  that $a_n = n^{- 3 / 2} \alpha^n  (\lambda + \Omicron (1 / n))$ where
  $\lambda$ is given by

  \tmsession{sage}{default}{\tmunfoldedio{sage:
  }{CBF(-i/(2*sqrt(pi)))*sqrt(sing[1])*(mat[1,2] +
  5*mat[1,3])}{[0.220043767112643 +/- 2.06e-16]}}

  {\noindent}This result agrees with the closed form $a_n \sim \sqrt{\alpha}
  (2 \pi \sqrt{2})^{- 3 / 2} n^{- 3 / 2} \alpha^n$ from~{\cite{Hirschhorn2012}}.

  In the same way, using the initial terms $b_0 = 0$, $b_1 = 6$ of $(b_n)$, we
  can compute the coefficient of~$f_1$ in the decomposition of $b (x)$ and
  check that the ratio of the two values we obtained is~$\zeta (3)$.

  \tmsession{sage}{default}{\tmunfoldedio{sage: }{6*mat[1,3]/(mat[1,2] +
  5*mat[1,3])}{[1.2020569031595942853997381615114499907650 +/- 6.08e-41]}}
\end{example}

\paragraph{Acknowledgements.}This work was prompted by discussions with
Fredrik Johansson. I am also indebted to the Sage developers, in particular
Vincent Delecroix, Jeroen Demeyer and Clemens Heuberger, who reviewed or
helped writing a number of Sage patches vital to this project, and to Christoph Lauter for comments on a version of this text.

\begin{thebibliography}{10}

\enlargethispage\baselineskip

\bibitem{Apery1979}
R.~Apéry.
\newblock Irrationalité de $\zeta(2)$ et $\zeta(3)$.
\newblock {\em Astérisque}, 61, page 11–13, 1979.

\bibitem{BenoitChyzakDarrasseGerholdMezzarobbaSalvy2010}
A.~Benoit, F.~Chyzak, A.~Darrasse, S.~Gerhold, M.~Mezzarobba, and B.~Salvy.
\newblock The dynamic dictionary of mathematical functions ({DDMF}).
\newblock In K.~Fukuda \emph{et al.}, editors,
  {\em Mathematical Software — ICMS 2010}, volume 6327 of {\em Lecture Notes
  in Computer Science}, page 35–41. Springer, 2010.

\bibitem{ChudnovskyChudnovsky1990}
D.~V. Chudnovsky and G.~V. Chudnovsky.
\newblock Computer algebra in the service of mathematical physics and number
  theory.
\newblock In D.~V. Chudnovsky and R.~D. Jenks, editors, {\em Computers in
  Mathematics}, volume 125 of {\em Lecture Notes in Pure and Applied
  Mathematics}, page 109–232, Stanford University, 1986. Dekker.

\bibitem{FlajoletSedgewick2009}
P.~Flajolet and R.~Sedgewick.
\newblock {\em Analytic Combinatorics}.
\newblock Cambridge University Press, 2009.

\bibitem{Hille1976}
E.~Hille.
\newblock {\em Ordinary differential equations in the complex domain}.
\newblock Wiley, 1976.
\newblock {D}over reprint, 1997.

\bibitem{Hirschhorn2012}
M.~D. Hirschhorn.
\newblock Estimating the {A}péry numbers.
\newblock {\em The Fibonacci Quarterly}, 50(2):129--131, 2012.

\bibitem{Johansson2013}
F.~Johansson.
\newblock Arb: a {C} library for ball arithmetic.
\newblock {\em ACM Communications in Computer Algebra}, 47(4):166–169, 2013.
\newblock \url{http://fredrikj.net/arb/}

\bibitem{Kauers2013}
M.~Kauers.
\newblock The holonomic toolkit.
\newblock {\em Computer Algebra in Quantum Field Theory: Integration, Summation
  and Special Functions, Texts and Monographs in Symbolic Computation}, 2013.

\bibitem{KauersJaroschekJohansson2015}
M.~Kauers, M.~Jaroschek, and F.~Johansson.
\newblock {O}re polynomials in {S}age.
\newblock In J.~Gutierrez, J.~Schicho, and M.~Weimann, editors, {\em Computer
  Algebra and Polynomials}, page 105–125. Springer, 2015.

\bibitem{Koutschan2013b}
C.~Koutschan.
\newblock Lattice {G}reen's functions of the higher-dimensional face-centered
  cubic lattices.
\newblock {\em Journal of Physics A: Mathematical and Theoretical},
  46(12):125005, 2013.

\bibitem{Lanczos1956}
C.~Lanczos.
\newblock {\em Applied analysis}.
\newblock Prentice-Hall, 1956.

\bibitem{Mallinger1996}
C.~Mallinger.
\newblock Algorithmic manipulation and transformations of univariate holonomic
  functions and sequences.
\newblock Diplomarbeit, RISC-Linz, 1996.

\bibitem{Mezzarobba-numbounds}
M.~Mezzarobba.
\newblock Truncation bounds for {D}-finite series.
\newblock In preparation.

\bibitem{Mezzarobba2010}
M.~Mezzarobba.
\newblock {NumGfun}: a package for numerical and analytic computation with
  {D}-finite functions.
\newblock In S.~M. Watt, editor, {\em {ISSAC} '10}, page 139–146. ACM, 2010.

\bibitem{Mezzarobba2011}
M.~Mezzarobba.
\newblock {\em Autour de l'évaluation numérique des fonctions {D}-finies}.
\newblock Thèse de doctorat, École polytechnique, 2011.

\bibitem{SalvyZimmermann1994}
B.~Salvy and P.~Zimmermann.
\newblock Gfun: A {M}aple package for the manipulation of generating and
  holonomic functions in one variable.
\newblock {\em ACM Transactions on Mathematical Software}, 20(2):163–177,
  1994.

\bibitem{Stanley1999}
R.~P. Stanley.
\newblock {\em Enumerative combinatorics}, volume~2.
\newblock Cambridge University Press, 1999.

\bibitem{Sage}
{The SageMath Developers}.
\newblock {SageMath} mathematics software, 2005–.
\newblock \url{http://www.sagemath.org/}

\bibitem{vdH1999}
J.~van~der Hoeven.
\newblock Fast evaluation of holonomic functions.
\newblock {\em Theoretical Computer Science}, 210(1):199–216, 1999.

\bibitem{vdH2001}
J.~van~der Hoeven.
\newblock Fast evaluation of holonomic functions near and in regular
  singularities.
\newblock {\em Journal of Symbolic Computation}, 31(6):717–743, 2001.

\bibitem{EnckevortStraten2006}
C.~van Enckevort and D.~van Straten.
\newblock Monodromy calculations of fourth order equations of {C}alabi-{Y}au
  type.
\newblock In J.~D. Lewis, S.-T. Yau, and N.~Yui, editors, {\em Mirror Symmetry
  {V}}, volume~38 of {\em AMS/IP Studies in Advanced Mathematics}, page
  539–559. International Press, 2006.

\end{thebibliography}

\end{document}

### 2019-05-17 papers/issac19demo/main.tex
inputenc
software demo: minor presentation fix
0.4

@@
 \usepackage{booktabs}
+\usepackage[utf8]{inputenc}
 
@@
 The operators $P$ are called \emph{telescopers,} and the corresponding operators $Q$ are called their \emph{certificates.}
+Our new version of ore\_algebra includes an implementation of Chyzak's algorithm~\cite{chyzak00} for creative telescoping in the general D-finite setting.
+
 Creative telescoping is a central operation for evaluating definite sums and integrals~\cite{chyzak14}.
@@
 \frac{{-}u^2 v^2{+}u^2{+}v^2{-}1}{t u^3 v^2{-}t u^3 v{+}t u^2 v^3{-}2 t u^2 v^2{+}2 t u^2 v{-}t u^2{-}t u v^3{+}2 t u v^2{-}2 t
-  u v+t u-t v^2+t v-u^2 v^2+u^2 v+u v^2-u v},
+  u v{+}t u{-}t v^2{+}t v{-}u^2 v^2{+}u^2 v{+}u v^2{-}u v},
 \]

### 2019-05-17 doc/conf.py

@@ -30,7 +30,7 @@ author = u'Manuel Kauers, Maximilian Jaroschek, Fredrik Johansson'
 # The short X.Y version
 version = u''
 # The full version, including alpha/beta/rc tags
-release = u'0.3'
+release = u'0.4'
 
 
 # -- General configuration ---------------------------------------------------

### 2019-05-17 setup.py
also increase version number in setup.py

@@ -23,7 +23,7 @@ class TestCommand(Command):
 
 setup(
     name = "ore_algebra",
-    version = "0.3",
+    version = "0.4",
     author = "Manuel Kauers, Maximilian Jaroschek, Fredrik Johansson",
     author_email = "manuel@kauers.de",
     license = "GPL",

### 2020-03-11 papers/icms2016/main.tex
icms2016: typos (sing → s in two places)

@@
   \tmsession{sage}{default}{\tminput{sage: }{mat =
-  diffop.numerical\_transition\_matrix([0, sing[1]], 1e-40)}}
+  diffop.numerical\_transition\_matrix([0, s[1]], 1e-40)}}
 
@@
   \tmsession{sage}{default}{\tmunfoldedio{sage:
-  }{CBF(-i/(2*sqrt(pi)))*sqrt(sing[1])*(mat[1,2] +
+  }{CBF(-i/(2*sqrt(pi)))*sqrt(s[1])*(mat[1,2] +
   5*mat[1,3])}{[0.220043767112643 +/- 2.06e-16]}}
